# Installing ferx {#sec-installation}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

ferx is an R package with a compiled Rust engine (ferx-core) inside. Installing it compiles that engine from source, so you need a Rust toolchain and the first install takes a long time. The ferx-r README notes that later installs are faster.

## Requirements

- **R** with a working build toolchain. The package's `SystemRequirements` are GNU make, Cargo and rustc (>= 1.70).
- **Rust**: a standard stable toolchain installed with [rustup](https://www.rust-lang.org/tools/install). No custom toolchain is required.
- **Windows**: Rtools matching your R version, plus the GNU-ABI Rust toolchain. See the Windows notes in the [ferx-r README](https://github.com/FeRx-NLME/ferx-r#windows).

## Installing the pinned build

This book documents one specific ferx-r commit, `{{< var ferx_r_sha_short >}}` (package version {{< var ferx_r_version >}}, engine ferx-core `{{< var ferx_core_sha_short >}}`). Every output in the book was produced with it. To get exactly the same behaviour, install that commit:

```r
install.packages("pak")
pak::pak("FeRx-NLME/ferx-r@{{< var ferx_r_sha >}}")
```

Installing without a commit gives you the latest development version. That is fine for your own work, but its output can differ from what this book shows:

```r
pak::pak("FeRx-NLME/ferx-r")
```

## Which version do you have? {#sec-versions}

```{r}
#| label: version-chapters
#| include: false
# The chapters that carry each kind of version callout, read from the chapter sources,
# so this list cannot drift from the callouts themselves.
chapter_dir <- if (dir.exists("chapters")) "chapters" else "."
chapter_files <- list.files(chapter_dir, pattern = "^[0-9]{2}-.*\\.qmd$", full.names = TRUE)
chapters_with <- function(title) {
  hits <- Filter(function(f) any(grepl(paste0("^\\s*## ", title), readLines(f))), chapter_files)
  ids <- vapply(hits, function(f) sub(".*\\{#(sec-[a-z-]+)\\}.*", "\\1",
                                      grep("^# .*\\{#sec-", readLines(f), value = TRUE)[1]), character(1))
  if (length(ids)) paste0("@", ids, collapse = ", ") else "none at this pin"
}
needs_newer <- chapters_with("Needs ferx-r newer than")
changed_after <- chapters_with("Changed after ferx-r")
```

The pinned build is newer than the last ferx-r release, {{< var ferx_r_release >}}. `packageVersion("ferx")` (below) tells you which one you have. With {{< var ferx_r_release >}} most of the book runs as shown, with three kinds of difference:

- **Features newer than {{< var ferx_r_release >}}.** Some functions, arguments, settings and examples do not exist in the release. @sec-reference-newer lists them, and each chapter that uses one says so in a callout titled "Needs ferx-r newer than {{< var ferx_r_release >}}": `r needs_newer`.
- **Behaviour that changed after {{< var ferx_r_release >}}.** A callout titled "Changed after ferx-r {{< var ferx_r_release >}}" marks it where the book relies on it: `r changed_after`.
- **Numbers.** The engine changed too, so your results can differ from the printed ones. FOCE fits move most: the FOCE objective changed after {{< var ferx_r_release >}}, which shifts their OFVs and, slightly, their estimates.

To get the book's output, install the pinned commit above. To stay on the release, install its tag, `pak::pak("FeRx-NLME/ferx-r@v{{< var ferx_r_release >}}")`, and read the callouts.

## Checking the installation

Load the package and confirm the version:

```{r}
#| label: version
library(ferx)
packageVersion("ferx")
```

The package ships example models and datasets, so a real fit is the quickest end-to-end check. It exercises the model parser, the data reader and the estimation engine:

```{r}
#| label: smoke-fit
ex <- ferx_example("warfarin")
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
fit$converged
```

If this prints `TRUE`, ferx is installed and working. @sec-complete-analysis walks through the same example as a full analysis.

::: {.callout-note}
## Stopping a long fit
`ferx_fit()` checks for R interrupts while it runs, so **Ctrl-C** (or RStudio's stop button) cancels a fit. The call then ends with an error saying the fit was cancelled. The ferx-core [installation page](https://ferx-nlme.org/ferx-core/installation.html) describes the behaviour in detail.
:::

## Reference

::: {.callout-tip}
## Reference
- ferx-r: [README: installation](https://github.com/FeRx-NLME/ferx-r#installation), including Windows and Docker
- ferx-core: [installation](https://ferx-nlme.org/ferx-core/installation.html)
- R help: `?ferx_example`, `?ferx_fit`
:::
