---
title: "Quickstart"
---

This chapter fits a one-compartment oral PK model to the bundled warfarin dataset.
It covers the complete loop: data → model → fit → inspect → plot.

## Load ferx

```{r}
#| label: setup
library(ferx)
library(ggplot2)
```

## The example dataset

`ferx_example()` returns paths to a bundled model file and a NONMEM-format CSV.
Calling it with no argument lists all bundled examples:

```{r}
#| label: list-examples
ferx_example()
```

```{r}
#| label: example-paths
ex <- ferx_example("warfarin")
ex
```

Peek at the data:

```{r}
#| label: peek-data
obs <- read.csv(ex$data)
head(obs)
```

The dataset follows the standard NONMEM column convention:

- `EVID = 1` for dose records (with non-zero `AMT`), `EVID = 0` for observations.
- `MDV = 1` masks a row from the likelihood (typically every dose row); `MDV = 0` for observed concentrations.
- `CMT` selects the compartment for dose / observation. For a one-compartment oral model with a single endpoint the dose lands in the absorption compartment and observations come from the central compartment — both are `CMT = 1` here because the analytical `one_cpt_oral` solver tracks them together.
- `RATE = 0` is a bolus dose; non-zero `RATE` indicates an infusion (zero-order input).

`DV` values written as `.` (NONMEM convention for missing) are read by `read.csv()` as text; ferx parses them as `NA` internally.

The warfarin dataset bundled with ferx has 10 subjects and 110 observations after dose rows are excluded.

## The model file

```{r}
#| label: show-model
ferx_model_show(ex$model)
```

The model DSL is described in detail in @sec-model-dsl.

## Fit the model

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(model = ex$model, data = ex$data, verbose = FALSE)
```

`verbose = FALSE` silences per-iteration progress; omit it to watch the optimiser.
Convergence is reached when the optimiser's stopping criterion is met (gradient
norm or function tolerance, depending on the optimiser).

## Inspect results

```{r}
#| label: summary
summary(fit)
```

`summary(fit)` prints a compact run report: convergence, method, OFV/AIC/BIC,
shrinkage, the Durbin-Watson autocorrelation statistic, the covariance step
status, and any active warnings. It also lists the `[fit_options]` actually in
effect (model file vs. call-time overrides).

For the full parameter table including standard errors, use `print(fit)`:

```{r}
#| label: print
print(fit)
```

For programmatic access, `ferx_estimates()` returns a tidy data frame:

```{r}
#| label: estimates
ferx_estimates(fit)
```

Notable columns:

- `transform` — natural scale of the parameter (`identity`, `variance`, `proportional`, ...).
- `estimate`, `se`, `rse_pct` — point estimate, standard error, and relative SE.
- `lower_95`, `upper_95` — asymptotic 95% confidence interval on the estimation scale.
- `estimate_natural` / `lower_95_natural` / `upper_95_natural` — back-transformed values where applicable (e.g. CV% from a log-normal omega).
- `init_as_sd` — `TRUE` if the model file declared the initial value on the SD scale (`(sd)` qualifier) rather than variance.

## Individual diagnostics

Per-observation diagnostics live in `fit$sdtab`; per-subject EBEs in `fit$ebe_etas`.

```{r}
#| label: sdtab
head(fit$sdtab)
head(fit$ebe_etas)
```

## Goodness-of-fit plots

The two most informative GOF plots are IPRED vs DV and CWRES vs PRED.

```{r}
#| label: gof
#| fig-width: 8
#| fig-height: 4
sdtab <- fit$sdtab

p1 <- ggplot(sdtab, aes(x = IPRED, y = DV)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, colour = "#1F6B7A") +
  labs(title = "IPRED vs DV", x = "Individual prediction", y = "Observed DV")

p2 <- ggplot(sdtab, aes(x = PRED, y = CWRES)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, colour = "#1F6B7A") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "#C8893E") +
  labs(title = "CWRES vs PRED", x = "Population prediction", y = "CWRES")

patchwork::wrap_plots(p1, p2)
```

For a full treatment of diagnostics see @sec-diagnostics.

## What's next

- @sec-model-workflow — pre-fit inspection and editing of `.ferx` files
- @sec-model-dsl — understand every section of the `.ferx` file
- @sec-fitting — fitting options, methods, and convergence
- @sec-diagnostics — full diagnostic suite and shrinkage
- @sec-simulation — simulate from the fitted model and overlay a VPC
