---
title: "Quickstart"
---

This chapter fits a one-compartment oral PK model to the bundled warfarin dataset.
It covers the complete loop: data → model → fit → inspect → plot.

## Load ferx

```{r}
#| label: setup
library(ferx)
library(ggplot2)
```

## The example dataset

`ferx_example()` returns paths to a bundled model file and NONMEM-format CSV.

```{r}
#| label: example-paths
ex <- ferx_example("warfarin")
ex
#> $model
#> [1] "/path/to/ferx/examples/models/warfarin.ferx"
#> $data
#> [1] "/path/to/ferx/examples/data/warfarin.csv"
```

Peek at the data:

```{r}
#| label: peek-data
obs <- read.csv(ex$data)
head(obs)
#>   ID TIME  DV EVID AMT CMT MDV
#> 1  1  0.0  NA    1  70   1   1
#> 2  1  0.5 3.1    0   0   2   0
#> 3  1  1.0 5.4    0   0   2   0
#> ...
```

The dataset follows the standard NONMEM column convention:
`EVID = 1` for dose rows, `EVID = 0` for observations, `CMT = 1` for the depot compartment, `CMT = 2` for the central compartment.

## The model file

```{r}
#| label: show-model
ferx_model_show(ex$model)
```

```
# One-compartment oral PK model (warfarin)

[parameters]
  theta TVCL(0.134, 0.001, 10.0)
  theta TVV(8.1, 0.1, 500.0)
  theta TVKA(1.0, 0.01, 50.0)

  omega ETA_CL ~ 0.07
  omega ETA_V  ~ 0.02
  omega ETA_KA ~ 0.40

  sigma PROP_ERR ~ 0.01

[individual_parameters]
  CL = TVCL * exp(ETA_CL)
  V  = TVV  * exp(ETA_V)
  KA = TVKA * exp(ETA_KA)

[structural_model]
  pk one_cpt_oral(cl=CL, v=V, ka=KA)

[error_model]
  DV ~ proportional(PROP_ERR)

[initial_values]
  theta = [0.2, 10.0, 1.5]
  omega = [0.09, 0.04, 0.30]
  sigma = [0.02]

[fit_options]
  method     = foce
  maxiter    = 300
  covariance = true
```

The model DSL is described in detail in @sec-model-dsl.

## Fit the model

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(model = ex$model, data = ex$data)
```

Fitting prints progress to the console and terminates when the gradient norm drops below the convergence threshold.

## Inspect results

```{r}
#| label: summary
summary(fit)
#>
#> ferx fit — warfarin (one_cpt_oral, FOCE)
#>
#> THETA (fixed effects):
#>   Parameter   Estimate   SE      CV%    95% CI
#>   TVCL         0.133    0.006   4.8    [0.121, 0.145]
#>   TVV          7.89     0.31    3.9    [7.28, 8.50]
#>   TVKA         1.49     0.16   10.6    [1.17, 1.81]
#>
#> OMEGA (between-subject variance):
#>   ETA_CL       0.0425   0.0109  CV~21%
#>   ETA_V        0.0162   0.0068  CV~13%
#>   ETA_KA       0.328    0.090   CV~63%
#>
#> SIGMA (residual variance):
#>   PROP_ERR     0.0162   0.0021  CV~13%
#>
#> OFV: 344.2   AIC: 358.2   BIC: 378.1
#> Subjects: 32   Observations: 251
```

Individual estimates and diagnostics are in `fit$sdtab`:

```{r}
#| label: sdtab
head(fit$sdtab)
#>   ID TIME  DV   PRED  IPRED   CWRES   IWRES  EBE_OFV N_OBS
#> 1  1  0.5  3.1  3.44   3.18  -0.98  -0.25   12.4     8
#> ...

# Per-subject ETAs are in fit$ebe_etas
head(fit$ebe_etas)
#>   ID  ETA_CL   ETA_V  ETA_KA
#> 1  1  -0.07    0.02    0.12
#> ...
```

## Goodness-of-fit plots

The two most informative GOF plots are IPRED vs DV and CWRES vs PRED.

```{r}
#| label: gof
#| fig-width: 8
#| fig-height: 4
sdtab <- fit$sdtab

p1 <- ggplot(sdtab, aes(x = IPRED, y = DV)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, colour = "#1F6B7A") +
  labs(title = "IPRED vs DV", x = "Individual prediction", y = "Observed DV")

p2 <- ggplot(sdtab, aes(x = PRED, y = CWRES)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, colour = "#1F6B7A") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "#C8893E") +
  labs(title = "CWRES vs PRED", x = "Population prediction", y = "CWRES")

patchwork::wrap_plots(p1, p2)
```

For a full treatment of diagnostics see @sec-diagnostics.

## What's next

- @sec-model-dsl — understand every section of the `.ferx` file
- @sec-fitting — fitting options, methods, and convergence
- @sec-diagnostics — full diagnostic suite and shrinkage
- @sec-simulation — simulate from the fitted model and overlay a VPC
