---
title: "Model Workflow"
---

## Overview {#sec-model-workflow}

`ferx_fit()` is the punch line, but a productive modeling session spends most
of its time *before* and *between* fits — scaffolding a model file, sanity-checking
initial estimates, editing one section, refitting. This chapter shows the
helpers that make that loop fast.

```{r}
#| label: setup
#| message: false
library(ferx)
```

## Scaffolding a new model file

`ferx_model_new()` writes a complete, valid `.ferx` file from a template.
Five templates ship today: `1cpt_oral`, `1cpt_iv`, `2cpt_oral`, `2cpt_iv`, and
`ode` (a generic ODE skeleton).

```{r}
#| label: scaffold-print
ferx_model_new(
  path      = tempfile(fileext = ".ferx"),
  template  = "1cpt_oral",
  edit      = FALSE,
  print     = TRUE,
  overwrite = TRUE
)
```

`print = TRUE` echoes the generated file. With `edit = TRUE` (the default), the
file is opened in your editor; `overwrite = FALSE` (the default) refuses to
clobber an existing file.

## Inspecting a model before fitting

`ferx_model_inspect()` parses the file and reports its structural shape — model
type, ETA list, IOV, residual error — without compiling or fitting:

```{r}
#| label: inspect
ex <- ferx_example("warfarin")
ferx_model_inspect(ex$model)
```

This is the cheapest way to confirm that the file you just edited still
declares the structure you think it does.

`ferx_model_validate()` is the syntax check: it returns `TRUE` if every
required section is present and parseable, and prints a per-section status
line:

```{r}
#| label: validate
ferx_model_validate(ex$model)
```

`ferx_model_show()` prints the raw file contents — handy in a Quarto chunk
when the reader wants to see what is being fit.

## Editing sections programmatically

For scripted workflows (e.g. covariate-search loops, simulation studies that
sweep theta starts), `ferx_model_section()` and `ferx_model_set_section()`
read and replace one block at a time:

```{r}
#| label: read-section
ferx_model_section(ex$model, "individual_parameters")
```

```{r}
#| label: edit-section
#| eval: false
# Copy the bundled model somewhere we can edit it
tmp <- tempfile(fileext = ".ferx")
ferx_model_edit(ex$model, dest = dirname(tmp),
                save_as = basename(tmp), .editor = identity)

# Replace one section
new_ip <- c(
  "  CL = TVCL * (WT / 70)^0.75 * exp(ETA_CL)",
  "  V  = TVV  * (WT / 70)      * exp(ETA_V)",
  "  KA = TVKA                  * exp(ETA_KA)"
)
ferx_model_set_section(tmp, "individual_parameters", new_ip)
```

`ferx_model_edit()` copies a bundled (or any read-only) model to a writable
location and opens it. Pass `.editor = identity` (or any no-op function) when
running non-interactively, e.g. in a knitted chapter or a CI pipeline.

## Pre-fit sanity checks

### `ferx_check_init()`

Before kicking off a 30-minute SAEM fit, verify that the model and the initial
estimates produce a finite, sensible OFV. `ferx_check_init()` runs a handful
of FOCEI iterations and returns a list with `$trace` (per-iteration OFV,
gradient norm, step norm) and `$summary` (start/end OFV, OFV drop,
convergence flag):

```{r}
#| label: check-init
#| eval: false
ci <- ferx_check_init(ex$model, ex$data)
ci$summary
head(ci$trace)
```

A finite start OFV with a downward trajectory means the model and inits are
healthy enough to commit to a real fit. A non-finite start OFV or an
explosive gradient norm usually points to a scale mismatch, the wrong CMT
numbering, or a parameter bound that excludes the true value.

### `ferx_inits_from_nca()`

For a new drug — or when porting a model from another tool — non-compartmental
analysis is a cheap source of starting values. `ferx_inits_from_nca()` runs an
NCA on the dataset and returns `theta` and `omega` starting values:

```{r}
#| label: inits-nca
#| cache: true
inits <- ferx_inits_from_nca(ex$model, ex$data, method = "nca")
inits$theta
inits$omega
```

The same routine is invoked from `ferx_fit()` via the `inits_from_nca`
argument:

| Value | Behaviour |
|---|---|
| `FALSE` (default) | Use initial values from the model file |
| `TRUE` / `"nca_sweep"` | Try both file inits and NCA inits; keep the lower-OFV fit |
| `"nca"` | Replace file inits with NCA-derived values |
| `"nca_ebe"` | Use NCA inits, then refine with a quick EBE pass |

## Putting it together

A typical first-pass workflow on a new dataset:

```{r}
#| eval: false
# 1. Scaffold a model from a template
path <- "my_model.ferx"
ferx_model_new(path, template = "1cpt_oral")

# 2. Edit (manually or programmatically), then validate
ferx_model_validate(path)
ferx_model_inspect(path)

# 3. Sanity-check inits before the real fit
ferx_check_init(path, "my_data.csv")

# 4. Fit (optionally bootstrap inits from NCA)
fit <- ferx_fit(path, "my_data.csv", inits_from_nca = TRUE)
```

## What's next

- @sec-model-dsl — every block of the `.ferx` file
- @sec-fitting — choosing methods, optimizers, and multi-start
