---
title: "Fitting"
---

## Overview {#sec-fitting}

`ferx_fit()` is the central function.
It compiles the model, runs the optimiser, and returns a `ferx_fit` object with estimates, standard errors, and individual diagnostics.

```{r}
#| label: setup
#| message: false
library(ferx)
ex <- ferx_example("warfarin")
```

## Estimation methods

Pass `method` to choose the algorithm:

| Method | Description | Typical use |
|---|---|---|
| `"foce"` | First-Order Conditional Estimation | Default, fast, well-tested |
| `"focei"` | FOCE with interaction | When residual error depends on ETAs |
| `"saem"` | Stochastic Approximation EM | Models with many ETAs or strong nonlinearity |
| `"gn"` | Gauss-Newton / BHHH | Quick exploration; no covariance |
| `"gn_hybrid"` | Gauss-Newton then FOCEI polish | Fast warm start, rigorous finish |

```{r}
#| label: fit-foce
#| cache: true
fit_foce <- ferx_fit(ex$model, ex$data, method = "foce")
```

```{r}
#| label: fit-focei
#| cache: true
fit_focei <- ferx_fit(ex$model, ex$data, method = "focei")
```

### Method chains

Pass a character vector to run methods in sequence.
Each stage is seeded with the previous stage's converged parameters.
Only the final stage produces the reported covariance and diagnostics.

```{r}
#| label: fit-chain
#| cache: true
fit_chain <- ferx_fit(ex$model, ex$data, method = c("saem", "focei"))
summary(fit_chain)
```

The chain summary shows the full sequence:

```
Method chain: saem → focei
```

### FOCE vs FOCEI

Both methods integrate over the ETA distribution using a second-order Laplace approximation.
The difference is in how the residual error is treated:

- **FOCE**: the residual function is linearised around ETA = 0 (no interaction between ETA and the error model)
- **FOCEI**: the residual function is linearised around the individual EBE — captures the interaction when `ipred` appears inside the error model

For proportional error on concentration data, FOCEI is typically more accurate.
The performance difference is small; use FOCEI by default.

## The covariance step

`covariance = TRUE` triggers the Hessian-based covariance matrix after convergence.
This provides standard errors, 95% confidence intervals, and the condition number.

```{r}
#| label: fit-cov
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE,
                optimizer_trace = TRUE)
```

If the covariance matrix is non-positive-definite or cannot be inverted, ferx
emits a warning and sets `covariance_status` to `Failed`; all SE fields are
`None`. A near-zero or negative eigenvalue usually indicates an
over-parameterised model or a boundary estimate. A future release will include
eigenvalue diagnostics in the warning
([ferx-core#17](https://github.com/FeRx-NLME/ferx-core/issues/17)).

## Parallelism

ferx uses parallelism at two independent levels. Understanding both helps you
get the most out of a multi-core machine.

### Per-subject parallelism (`threads`)

The inner loops — EBE search for FOCE/FOCEI, subject sampling for SAEM — run
in parallel across subjects on `threads` workers:

```{r}
#| eval: false
fit <- ferx_fit(ex$model, ex$data, threads = parallel::detectCores(logical = FALSE))
```

Passing `NULL` (the default) uses one worker per logical CPU.
On HPC shared nodes, cap the thread count to avoid contention:

```{r}
#| eval: false
fit <- ferx_fit(ex$model, ex$data, threads = 4L)
```

This speeds up a **single fit** by distributing subjects across cores.
It does not help when the model is stuck in a local minimum.

### Multi-start optimization (`n_starts`)

Some models have multimodal OFV surfaces where a single fit can converge to a
local minimum rather than the global optimum. This is common with:

- Michaelis-Menten (saturable) elimination — the Vmax/Km pair produces a
  narrow ridge where many combinations give similar predicted concentrations.
- Full-block omega with three or more ETAs.
- Many correlated covariate parameters.

`n_starts` runs independent fits from perturbed starting values **in parallel**
and returns the result with the lowest OFV:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(n_starts = 8L)
)
```

Start 0 always uses the exact initial values from the model file. Starts 1–7
apply a log-space perturbation of size `start_sigma` (default 0.3, roughly
±30% CV) to each theta. On an 8-core machine, 8 starts cost approximately the
same wall time as a single run.

For ridge-shaped surfaces — where the optimiser needs to explore a wider
neighbourhood — increase `start_sigma`:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    n_starts    = 8L,
    start_sigma = 0.5   # wider perturbation; default is 0.3
  )
)
```

Pin `multi_start_seed` to reproduce a specific set of perturbations:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    n_starts         = 8L,
    multi_start_seed = 123L
  )
)
```

### Combining both levels

`threads` and `n_starts` are orthogonal and compose naturally. With
`n_starts = 4` and `threads = 2`, ferx runs four fits in parallel, each
using two worker threads for its inner loops — fully utilizing an 8-core
machine:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  threads  = 2L,
  settings = list(n_starts = 4L)
)
```

::: callout-tip
**Rule of thumb:** use `threads` to speed up a single well-conditioned fit;
use `n_starts` when you suspect local minima. On a shared HPC node where you
have 8 cores, `n_starts = 8, threads = 1` is usually better than
`n_starts = 1, threads = 8` for poorly-identified models.
:::

## Reading the output

```{r}
#| label: summary
summary(fit)
```

Key fields in the printed summary:

| Field | Meaning |
|---|---|
| THETA estimate | Point estimate at convergence |
| SE | Standard error from the covariance step |
| CV% | Coefficient of variation = SE / estimate × 100 |
| 95% CI | Estimate ± 1.96 SE |
| OMEGA | Between-subject variance (diagonal elements) |
| CV~ | Approximate BSV in % = √ω × 100 |
| OFV | Objective function value (−2 log-likelihood) |
| AIC | OFV + 2 × npar |
| BIC | OFV + ln(nobs) × npar |

### Accessing individual fields

```{r}
#| label: fields
fit$theta         # named numeric vector of theta estimates
fit$omega         # omega matrix
fit$sigma         # sigma vector
fit$se_theta      # SEs for theta
fit$ofv           # scalar OFV
fit$aic
fit$bic
fit$sdtab         # individual-level diagnostics (ID, TIME, DV, PRED, IPRED, CWRES, ...)
```

## Convergence diagnostics

```{r}
#| label: trace
#| cache: true
ferx_plot_trace(fit)
```

`ferx_plot_trace()` draws the OFV and each theta/omega/sigma parameter over iterations.
A well-converged fit shows smooth descent with a stable plateau.

If you want the raw trace data:

```{r}
#| label: trace-data
tr <- ferx_trace(fit)
head(tr)
#>   iteration      ofv  TVCL   TVV  TVKA
#> 1         1  423.1   0.134  8.10  1.00
#> ...
```

## Common convergence issues

| Symptom | Likely cause | Fix |
|---|---|---|
| OFV oscillates | Step size too large | Lower `maxiter`; narrow bounds |
| `maxiter` hit without convergence | Model is misspecified | Check structural model; simplify |
| Covariance non-positive-definite | Over-parameterised | Remove insignificant ETAs or reduce model |
| Parameter hits bound | Bound too tight | Widen bounds; re-parameterise |

## Comparing models

Use OFV differences for nested models (the likelihood ratio test) or AIC/BIC for non-nested:

```{r}
#| label: compare
#| cache: true
fit_base <- ferx_fit(ex$model, ex$data)

# A model with fixed ETA_KA (simpler)
ex_no_ka_bsv <- ferx_example("warfarin")   # would use a modified model
# delta_ofv <- fit_base$ofv - fit_simpler$ofv
# p_value   <- pchisq(delta_ofv, df = 1, lower.tail = FALSE)
```

The chi-squared test with df = difference in number of parameters.
