---
title: "Fitting"
---

## Overview {#sec-fitting}

`ferx_fit()` is the central function.
It compiles the model, runs the optimiser, and returns a `ferx_fit` object with estimates, standard errors, and individual diagnostics.

```{r}
#| label: setup
#| message: false
library(ferx)
ex <- ferx_example("warfarin")
```

## Estimation methods

Pass `method` to choose the algorithm:

| Method | Description | Typical use |
|---|---|---|
| `"foce"` | First-Order Conditional Estimation | Default, fast, well-tested |
| `"focei"` | FOCE with interaction | When residual error depends on ETAs |
| `"saem"` | Stochastic Approximation EM | Models with many ETAs or strong nonlinearity |
| `"gn"` | Gauss-Newton / BHHH | Quick exploration; no covariance |
| `"gn_hybrid"` | Gauss-Newton then FOCEI polish | Fast warm start, rigorous finish |

```{r}
#| label: fit-foce
#| cache: true
fit_foce <- ferx_fit(ex$model, ex$data, method = "foce")
```

```{r}
#| label: fit-focei
#| cache: true
fit_focei <- ferx_fit(ex$model, ex$data, method = "focei")
```

### Method chains

Pass a character vector to run methods in sequence.
Each stage is seeded with the previous stage's converged parameters.
Only the final stage produces the reported covariance and diagnostics.

```{r}
#| label: fit-chain
#| eval: false
fit_chain <- ferx_fit(ex$model, ex$data, method = c("saem", "focei"))
```

A common production pattern is `c("saem", "focei")` — SAEM finds the basin of
attraction robustly, FOCEI polishes and produces the covariance step. A second
useful chain is `c("focei", "imp")`: the final `imp` stage runs importance
sampling around the FOCEI mode and reports the marginal log-likelihood and
effective sample size, which give an honest read on the Laplace
approximation. See @sec-uncertainty for the `imp` diagnostics.

### FOCE vs FOCEI

Both methods integrate over the ETA distribution using a second-order Laplace approximation.
The difference is in how the residual error is treated:

- **FOCE**: the residual function is linearised around ETA = 0 (no interaction between ETA and the error model)
- **FOCEI**: the residual function is linearised around the individual EBE — captures the interaction when `ipred` appears inside the error model

For proportional error on concentration data, FOCEI is typically more accurate.
The performance difference is small; use FOCEI by default.

## Pre-fit sanity checks

Two helpers run before the expensive fit and catch most "won't converge" problems early.

### Initial-value check

`ferx_check_init()` runs a handful of optimiser iterations from the model file's
initial estimates and reports whether the OFV is finite, whether predictions
match the observed scale, and whether the gradient is well-defined. It is much
faster than a full fit and detects sign errors, scale mismatches, and missing
columns without burning CPU time.

```{r}
#| label: check-init
#| eval: false
ferx_check_init(ex$model, ex$data)
```

### NCA-based initial estimates

If you do not have good initial estimates, `ferx_inits_from_nca()` derives them
from a per-subject non-compartmental analysis and returns starting `theta` and
`omega` values:

```{r}
#| label: inits-nca
#| eval: false
inits <- ferx_inits_from_nca(ex$model, ex$data, method = "nca")
inits$theta
inits$omega
```

The same routine is also available inside `ferx_fit()` via the `inits_from_nca`
argument; pass `"nca"`, `"nca_sweep"` (default when `inits_from_nca = TRUE`), or
`"nca_ebe"` to swap initial values in automatically.

## The covariance step

`covariance = TRUE` triggers the Hessian-based covariance matrix after convergence.
This provides standard errors, 95% confidence intervals, and the condition number.

```{r}
#| label: fit-cov
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "foce", covariance = TRUE,
                optimizer_trace = TRUE, verbose = FALSE)
```

After convergence inspect `fit$covariance_status`:

- `"computed"` — Hessian is positive-definite; SEs and 95% CIs are available in `ferx_estimates(fit)`.
- `"failed"` — covariance matrix could not be inverted or was non-positive-definite. `fit$cov_matrix` and the `se`/CI columns are `NA`. Likely causes are over-parameterisation, a boundary estimate, or a parameter that is structurally non-identifiable.

When the covariance step fails, try (a) removing one ETA at a time to find the
unidentified component, (b) checking for parameters at their bounds, or
(c) switching to SAEM, which does not depend on the Hessian for convergence.

## Parallelism

ferx uses parallelism at two independent levels. Understanding both helps you
get the most out of a multi-core machine.

### Per-subject parallelism (`threads`)

The inner loops run in parallel across subjects on `threads` workers.
This applies to all estimation methods:

| Method | What is parallelised |
|--------|----------------------|
| FOCE / FOCEI | EBE (inner) optimisation per subject |
| SAEM | Metropolis-Hastings sampling per subject |
| GN / GN hybrid | Per-subject gradient evaluations |

```{r}
#| eval: false
fit <- ferx_fit(ex$model, ex$data, threads = parallel::detectCores(logical = FALSE))
```

Passing `NULL` (the default) uses one worker per logical CPU.
On HPC shared nodes, cap the thread count to avoid contention:

```{r}
#| eval: false
fit <- ferx_fit(ex$model, ex$data, threads = 4L)
```

This speeds up a **single fit** by distributing subjects across cores.
It does not help when the model is stuck in a local minimum.

`threads` controls the inner-loop pool only. The `n_starts` outer parallelism
(see below) always uses the global rayon pool regardless of `threads`.

### Multi-start optimization (`n_starts`)

Some models have multimodal OFV surfaces where a single fit can converge to a
local minimum rather than the global optimum. This is common with:

- Michaelis-Menten (saturable) elimination — the Vmax/Km pair produces a
  narrow ridge where many combinations give similar predicted concentrations.
- Full-block omega with three or more ETAs.
- Many correlated covariate parameters.

`n_starts` runs independent fits from perturbed starting values **in parallel**
and returns the result with the lowest OFV:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(n_starts = 8L)
)
```

Start 0 always uses the exact initial values from the model file. Starts 1–7
apply a log-space perturbation of size `start_sigma` (default 0.3, roughly
±30% CV) to each theta. On an 8-core machine, 8 starts cost approximately the
same wall time as a single run.

For ridge-shaped surfaces — where the optimiser needs to explore a wider
neighbourhood — increase `start_sigma`:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    n_starts    = 8L,
    start_sigma = 0.5   # wider perturbation; default is 0.3
  )
)
```

Pin `multi_start_seed` to reproduce a specific set of perturbations:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    n_starts         = 8L,
    multi_start_seed = 123L
  )
)
```

**Method and optimizer behaviour:**

| Method | Multi-start behaviour |
|--------|-----------------------|
| FOCE / FOCEI | Fully supported; most useful — gradient-based methods are most prone to local minima |
| GN / GN hybrid | Fully supported; same local-minimum risk as FOCE |
| SAEM | Supported; each start gets an independent MH sampler seed (`saem_seed + k`) so trajectories are stochastically distinct. SAEM is inherently more robust to local minima than FOCE, but multi-start still helps on very multimodal surfaces |

`n_starts` is optimizer-agnostic: it works with any `optimizer` setting
(`slsqp`, `bobyqa`, `trust_region`, `lbfgs`, etc.) — each start runs the
full local optimisation pipeline with its own perturbed theta.

::: callout-warning
**`global_search` interaction:** when `global_search = true` and `n_starts > 1`,
the CRS2-LM gradient-free pre-search only runs on start 0. CRS2-LM ignores the
starting point by design, so running it on perturbed starts would override the
perturbation and waste cores. ferx emits a warning on `fit$warnings` when this
combination is detected.
:::

### Combining both levels

`threads` and `n_starts` are orthogonal and compose naturally. With
`n_starts = 4` and `threads = 2`, ferx runs four fits in parallel, each
using two worker threads for its inner loops — fully utilizing an 8-core
machine:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  threads  = 2L,
  settings = list(n_starts = 4L)
)
```

::: callout-tip
**Rule of thumb:** use `threads` to speed up a single well-conditioned fit;
use `n_starts` when you suspect local minima. On a shared HPC node where you
have 8 cores, `n_starts = 8, threads = 1` is usually better than
`n_starts = 1, threads = 8` for poorly-identified models.
:::

## Optimizer choice

The estimation **method** (FOCE/FOCEI/SAEM/GN) decides what likelihood is
maximised; the **optimizer** decides how. ferx exposes several optimisers
through the `[fit_options]` block (or `settings = list(optimizer = ...)`):

| Optimizer | Notes |
|---|---|
| `slsqp` (default) | Sequential quadratic programming; gradient-based, handles bounds. Robust for FOCE/FOCEI. |
| `bfgs`, `lbfgs`, `nlopt_lbfgs` | Quasi-Newton variants; sometimes faster on smooth surfaces. |
| `mma` | Method of moving asymptotes; alternative SQP-style. |
| `bobyqa` | Gradient-free; useful when finite-difference gradients are noisy. |
| `trust_region` | Trust-region with bound handling; good for poorly-scaled problems. |

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(optimizer = "trust_region")
)
```

### Global pre-search

`global_search = TRUE` runs NLopt's CRS2-LM gradient-free search to find a good
basin before the local optimiser kicks in. Useful when initial estimates are
weak (e.g. translated from a different drug) and you do not want to invoke a
full SAEM stage:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(global_search = TRUE, global_maxeval = 200)
)
```

Combine with `n_starts` (see below) only when you understand the interaction
described in the warning further down.

## Reading the output

```{r}
#| label: summary
summary(fit)
```

`summary(fit)` is the run report: convergence, method, OFV/AIC/BIC, shrinkage,
covariance status, the effective `[fit_options]`, and any warnings. The
parameter table itself (THETA / OMEGA / SIGMA point estimates + SE) is shown
by `print(fit)` or as a tidy data frame by `ferx_estimates(fit)`.

### Accessing individual fields

```{r}
#| label: fields
fit$theta         # named numeric vector of theta estimates
fit$omega         # omega matrix
fit$sigma         # sigma vector
fit$se_theta      # SEs for theta
fit$ofv           # scalar OFV
fit$aic
fit$bic
fit$sdtab         # individual-level diagnostics (ID, TIME, DV, PRED, IPRED, CWRES, ...)
```

## Convergence diagnostics

```{r}
#| label: trace
#| cache: true
ferx_plot_trace(fit)
```

`ferx_plot_trace()` draws the OFV and each theta/omega/sigma parameter over iterations.
A well-converged fit shows smooth descent with a stable plateau.

If you want the raw trace data:

```{r}
#| label: trace-data
tr <- ferx_trace(fit)
head(tr)
```

`ferx_trace()` requires the fit to have been produced with
`optimizer_trace = TRUE` (as in the fit at the top of this section); otherwise
it errors with a clear message.

## Saving and reloading a fit

For long-running fits, persist the result to disk and reload later without
re-fitting. The fit object embeds SHA-256 hashes of the model file and dataset
so `ferx_load_fit()` warns you if either has changed since the fit was saved.

```{r}
#| eval: false
ferx_save_fit(fit, "warfarin.fitrx")           # data not embedded by default
ferx_save_fit(fit, "warfarin.fitrx", include_data = TRUE)

fit_reloaded <- ferx_load_fit("warfarin.fitrx")
```

`include_data = TRUE` embeds the dataset as a base64 blob so the `.fitrx` file
is self-contained — useful for sharing a fit across machines.

## Common convergence issues

| Symptom | Likely cause | Fix |
|---|---|---|
| OFV oscillates | Step size too large | Lower `maxiter`; narrow bounds |
| `maxiter` hit without convergence | Model is misspecified | Check structural model; simplify |
| Covariance non-positive-definite | Over-parameterised | Remove insignificant ETAs or reduce model |
| Parameter hits bound | Bound too tight | Widen bounds; re-parameterise |

## Comparing models

Use OFV differences for nested models (the likelihood ratio test) or AIC/BIC for non-nested:

```{r}
#| label: compare
#| cache: true
fit_base <- ferx_fit(ex$model, ex$data)

# A model with fixed ETA_KA (simpler)
ex_no_ka_bsv <- ferx_example("warfarin")   # would use a modified model
# delta_ofv <- fit_base$ofv - fit_simpler$ofv
# p_value   <- pchisq(delta_ofv, df = 1, lower.tail = FALSE)
```

The chi-squared test with df = difference in number of parameters.
