---
title: "Fitting"
---

## Overview {#sec-fitting}

`ferx_fit()` is the central function.
It compiles the model, runs the optimiser, and returns a `ferx_fit` object with estimates, standard errors, and individual diagnostics.

```{r}
#| label: setup
#| message: false
library(ferx)
ex <- ferx_example("warfarin")
```

## Estimation methods

A fit has two layers that are worth keeping distinct from the start:

- the **method** (`foce`, `focei`, `saem`, `gn`, `gn_hybrid`, `imp`) decides
  *what* objective is maximised — how the likelihood integrates over the
  random effects;
- the **optimizer** (`bobyqa`, `slsqp`, `trust_region`, …) decides *how* that
  objective is minimised over the parameter vector.

Not every optimizer applies to every method, and each method has its own
tuning knobs. The [Optimizer and method settings](#sec-optimizer-settings)
section below lays out the full grid; this section covers the method choice
itself.

Pass `method` to choose the algorithm:

| Method | Description | Typical use |
|---|---|---|
| `"foce"` | First-Order Conditional Estimation | Default, fast, well-tested |
| `"focei"` | FOCE with interaction | When residual error depends on ETAs |
| `"saem"` | Stochastic Approximation EM | Models with many ETAs or strong nonlinearity; supports IOV |
| `"gn"` | Gauss-Newton / BHHH | Quick exploration; runs its own loop, no covariance |
| `"gn_hybrid"` | Gauss-Newton then FOCEI polish | Fast warm start, rigorous finish |
| `"imp"` | Importance-sampling marginal log-likelihood | Terminal diagnostic stage only; see @sec-uncertainty |

```{r}
#| label: fit-foce
#| cache: true
fit_foce <- ferx_fit(ex$model, ex$data, method = "foce")
```

```{r}
#| label: fit-focei
#| cache: true
fit_focei <- ferx_fit(ex$model, ex$data, method = "focei")
```

### Method chains

Pass a character vector to run methods in sequence.
Each stage is seeded with the previous stage's converged parameters.
Only the final stage produces the reported covariance and diagnostics.

```{r}
#| label: fit-chain
#| eval: false
fit_chain <- ferx_fit(ex$model, ex$data, method = c("saem", "focei"))
```

A common production pattern is `c("saem", "focei")` — SAEM finds the basin of
attraction robustly, FOCEI polishes and produces the covariance step. A second
useful chain is `c("focei", "imp")`: the final `imp` stage runs importance
sampling around the FOCEI mode and reports the marginal log-likelihood and
effective sample size, which give an honest read on the Laplace
approximation. See @sec-uncertainty for the `imp` diagnostics.

### FOCE vs FOCEI

Both methods integrate over the ETA distribution using a second-order Laplace approximation.
The difference is in how the residual error is treated:

- **FOCE**: the residual function is linearised around ETA = 0 (no interaction between ETA and the error model)
- **FOCEI**: the residual function is linearised around the individual EBE — captures the interaction when `ipred` appears inside the error model

For proportional error on concentration data, FOCEI is typically more accurate.
The performance difference is small; use FOCEI by default.

## Pre-fit sanity checks

Two helpers run before the expensive fit and catch most "won't converge" problems early.

### Initial-value check

`ferx_check_init()` runs a handful of optimiser iterations from the model
file's initial estimates and returns a list with a `$trace` data frame (per-
iteration OFV, gradient norm, step norm) and a `$summary` row (start OFV,
end OFV, OFV drop, convergence flag). A finite start OFV and a non-degenerate
gradient mean the model is set up correctly enough to attempt the full fit;
a non-finite OFV or an explosive gradient norm flags scale or specification
problems before you commit to a long run.

```{r}
#| label: check-init
#| eval: false
ferx_check_init(ex$model, ex$data)
```

### NCA-based initial estimates {#sec-nca-inits}

If you do not have good initial estimates, `ferx_inits_from_nca()` derives them
from a per-subject non-compartmental analysis and returns starting `theta` and
`omega` values:

```{r}
#| label: inits-nca
#| eval: false
inits <- ferx_inits_from_nca(ex$model, ex$data, method = "nca")
inits$theta
inits$omega
```

The same routine is also available inside `ferx_fit()` via the `inits_from_nca`
argument; pass `"nca"`, `"nca_sweep"` (default when `inits_from_nca = TRUE`), or
`"nca_ebe"` to swap initial values in automatically.

## The covariance step

`covariance = TRUE` triggers the Hessian-based covariance matrix after convergence.
This provides standard errors, 95% confidence intervals, and the condition number.

```{r}
#| label: fit-cov
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "foce", covariance = TRUE,
                optimizer_trace = TRUE, verbose = FALSE)
```

After convergence inspect `fit$covariance_status`:

- `"computed"` — Hessian is positive-definite; SEs and 95% CIs are available in `ferx_estimates(fit)`.
- `"failed"` — covariance matrix could not be inverted or was non-positive-definite. `fit$cov_matrix` and the `se`/CI columns are `NA`. Likely causes are over-parameterisation, a boundary estimate, or a parameter that is structurally non-identifiable.

When the covariance step fails, try (a) removing one ETA at a time to find the
unidentified component, (b) checking for parameters at their bounds, or
(c) switching to SAEM, which does not depend on the Hessian for convergence.

## Parallelism

ferx uses parallelism at two independent levels. Understanding both helps you
get the most out of a multi-core machine.

### Per-subject parallelism (`threads`)

The inner loops run in parallel across subjects on `threads` workers.
This applies to all estimation methods:

| Method | What is parallelised |
|--------|----------------------|
| FOCE / FOCEI | EBE (inner) optimisation per subject |
| SAEM | Metropolis-Hastings sampling per subject |
| GN / GN hybrid | Per-subject gradient evaluations |

```{r}
#| eval: false
fit <- ferx_fit(ex$model, ex$data, threads = parallel::detectCores(logical = FALSE))
```

Passing `NULL` (the default) uses one worker per logical CPU.
On HPC shared nodes, cap the thread count to avoid contention:

```{r}
#| eval: false
fit <- ferx_fit(ex$model, ex$data, threads = 4L)
```

This speeds up a **single fit** by distributing subjects across cores.
It does not help when the model is stuck in a local minimum.

`threads` controls the inner-loop pool only. The `n_starts` outer parallelism
(see below) always uses the global rayon pool regardless of `threads`.

### Multi-start optimization (`n_starts`) {#sec-multistart}

Some models have multimodal OFV surfaces where a single fit can converge to a
local minimum rather than the global optimum. This is common with:

- Michaelis-Menten (saturable) elimination — the Vmax/Km pair produces a
  narrow ridge where many combinations give similar predicted concentrations.
- Full-block omega with three or more ETAs.
- Many correlated covariate parameters.

`n_starts` runs independent fits from perturbed starting values **in parallel**
and returns the result with the lowest OFV:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(n_starts = 8L)
)
```

Start 0 always uses the exact initial values from the model file. Starts 1–7
apply a log-space perturbation of size `start_sigma` (default 0.3, roughly
±30% CV) to each theta. On an 8-core machine, 8 starts cost approximately the
same wall time as a single run.

For ridge-shaped surfaces — where the optimiser needs to explore a wider
neighbourhood — increase `start_sigma`:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    n_starts    = 8L,
    start_sigma = 0.5   # wider perturbation; default is 0.3
  )
)
```

Pin `multi_start_seed` to reproduce a specific set of perturbations:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    n_starts         = 8L,
    multi_start_seed = 123L
  )
)
```

**Method and optimizer behaviour:**

| Method | Multi-start behaviour |
|--------|-----------------------|
| FOCE / FOCEI | Fully supported; most useful — gradient-based methods are most prone to local minima |
| GN / GN hybrid | Fully supported; same local-minimum risk as FOCE |
| SAEM | Supported; each start gets an independent MH sampler seed (`saem_seed + k`) so trajectories are stochastically distinct. SAEM is inherently more robust to local minima than FOCE, but multi-start still helps on very multimodal surfaces |

`n_starts` is optimizer-agnostic: it works with any `optimizer` setting
(`slsqp`, `bobyqa`, `trust_region`, `lbfgs`, etc.) — each start runs the
full local optimisation pipeline with its own perturbed theta.

::: callout-warning
**`global_search` interaction:** when `global_search = true` and `n_starts > 1`,
the CRS2-LM gradient-free pre-search only runs on start 0. CRS2-LM ignores the
starting point by design, so running it on perturbed starts would override the
perturbation and waste cores. ferx emits a warning on `fit$warnings` when this
combination is detected.
:::

### Combining both levels

`threads` and `n_starts` are orthogonal and compose naturally. With
`n_starts = 4` and `threads = 2`, ferx runs four fits in parallel, each
using two worker threads for its inner loops — fully utilizing an 8-core
machine:

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  threads  = 2L,
  settings = list(n_starts = 4L)
)
```

::: callout-tip
**Rule of thumb:** use `threads` to speed up a single well-conditioned fit;
use `n_starts` when you suspect local minima. On a shared HPC node where you
have 8 cores, `n_starts = 8, threads = 1` is usually better than
`n_starts = 1, threads = 8` for poorly-identified models.
:::

## Optimizer and method settings {#sec-optimizer-settings}

Everything beyond the dedicated `ferx_fit()` arguments is tuned through the
`settings` list (or, equivalently, the model file's `[fit_options]` block).
Keys are validated — an unknown key, a malformed value, or a key that
duplicates a dedicated argument (`method`, `covariance`, `bloq_method`,
`threads`, `sir`) raises an error rather than being silently ignored.

**Precedence.** Dedicated arguments win over `settings`, which win over the
model file's `[fit_options]`. A `warning()` fires whenever a call-time value
overrides a *different* value baked into the model file. Inspect
`fit$call_settings` and `fit$model_file_settings` to audit what actually took
effect.

### Which settings apply to which method

The single most common source of confusion is passing an option to a method
that ignores it. This grid is the map:

| Setting | FOCE / FOCEI | GN-hybrid | pure GN | SAEM | IMP stage |
|---|:--:|:--:|:--:|:--:|:--:|
| `optimizer` | ✓ | polish only | — | — | — |
| `maxiter` | ✓ | ✓ | ✓ | — (use `n_*`) | — |
| `global_search`, `global_maxeval` | ✓ | polish only | — | — | — |
| `stagnation_guard` | ✓ | ✓ | — | — | — |
| `reconverge_gradient_interval` | ✓ | ✓ | — | — | — |
| `steihaug_max_iters` (trust-region) | ✓ | ✓ | — | — | — |
| `gn_lambda` | — | ✓ | ✓ | — | — |
| `n_exploration`, `n_convergence`, `n_mh_steps`, `adapt_interval`, `omega_burnin`, `n_leapfrog` | — | — | — | ✓ | — |
| `is_samples`, `is_proposal_df`, `is_seed`, `is_low_ess_threshold` | — | — | — | — | ✓ |
| `inner_maxiter`, `inner_tol` (shared) | ✓ | ✓ | ✓ | ✓ | — |

Setting an option that the active method ignores is harmless for most keys but
`optimizer` under `method = "saem"` emits a warning — SAEM has no single outer
minimisation, so there is no algorithm to select.

### The outer optimizer

For FOCE / FOCEI (and the FOCEI **polish** phase of `gn_hybrid`), the outer
optimizer minimises the population OFV over the transformed parameter vector
`[log θ, chol(Ω), log σ]`:

| Optimizer | Kind | Use it when |
|---|---|---|
| `bobyqa` **(default)** | Derivative-free quadratic trust-region | Start here. Re-evaluates the EBEs at every trial point, so it never sees the fixed-EBE gradient bias that stalls gradient methods on ODE/PD models, sparse data, and Hill-ridge identifiability problems. |
| `slsqp` | Gradient-based SQP | A smooth, well-conditioned analytical-PK model with many parameters, where derivative-free search needs too many samples to triangulate the surface. |
| `trust_region` | Newton trust-region (Steihaug CG) | Large θ + Ω, or paired with `inits_from_nca` (see [NCA-based initial estimates](#sec-nca-inits)) — second-order curvature helps when you are already in the right basin. |
| `lbfgs`, `nlopt_lbfgs`, `mma`, `bfgs` | Quasi-Newton / SQP variants | Rarely needed; prefer `bobyqa` or `slsqp`. |

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(optimizer = "slsqp")   # gradient-based; bobyqa is the default
)
```

::: {.callout-note}
## Why `bobyqa` is the default
The default was `slsqp` in earlier releases. On ODE/PD models and sparse data,
the cheap fixed-EBE finite-difference gradient drives `slsqp` to local minima
*hundreds* of OFV units above the true optimum. `bobyqa` doesn't use a gradient,
doesn't see the bias, and reaches the same or lower OFV in the same or less wall
time. Pure `gn` and pure `saem` ignore `optimizer` entirely.
:::

### Fixed-EBE gradient bias and `reconverge_gradient_interval`

This is the knob behind the default change, and the first thing to reach for
when a *gradient-based* optimizer stalls above an OFV you expected to beat.

By default (`reconverge_gradient_interval = 0`) the population gradient holds
each subject's inner EBE solution fixed — cheap, but it omits the response of
that inner solution to θ and Ω. On ill-conditioned non-IOV models the omitted
term is exactly what lets `slsqp` stall. Reconverging the inner loop during the
gradient recovers the full surface at roughly **5–6× the per-gradient cost**:

- `0` — never reconverge (cheap fixed-EBE gradient, the default);
- `1` — reconverge on every gradient evaluation (full surface, full cost);
- `N` — reconverge every `N`-th evaluation, cheap gradient in between, which
  often closes most of the gap for a fraction of the cost.

On a cefepime 2-compartment benchmark the trade-off looks like this:

| Configuration | OFV | Wall time |
|---|---:|---:|
| `slsqp` (fixed-EBE) | 68,252 | 390 s |
| `slsqp` + `interval = 10` | 66,118 | 633 s |
| `slsqp` + `interval = 1` | **65,485** | 1,871 s |
| `bobyqa` (default, derivative-free) | 65,598 | 315 s |

The lesson the default encodes: `bobyqa` reaches a near-optimal OFV faster than
any reconverged `slsqp` setting. Reach for reconvergence only when you *must*
use a gradient optimizer (e.g. a very high parameter count where derivative-free
search is too slow). **IOV models always reconverge and ignore this setting.**

### Letting the optimizer run: `stagnation_guard`

By default ferx short-circuits the NLopt outer loop once the OFV plateau is
numerically flat (no improvement above `1e-3` over a rolling window), so SLSQP /
L-BFGS terminate quickly instead of grinding through the full `maxiter` budget
at full inner-loop cost. Set `stagnation_guard = FALSE` to disable the guard and
let the optimizer run to its own xtol / ftol or to `maxiter` — useful when real
OFV improvements are genuinely smaller than the guard's threshold (some
γ-bearing FOCEI problems), or when debugging.

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(
    optimizer                    = "slsqp",
    reconverge_gradient_interval = 1,      # recover the full gradient surface
    stagnation_guard             = FALSE   # run to slsqp's own tolerances
  )
)
```

### Global pre-search

`global_search = TRUE` runs NLopt's CRS2-LM derivative-free search to find a
good basin *before* the local optimizer takes over — useful when initial
estimates are weak (e.g. translated from a different drug) and you would rather
not spend a full SAEM stage. It requires lower/upper bounds on every theta and
hands off to the `optimizer` once `global_maxeval` is exhausted.

```{r}
#| eval: false
fit <- ferx_fit(
  ex$model, ex$data,
  settings = list(global_search = TRUE, global_maxeval = 2000)
)
```

Combine with `n_starts` (see [Multi-start optimization](#sec-multistart)) only
when you understand the interaction described in that section's warning.

### SAEM-specific settings

SAEM controls its own iteration budget instead of `maxiter`, and exposes the
sampler internals:

| Setting | Default | What it does |
|---|---|---|
| `n_exploration` | 150 | Stochastic exploration-phase iterations. |
| `n_convergence` | 250 | Averaging / convergence-phase iterations. |
| `n_mh_steps` | 10 | Metropolis-Hastings steps per subject per iteration. |
| `adapt_interval` | 50 | How often (iterations) the MH proposal covariance is adapted. |
| `omega_burnin` | 20 | Iterations holding Ω fixed while the sampler warms up. Guards against Ω collapse on sparse data; set `0` to disable. |
| `seed` / `saem_seed` | 12345 | RNG seed for the MH sampler. |
| `n_leapfrog` / `saem_n_leapfrog` | 0 | `0` = Metropolis-Hastings; a positive value swaps in one **HMC** proposal per subject per iteration. Requires an autodiff build. |

```{r}
#| eval: false
fit_saem <- ferx_fit(
  ex$model, ex$data,
  method   = "saem",
  settings = list(n_exploration = 300, n_convergence = 400, omega_burnin = 30)
)
```

If your SAEM fit on sparse data shows a tiny `omega` paired with an inflated
additive `sigma`, raise `omega_burnin` before anything else — that pattern is
the classic Ω-collapse the burn-in is there to prevent.

### Gauss-Newton, IMP and SIR settings

| Setting | Applies to | Default | What it does |
|---|---|---|---|
| `gn_lambda` | `gn`, `gn_hybrid` | 0.01 | Levenberg-Marquardt damping; larger = more conservative steps. |
| `steihaug_max_iters` | `trust_region` | engine | CG iteration budget for the trust-region subproblem; raise on ill-conditioned problems. |
| `is_samples` | `imp` stage | 1000 | Importance samples per subject (halving the MC SE needs 4× the samples). |
| `is_proposal_df` | `imp` stage | 5 | Degrees of freedom for the Student-t proposal. |
| `is_seed` | `imp` stage | engine | RNG seed for the importance-sampling step. |
| `is_low_ess_threshold` | `imp` stage | 0.1 | ESS fraction below which a subject is flagged. |

For the SIR (`sir_samples`, `sir_resamples`, `sir_seed`) and multi-start
(`n_starts`, `start_sigma`, `multi_start_seed`) knobs, see @sec-uncertainty and
[Multi-start optimization](#sec-multistart) respectively — they have dedicated
sections rather than being repeated here.

## Reading the output

```{r}
#| label: summary
summary(fit)
```

`summary(fit)` is the run report: convergence, method, OFV/AIC/BIC, shrinkage,
covariance status, the effective `[fit_options]`, and any warnings. The
parameter table itself (THETA / OMEGA / SIGMA point estimates + SE) is shown
by `print(fit)` or as a tidy data frame by `ferx_estimates(fit)`.

### Accessing individual fields

```{r}
#| label: fields
fit$theta         # named numeric vector of theta estimates
fit$omega         # omega matrix
fit$sigma         # sigma vector
fit$se_theta      # SEs for theta
fit$ofv           # scalar OFV
fit$aic
fit$bic
fit$sdtab         # individual-level diagnostics (ID, TIME, DV, PRED, IPRED, CWRES, ...)
```

## Convergence diagnostics

```{r}
#| label: trace
#| cache: true
ferx_plot_trace(fit)
```

`ferx_plot_trace()` draws the OFV and each theta/omega/sigma parameter over iterations.
A well-converged fit shows smooth descent with a stable plateau.

If you want the raw trace data:

```{r}
#| label: trace-data
tr <- ferx_trace(fit)
head(tr)
```

`ferx_trace()` requires the fit to have been produced with
`optimizer_trace = TRUE` (as in the fit at the top of this section); otherwise
it errors with a clear message.

## Saving and reloading a fit

For long-running fits, persist the result to disk and reload later without
re-fitting. The fit object embeds SHA-256 hashes of the model file and dataset
so `ferx_load_fit()` warns you if either has changed since the fit was saved.

```{r}
#| eval: false
ferx_save_fit(fit, "warfarin.fitrx")           # data not embedded by default
ferx_save_fit(fit, "warfarin.fitrx", include_data = TRUE)

fit_reloaded <- ferx_load_fit("warfarin.fitrx")
```

`include_data = TRUE` embeds the dataset as a base64 blob so the `.fitrx` file
is self-contained — useful for sharing a fit across machines.

## Common convergence issues

| Symptom | Likely cause | Fix |
|---|---|---|
| OFV oscillates | Step size too large | Lower `maxiter`; narrow bounds |
| `maxiter` hit without convergence | Model is misspecified | Check structural model; simplify |
| Covariance non-positive-definite | Over-parameterised | Remove insignificant ETAs or reduce model |
| Parameter hits bound | Bound too tight | Widen bounds; re-parameterise |

## Comparing models

Use OFV differences for nested models (the likelihood ratio test) or AIC/BIC for non-nested:

```{r}
#| label: compare
#| cache: true
fit_base <- ferx_fit(ex$model, ex$data)

# A model with fixed ETA_KA (simpler)
ex_no_ka_bsv <- ferx_example("warfarin")   # would use a modified model
# delta_ofv <- fit_base$ofv - fit_simpler$ofv
# p_value   <- pchisq(delta_ofv, df = 1, lower.tail = FALSE)
```

The chi-squared test with df = difference in number of parameters.
