# Writing and managing model files {#sec-model-files}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

```{r}
#| label: stage-banner
#| echo: false
#| output: asis
book_stage_banner("MODEL")
```

## Where you are

The dataset from @sec-data is checked. The next step is a model file: a plain-text `.ferx` file that describes the parameters, the structural model, the residual error and the default fit settings. Keeping the model in a file separates it from R code, makes it easy to version, and lets the same file run from R or from the ferx command line.

This chapter shows how to read, create, check and edit model files from R. It uses the base model of the workflow thread, `two_cpt_oral_base`: a two-compartment oral model without covariates, for the same dataset.

## The data

The base model ships with the same dataset as @sec-data:

```{r}
#| label: example
base <- ferx_example("two_cpt_oral_base")
basename(unlist(base))
```

## Reading a model file

`ferx_model_show()` prints the file:

```{r}
#| label: show
ferx_model_show(base$model)
```

A model file is a sequence of `[blocks]`. These five appear in almost every model:

| Block | What it declares | ferx-core reference |
|---|---|---|
| `[parameters]` | Fixed effects `theta NAME(initial, lower, upper)`, between-subject variances `omega NAME ~ value`, residual error `sigma` (`(sd)` marks an initial value given as a standard deviation) | [parameters](https://ferx-nlme.github.io/ferx-core/model-file/parameters.html) |
| `[individual_parameters]` | Each subject's parameters as expressions of thetas, etas and covariates | [individual parameters](https://ferx-nlme.github.io/ferx-core/model-file/individual-parameters.html) |
| `[structural_model]` | A built-in PK model (`pk two_cpt_oral(...)`), an ODE system, or a compartment-free model | [structural model](https://ferx-nlme.github.io/ferx-core/model-file/structural-model.html) |
| `[error_model]` | The residual error model for `DV` | [error model](https://ferx-nlme.github.io/ferx-core/model-file/error-model.html) |
| `[fit_options]` | Default estimation settings; arguments to `ferx_fit()` override them | [fit options](https://ferx-nlme.github.io/ferx-core/model-file/fit-options.html) |

### What an expression may contain

The expressions in `[individual_parameters]` and `[odes]` take a fixed set of functions. Rather than reproduce the list, ask the engine: an unrecognised name is a parse error that names the whole supported set, so this stays true at whatever build you are on.

```{r}
#| label: expression-functions
fn_probe <- file.path(book_tempdir("model-files"), "unknown_function.ferx")
writeLines(sub("  CL = TVCL * exp(ETA_CL)", "  CL = TVCL * tanh(ETA_CL)",
               readLines(ferx_example("warfarin")$model), fixed = TRUE), fn_probe)
tryCatch(ferx_fit(fn_probe, ferx_example("warfarin")$data, verbose = FALSE),
         error = function(e) sub("^Error parsing model: ", "", conditionMessage(e)))
```

`logit()` and `inv_logit()` are how a parameter is kept inside (0, 1) -- @sec-variability gives the forms and @sec-first-fit shows how they are reported. Names match case-insensitively, so `EXP(...)` works as well as `exp(...)`.

This model also has a `[covariates]` block, which declares `WT` and `CRCL`. The other blocks are introduced in the chapters where they are used:

| Block | Chapter |
|---|---|
| `[data]`, `[data_selection]` | @sec-data |
| `[odes]`, `[initial_conditions]`, `[scaling]` | @sec-structural-models |
| `[covariates]`, `[covariate_model]` | @sec-covariates |
| `[mixture]` | @sec-variability |
| `[derived]`, `[output]` | @sec-tables-figures |
| `[simulation]` | @sec-simulation |
| `[binary_model]` | @sec-binary |
| `[event_model]` | @sec-time-to-event |
| `[adaptive_dosing]` | @sec-adaptive-dosing |
| `[diffusion]`, `[covariate_nn]` | @sec-experimental |
| `[markov_model]` | not available in ferx-r (@sec-reference) |

## Checking a model before fitting

`ferx_model_inspect()` parses the file without fitting and reports how ferx interprets it:

```{r}
#| label: inspect
structure_info <- ferx_model_inspect(base$model)
structure_info$theta_names
```

`ferx_model_validate()` runs the engine's parser and structural checks. With `data`, it also runs the checks that need the dataset, such as whether declared covariates are present:

```{r}
#| label: validate
check <- ferx_model_validate(base$model, data = base$data)
check$ok
```

The report is printed, and the invisible return value holds `ok`, the `model` and `data` names, and a `diagnostics` data frame. To see a failing check, misspell a block name in a copy of the model:

```{r}
#| label: validate-typo
model_dir <- book_tempdir("model-files")
typo <- file.path(model_dir, "typo.ferx")
writeLines(sub("^\\[fit_options\\]", "[fit_option]", readLines(base$model)), typo)

typo_check <- ferx_model_validate(typo)
typo_check$ok
typo_check$diagnostics[, c("severity", "code", "block", "line", "suggestion")]
```

Diagnostic codes are stable identifiers: `E_*` for errors, `W_*` for warnings. @sec-reference lists them all with a one-line meaning, and the ferx-core [check report](https://ferx-nlme.github.io/ferx-core/file-formats/check-report.html) page gives the full text of each.

## Starting a new model from a template

`ferx_model()` can scaffold a new model file from a built-in template. The templates are `"1cpt_oral"`, `"1cpt_iv"`, `"2cpt_oral"`, `"2cpt_iv"` and `"ode"`. With `print = TRUE` it only prints the skeleton:

```{r}
#| label: template-print
ferx_model(template = "2cpt_oral", print = TRUE)
```

With a `path` it writes the file and returns a `ferx_model` object. Use `edit = FALSE` to skip opening an editor, and `overwrite = TRUE` to replace an existing file:

```{r}
#| label: template-write
scaffold <- ferx_model(template = "2cpt_oral", path = file.path(model_dir, "run1.ferx"),
                       edit = FALSE, overwrite = TRUE)
scaffold
ferx_model_validate(scaffold$model, data = base$data)$ok
```

A template is a starting point. Adapt its initial estimates, bounds and blocks to your drug and data before you rely on it.

## The `ferx_model` object

`ferx_model(data, model)` bundles a model file with a dataset. `print()` on a `ferx_model` shows both paths and the parsed structure:

```{r}
#| label: model-object
m <- ferx_model(base$data, base$model)
m
```

`data` comes first so that a dataset can be piped in. `ferx_fit()` accepts the object directly and takes the data from it:

```{r}
#| label: model-object-fit
fit <- base$data |> ferx_model(base$model) |> ferx_fit(verbose = FALSE)
fit$converged
```

Fitting itself is the topic of @sec-first-fit.

### Piping through an analysis

The argument order across the package is chosen so that a whole step reads as one pipeline: `ferx_model()` takes the data first, and the accessors that summarise a fit take the fit first. So a dataset can go to a fitted model and on to its warnings without naming an intermediate object:

```{r}
#| label: pipeline
base$data |>
  ferx_model(base$model) |>
  ferx_fit(verbose = FALSE) |>
  ferx_get_warnings(as_df = TRUE) |>
  select(severity, category)
```

The same holds for the other fit accessors, and for simulation and prediction, where the model comes first and a fit is passed by name:

```{r}
#| label: pipeline-more
fit |> check_diagnostics() |> getElement("shrinkage") |> head(3)
base$model |> ferx_simulate(base$data, n_sim = 1, seed = 1) |> head(3)
```

Which object pipes in is decided by the name of each function's first argument, as the argument table at the end of this chapter records. The four `ferx_model_*` helpers whose first argument is `path` -- `ferx_model_show()`, `ferx_model_inspect()`, `ferx_model_validate()` and `ferx_model_edit()` -- read a model *file*, so a path pipes into them but a `ferx_model` object does not. The two whose first argument is `x`, `ferx_model_get_section()` and `ferx_model_set_section()`, take either.

```{r}
#| label: pipeline-validate
try(m |> ferx_model_validate())
m |> ferx_model_get_section("error_model", strip = TRUE)
```

## Editing model files from R

### Reading and replacing a block

`ferx_model_get_section()` returns the lines of one block. It prints them and returns them invisibly; `strip = TRUE` trims leading whitespace:

```{r}
#| label: get-section
param_lines <- ferx_model_get_section(m, "parameters", strip = TRUE)
head(param_lines, 5)
```

`ferx_model_set_section()` replaces the body of a block with new `lines`. Given a `ferx_model` that points to a bundled example, it first copies the file to a temporary directory and edits the copy. The installed example stays unchanged:

```{r}
#| label: set-section
m_focei <- ferx_model_set_section(m, "fit_options", c(
  "  method     = focei",
  "  maxiter    = 300",
  "  covariance = true"
))
m_focei$model == m$model
ferx_model_get_section(m_focei, "fit_options")
```

### Opening a model in an editor

`ferx_model_edit()` opens a model file in your editor. A bundled example is copied to `dest` first (default: the working directory), and `overwrite` controls whether an existing copy may be replaced. `save_as` copies the file after editing, which is a simple way to keep numbered versions of a model (`run1.ferx`, `run2.ferx`, …). The `.editor` argument replaces the editor function. The book passes a function that does nothing, so the example can run while the book is built:

```{r}
#| label: edit
no_editor <- function(...) invisible(NULL)
my_copy <- ferx_model_edit(base$model, dest = model_dir, overwrite = TRUE, .editor = no_editor)
run2 <- ferx_model_edit(my_copy, save_as = file.path(model_dir, "run2.ferx"), .editor = no_editor)
basename(c(my_copy, run2))
```

In an interactive session, call `ferx_model_edit(path)` without `.editor`.

## Options that matter

| Function | Argument | Meaning |
|---|---|---|
| `ferx_model_show()` | `path` | Model file to print |
| `ferx_model_inspect()` | `path` | Model file, or a fit (reads its stored structure) |
| `ferx_model_validate()` | `path` | Model file (must exist and end in `.ferx`) |
| | `data` | Optional dataset for data-dependent checks |
| `ferx_model()` | `data` | Dataset path; if missing, the model's `[data]` block is used at fit time |
| | `model` | Existing model file to wrap |
| | `template` | Scaffold from `"1cpt_oral"`, `"1cpt_iv"`, `"2cpt_oral"`, `"2cpt_iv"` or `"ode"` |
| | `path` | Where to write the scaffold |
| | `overwrite` | Replace an existing file at `path` |
| | `edit` | Open the new file in an editor (default `TRUE`) |
| | `print` | Print the skeleton only |
| `ferx_model_get_section()` | `x`, `section`, `strip` | Model object or path, block name without brackets, trim whitespace |
| `ferx_model_set_section()` | `x`, `section`, `lines` | Model object or path, block name, replacement body lines |
| `ferx_model_edit()` | `path`, `dest`, `overwrite`, `save_as`, `.editor` | File, copy destination, replace existing copy, save a copy after editing, editor function |

## Pitfalls

- **`ferx_model_set_section()` on a plain path edits that file in place.** Copy-on-write only applies to a `ferx_model` object pointing into the installed package. Never pass `ferx_example(...)$model` itself as a path. Wrap it with `ferx_model()` or copy it first.
- **`ferx_model_set_section()` replaces an existing block. It cannot add a new one.** To add a block such as `[data]`, append the lines to a copy of the file (@sec-data does this).
- **Argument order of `ferx_model()`.** `data` is the first argument. An old-style call `ferx_model("model.ferx")` still works but warns about the changed order. Name `model =` to be explicit.

## Summary

- Read a model with `ferx_model_show()`. Check how it is parsed with `ferx_model_inspect()`, and validate it (optionally against the data) with `ferx_model_validate()`.
- Start new models from `ferx_model(template = ...)`. Keep versions with `ferx_model_edit(save_as = ...)`.
- Edit blocks programmatically with `ferx_model_get_section()` and `ferx_model_set_section()`, on a `ferx_model` object or a copy.

Next: @sec-first-fit fits this base model.

::: {.callout-tip}
## Reference
- R help: `?ferx_model`, `?ferx_model_show`, `?ferx_model_inspect`, `?ferx_model_validate`, `?ferx_model_get_section`, `?ferx_model_set_section`, `?ferx_model_edit`
- ferx-core: [model file reference](https://ferx-nlme.github.io/ferx-core/model-file/index.html), [check report codes](https://ferx-nlme.github.io/ferx-core/file-formats/check-report.html)
:::
