---
title: "Diagnostics"
---

## Overview {#sec-diagnostics}

After fitting, the `$sdtab` data frame contains everything needed for standard goodness-of-fit diagnostics.

```{r}
#| label: setup
#| message: false
library(ferx)
library(ggplot2)
library(dplyr)

ex  <- ferx_example("warfarin")
```

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
sdtab <- fit$sdtab
```

## The `sdtab` columns

```{r}
#| label: sdtab-cols
names(sdtab)
#> [1] "ID"    "TIME"  "DV"    "PRED"  "IPRED" "CWRES" "IWRES" "EBE_OFV" "N_OBS"
names(fit$ebe_etas)
#> [1] "ID"    "ETA_CL" "ETA_V" "ETA_KA"
```

`sdtab` contains per-observation diagnostics.
Per-subject ETAs are in `fit$ebe_etas` (one row per subject).

| Column | Location | Description |
|---|---|---|
| `PRED` | `sdtab` | Population prediction (eta = 0) |
| `IPRED` | `sdtab` | Individual prediction (fitted ETAs) |
| `CWRES` | `sdtab` | Conditional weighted residual |
| `IWRES` | `sdtab` | Individual weighted residual |
| `ETA_*` | `fit$ebe_etas` | Empirical Bayes estimate (EBE) for each BSV parameter |

## Core GOF plots

### IPRED vs DV

```{r}
#| label: ipred-dv
#| fig-cap: "Individual predictions vs observed DV. Points should scatter symmetrically around the line of identity."
ggplot(sdtab, aes(x = IPRED, y = DV)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, colour = "#1F6B7A", linewidth = 1) +
  labs(x = "IPRED", y = "Observed DV") +
  theme_minimal()
```

### CWRES vs PRED

```{r}
#| label: cwres-pred
#| fig-cap: "CWRES vs population prediction. Should be centred near 0 with ≥95% of points within ±2."
ggplot(sdtab, aes(x = PRED, y = CWRES)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0,      colour = "#1F6B7A", linewidth = 1) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "#C8893E") +
  labs(x = "PRED", y = "CWRES") +
  theme_minimal()
```

### CWRES vs TIME

```{r}
#| label: cwres-time
ggplot(sdtab, aes(x = TIME, y = CWRES)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0,      colour = "#1F6B7A", linewidth = 1) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "#C8893E") +
  labs(x = "TIME (h)", y = "CWRES") +
  theme_minimal()
```

### PRED vs DV

```{r}
#| label: pred-dv
ggplot(sdtab, aes(x = PRED, y = DV)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, colour = "#1F6B7A", linewidth = 1) +
  labs(x = "PRED", y = "Observed DV") +
  theme_minimal()
```

## ETA shrinkage

Shrinkage measures how much the individual ETAs are pulled toward zero relative to the omega variance.
High shrinkage (> 30%) means the data are not informative enough to reliably estimate individual deviations, and ETA-based diagnostics become unreliable.

```{r}
#| label: shrinkage
summary(fit)$shrinkage_eta
#>  ETA_CL   ETA_V  ETA_KA
#>   0.142   0.283   0.452
```

ETA shrinkage > 0.3 is a warning sign.
It does not mean the model is wrong, but ETA-based covariate plots and individual parameter histograms should be interpreted cautiously.

## ETA distribution

ETAs should be approximately normally distributed around zero.
Per-subject EBEs are stored in `fit$ebe_etas` (one row per subject).

```{r}
#| label: eta-dist
#| fig-cap: "Distribution of empirical Bayes estimates. Roughly normal with mean near 0."
etas_long <- fit$ebe_etas |>
  tidyr::pivot_longer(-ID, names_to = "eta", values_to = "value")

ggplot(etas_long, aes(x = value)) +
  geom_histogram(bins = 20, fill = "#1F6B7A", alpha = 0.7) +
  facet_wrap(~eta, scales = "free") +
  labs(x = "EBE", y = "Count") +
  theme_minimal()
```

## ETA-covariate correlations

`ferx_eta_cov()` computes Pearson correlations between ETAs and subject-level covariates.

```{r}
#| label: eta-cov
obs <- read.csv(ex$data)
ferx_eta_cov(fit, obs)
#>    eta      covariate     r   p_val flag
#>  ETA_CL   WT          0.42  0.016  *
#>  ETA_V    WT          0.38  0.031  *
#>  ...
```

Significant correlations (p < 0.05) flag covariates worth testing in the structural model.

## Parameter correlation matrix

After a covariance step, inspect correlations between estimated parameters.
Off-diagonal values near ±1 indicate structural identifiability problems.

```{r}
#| label: cor-matrix
if (!is.null(fit$cov_matrix)) {
  ferx_cor_matrix(fit)
} else {
  message("Covariance matrix not available (covariance step may have failed).")
}
```

## Parameter estimates table

`ferx_estimates()` returns a tidy data frame for programmatic use or table export:

```{r}
#| label: estimates
ferx_estimates(fit)
```

## Epsilon shrinkage

Epsilon shrinkage measures compression in the residual distribution:

```{r}
#| label: eps-shrinkage
summary(fit)$shrinkage_eps
#>  PROP_ERR
#>   0.094
```

Low epsilon shrinkage (< 20%) is generally acceptable.

## IWRES autocorrelation (`check_diagnostics`)

Residual autocorrelation — IWRES values within a subject that are systematically positive then negative (or vice versa) — is a sign that the structural model is missing dynamics.
ferx quantifies this with two pooled statistics computed across all subjects:

| Statistic | Meaning |
|---|---|
| Durbin-Watson (DW) | 2 = no autocorrelation; < 1.5 = positive; > 2.5 = negative |
| Lag-1 Pearson r | Direction and magnitude of serial correlation |

`check_diagnostics()` returns both in a tidy list alongside shrinkage:

```{r}
#| label: check-diagnostics
diag <- check_diagnostics(fit)
diag$autocorrelation
#>   dw_statistic  lag1_r                  flag
#>           1.84  0.0821  no autocorrelation
diag$shrinkage
#>    param type shrinkage shrinkage_pct
#>   ETA_CL  eta     0.142          14.2
#>    ETA_V  eta     0.283          28.3
#>   ETA_KA  eta     0.452          45.2
#>      EPS  eps     0.094           9.4
```

`print(fit)` shows the same values in a `--- Diagnostics ---` block, and a `message()` is emitted automatically when DW < 1.5 or DW > 2.5 with ordered remedies.

### Interpreting the DW statistic

| DW range | Interpretation | Suggested action |
|---|---|---|
| < 1.5 | Positive autocorrelation | Check time ordering; add transit compartment; consider IOV on ka/F; SDE |
| 1.5 – 2.5 | No concern | — |
| > 2.5 | Negative autocorrelation | Check for over-parameterisation or misspecified error model |

### IWRES vs time plot

```{r}
#| label: iwres-time
#| fig-cap: "IWRES vs time coloured by subject. Systematic patterns within subjects indicate missing dynamics."
if (!is.null(diag$autocorrelation)) {
  dw_val <- diag$autocorrelation$dw_statistic
  r_val  <- diag$autocorrelation$lag1_r
  ggplot(sdtab, aes(x = TIME, y = IWRES, colour = factor(ID))) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_hline(yintercept = 0, colour = "#1F6B7A", linewidth = 1) +
    labs(
      x = "TIME (h)", y = "IWRES", colour = "Subject",
      title = sprintf("IWRES vs Time  |  DW = %.2f  lag-1 r = %.3f", dw_val, r_val)
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}
```
