---
aliases:
  - /chapters/07-parameter-transforms.html
---

# Initial estimates and a first fit {#sec-first-fit}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

```{r}
#| label: stage-banner
#| echo: false
#| output: asis
book_stage_banner("ESTIMATE")
```

## Where you are

The dataset is checked (@sec-data) and the base model is written and validated (@sec-model-files). This chapter produces the first fit. It covers where starting values come from, how to fit, what the fit object contains, and how to read the warnings ferx attaches to it. Choosing and tuning estimation methods follows in @sec-estimation-methods.

## The data

The base model and its dataset from the previous chapters:

```{r}
#| label: example
base <- ferx_example("two_cpt_oral_base")
```

## Starting values

### From the model file

Every `theta`, `omega` and `sigma` in `[parameters]` carries an initial value. For thetas it is the first number in `theta NAME(initial, lower, upper)`:

```{r}
#| label: model-inits
init_lines <- ferx_model_get_section(base$model, "parameters")
```

### From the data: NCA-based starting values

`ferx_inits_from_nca()` derives starting values from the data using non-compartmental analysis (NCA). It returns them for inspection without fitting. The `method` is `"nca_sweep"` (default), `"nca"` or `"nca_ebe"`. They differ in how much refinement runs on top of the NCA estimates; `?ferx_inits_from_nca` describes each.

```{r}
#| label: nca-inits
nca <- ferx_inits_from_nca(base$model, base$data)
nca
```

The result is a `ferx_inits` object, and `print()` shows the suggested thetas, the omega matrix and any notes. On this dataset all three methods return the same thetas:

```{r}
#| label: nca-methods
identical(nca$theta, ferx_inits_from_nca(base$model, base$data, method = "nca")$theta) &&
  identical(nca$theta, ferx_inits_from_nca(base$model, base$data, method = "nca_ebe")$theta)
```

To fit from these values directly, pass `inits_from_nca` to `ferx_fit()`: `TRUE`, or one of the method names. The same option exists as the `inits_from_nca` key in `[fit_options]` or `settings`. The *Variants* section below shows that this is not always an improvement.

### A pilot run

`ferx_check_init()` runs a short pilot fit with the optimizer trace switched on. It returns the `fit`, the `trace` (one row per objective evaluation) and a one-row `summary`:

```{r}
#| label: check-init
pilot <- ferx_check_init(base$model, base$data)
pilot$summary
```

The two R warnings come from `ferx_check_init()` itself. It overrides `covariance` and `maxiter` from the model file, and ferx reports every such override. A falling objective function value (OFV) shows that the optimizer moves away from the starting values in a useful direction. On this model the pilot already reached convergence. The trace shows the path:

```{r}
#| label: fig-pilot-trace
#| fig-cap: "OFV at each objective evaluation of the pilot run."
ggplot(pilot$trace, aes(iter, ofv)) +
  geom_line() +
  coord_cartesian(ylim = range(pilot$trace$ofv[pilot$trace$iter > 10])) +
  labs(x = "Objective evaluation", y = "OFV")
```

## Minimal runnable call

```{r}
#| label: fit
fit <- ferx_fit(base$model, base$data, verbose = FALSE)
fit
```

`print()` on the fit gives one screen with everything you need to judge it:

- **Header**: method, gradient, number of subjects and observations, convergence status and OFV/AIC/BIC.
- **Model structure**, as ferx derived it from the file.
- **THETA**, **OMEGA** and **SIGMA** estimates with standard errors. Omegas are variances; for log-normal etas the CV% is shown alongside.
- **Shrinkage**, the **diagnostics** line (covariance step, condition number, Durbin-Watson statistic of the residuals) and run information.
- **Settings**: every setting, and whether it came from the model file or the call.
- A **warning count**.

The fit used `method = focei` and `gradient = fd` from the model's `[fit_options]`, marked `[model only]`.

## Reading the result

### `summary()`

`summary()` returns a `ferx_summary` object. Its print method shows run metadata and all warnings in one compact block, without the parameter tables:

```{r}
#| label: summary
summary(fit)
```

The last warning line, the number of EBE fallbacks, comes from `fit$total_ebe_fallbacks` (`r fit$total_ebe_fallbacks` here). It is a run statistic rather than a structured fit warning, so `ferx_get_warnings()` below does not list it.

### Parameter estimates

`fit$estimates` is the tidy table of all estimated parameters. The row names are the parameter names:

```{r}
#| label: estimates
fit$estimates[, c("transform", "estimate", "se", "rse_pct", "lower_95", "upper_95")]
```

`ferx_coef()` and `ferx_se()` pull values by `param` name. Unlike indexing the table, a misspelt name is an error that suggests the closest match:

```{r}
#| label: coef-se
ferx_coef(fit, c("TVCL", "TVV1"))
ferx_se(fit, c("TVCL", "TVV1"))
try(ferx_coef(fit, "TVCLL"))
```

### Parameter transformations and the `transform` column

A parameter is not always reported on the scale you want to read. The `transform` column of `fit$estimates` names the **parameterisation** the row came from, which is what you need in order to know whether `estimate` is the number you want:

| `transform` | `estimate` holds | Which parameters |
|---|---|---|
| `identity` | the value itself | a `theta` written as `TVCL * exp(ETA_CL)` |
| `log` | the log of the value | a `theta` written as `exp(TVCL + ETA_CL)` |
| `logit` | the logit of the value | a `theta` written as `inv_logit(THETA_F + ETA_F)` |
| `logit_probability` | the value itself, on (0, 1) | a `theta` declared on the (0, 1) scale and used as `inv_logit(logit(THETA_F) + ETA_F)` |
| `variance` | a variance | `omega` and `kappa` diagonals |
| `proportional` | a coefficient of variation on the fraction scale | a proportional `sigma` |
| `additive` | a standard deviation, in the units the error model sees | an additive `sigma` |

The first four are the transforms a fixed effect can carry, and `fit$theta_transforms` lists them for the thetas alone. Bounds do not appear here: this column reflects how the parameter is written in `[individual_parameters]`, not the internal scale the optimizer packs it onto, and every bounded theta of this model reads `identity`:

```{r}
#| label: theta-transforms
fit$theta_transforms
```

`log` and `logit` put `estimate` on the transformed scale and the interpretable value in `estimate_natural`, with `lower_95_natural` and `upper_95_natural` beside it. Those columns are `NA` for `identity`, `variance`, `proportional` and `additive` rows, which need no back-transform. `logit_probability` needs none either -- the theta is declared on (0, 1) and the engine returns it there, so `estimate` is already the number you want and `estimate_natural` repeats it. The `bioavailability` example (@sec-absorption) uses that form:

```{r}
#| label: natural-scale
fit_logit <- ferx_fit(ferx_example("bioavailability")$model,
                      ferx_example("bioavailability")$data, verbose = FALSE)
fit_logit$estimates[, c("transform", "estimate", "lower_95", "estimate_natural", "lower_95_natural")]
```

`THETA_F` is the typical bioavailability, `r round(fit_logit$estimates["THETA_F", "estimate"], 3)`. The model file declares it on that scale (`theta THETA_F(0.70, 0.001, 0.999)`), and the engine's own individual parameters agree: reconstructing each subject's `F` from `estimate` reproduces them exactly.

```{r}
#| label: logit-probability-check
typical_f <- fit_logit$estimates["THETA_F", "estimate"]
eta_f <- fit_logit$ebe_etas$ETA_F
reconstructed <- 1 / (1 + exp(-(log(typical_f / (1 - typical_f)) + eta_f)))
max(abs(reconstructed - fit_logit$individual_estimates$F))
```

For a `log` theta the back-transform is real, and `estimate_natural` is exactly `exp(estimate)`:

```{r}
#| label: log-back-transform
log_model <- file.path(book_tempdir("first-fit"), "log_theta.ferx")
writeLines(sub("  CL = TVCL * exp(ETA_CL)", "  CL = exp(TVCL + ETA_CL)",
               sub("  theta TVCL(0.134, 0.001, 10.0)", "  theta TVCL(-2.0, -7.0, 2.3)",
                   readLines(ferx_example("warfarin")$model), fixed = TRUE),
               fixed = TRUE), log_model)
fit_log <- ferx_fit(log_model, ferx_example("warfarin")$data, verbose = FALSE)
log_row <- fit_log$estimates["TVCL", ]
c(transform = log_row$transform,
  difference = log_row$estimate_natural - exp(log_row$estimate))
```

The two confidence intervals on a row are built differently. `lower_95` and `upper_95` are always the symmetric interval `estimate ± 1.96 × se` on the scale `estimate` is reported on. On a bounded parameter that interval is not held inside the bounds, so for a probability read it as a measure of precision rather than as limits. For a `logit_probability` theta, `lower_95_natural` and `upper_95_natural` are formed on the logit scale instead -- the standard error is carried there by the delta method, `se / (p (1 - p))`, and the interval brought back with `inv_logit()` -- so they cannot leave [0, 1], and reach its edge only for an estimate close to a bound with a large standard error:

```{r}
#| label: logit-probability-interval
fit_logit$estimates["THETA_F", c("estimate", "lower_95", "upper_95",
                                 "lower_95_natural", "upper_95_natural")]
```

Here `upper_95` is `r round(fit_logit$estimates["THETA_F", "upper_95"], 3)`, a bioavailability above 1, while the natural-scale interval runs from `r round(fit_logit$estimates["THETA_F", "lower_95_natural"], 3)` to `r round(fit_logit$estimates["THETA_F", "upper_95_natural"], 3)`; for a probability, quote that one. `print()` shows it on a `(95% CI)` line under the theta.

That interval is wide for a reason unrelated to how it is built: with oral data only, `F`, `CL` and `V` cannot be separated (@sec-absorption), and the fit says so:

```{r}
#| label: logit-probability-warnings
ferx_get_warnings(fit_logit, as_df = TRUE)[, c("category", "severity")]
```

Declaring the same bioavailability on a `logit` theta instead, in a separate fit, gives nearly the same natural-scale columns. The agreement is approximate -- the delta method is a first-order approximation, and two fits stop at slightly different optima -- and only the natural-scale columns are comparable: `estimate`, `lower_95` and `upper_95` are on different scales in the two rows.

```{r}
#| label: logit-twin
logit_model <- file.path(book_tempdir("first-fit"), "bioavailability_logit.ferx")
bio_lines <- readLines(ferx_example("bioavailability")$model)
twin_lines <- sub("theta THETA_F(0.70, 0.001, 0.999)  # typical bioavailability, directly on (0,1) scale",
                  "theta THETA_F(0.847, -5.0, 5.0)  # typical bioavailability on the logit scale",
                  bio_lines, fixed = TRUE)
twin_lines <- sub("inv_logit(logit(THETA_F) + ETA_F)", "inv_logit(THETA_F + ETA_F)",
                  twin_lines, fixed = TRUE)
stopifnot(sum(twin_lines != bio_lines) == 2)   # both lines really were rewritten
writeLines(twin_lines, logit_model)
fit_twin <- ferx_fit(logit_model, ferx_example("bioavailability")$data, verbose = FALSE)
natural <- c("transform", "estimate_natural", "lower_95_natural", "upper_95_natural")
natural_gap <- max(abs(unlist(fit_logit$estimates["THETA_F", natural[-1]]) -
                       unlist(fit_twin$estimates["THETA_F", natural[-1]])))
rbind(declared_on_0_1 = fit_logit$estimates["THETA_F", natural],
      declared_as_logit = fit_twin$estimates["THETA_F", natural])
```

The largest difference between the two rows' natural-scale columns is `r signif(natural_gap, 2)`.

For the random effects, `fit$eta_param_types` says how each eta enters the model -- `log_normal`, `additive`, `logit`, `logit_probability` or `custom`:

```{r}
#| label: eta-param-types
setNames(fit$eta_param_types, rownames(fit$omega))
```

`fit$eta_log_transformed` is a narrower flag and not a substitute: it is `TRUE` only for the `TVCL * exp(ETA_CL)` form, so an eta written `exp(TVCL + ETA_CL)` is log-normal and still reads `FALSE`.

Nothing here converts for you. An omega row is a variance, so a between-subject CV% is yours to compute (@sec-tables-figures), and a log-transform-both-sides model reports an `additive` sigma on the log scale rather than in the units of `DV` (@sec-variability).

### The fit object

A fit is a named list with `r length(fit)` elements. These are the ones you will use most:

| What you want | Where |
|---|---|
| Parameter table | `fit$estimates` |
| Residuals, `PRED`, `IPRED` per observation | `fit$sdtab` (@sec-diagnostics) |
| Individual parameters per subject | `fit$individual_estimates` |
| Random effects per subject | `fit$ebe_etas` |
| OFV, AIC, BIC, convergence | `fit$ofv`, `fit$aic`, `fit$bic`, `fit$converged` |
| Shrinkage | `fit$shrinkage_eta`, `fit$shrinkage_eps` |
| Parameter correlations | `fit$cor_matrix` |

```{r}
#| label: individual
head(fit$individual_estimates)
```

`?ferx_fit` documents every element. @sec-reference lists them all with the chapter that uses each one.

### Warnings

ferx classifies everything it notices during a fit. Each warning has a **severity** and a stable **category** token:

| Severity | Meaning |
|---|---|
| critical | The result cannot be trusted as it is (for example no convergence or an ill-conditioned covariance) |
| warning | The result stands, but a caveat applies |
| info | A note; no action needed |

`ferx_get_warnings()` prints them grouped by severity. With `as_df = TRUE` it returns a data frame you can filter:

```{r}
#| label: warnings
ferx_get_warnings(fit)
ferx_get_warnings(fit, as_df = TRUE)[, c("severity", "category")]
```

`fit$warnings` holds the same messages as plain text. Because the category tokens are stable, scripts can branch on them. For example, reject any fit with a critical warning. `check_diagnostics()` returns the residual autocorrelation and shrinkage behind two common warnings as data frames:

```{r}
#| label: check-diagnostics
check_diagnostics(fit)
```

What these diagnostics mean and how to act on them is the subject of @sec-diagnostics.

## Options that matter

Arguments of `ferx_fit()` introduced here. A call-time value always overrides the model file's `[fit_options]`, and ferx reports the override:

| Argument | Default | Meaning |
|---|---|---|
| `model` | — | Model file or `ferx_model` object |
| `data` | `NULL` | Dataset; `NULL` uses the `ferx_model` object's data or the model's `[data]` block |
| `method` | `NULL` | Estimation method(s); `NULL` uses the model file, otherwise `"focei"` (@sec-estimation-methods) |
| `covariance` | `NULL` | Run the covariance step (standard errors); `NULL` uses the model file, which defaults to `TRUE` |
| `verbose` | `NULL` | Print progress; `NULL` uses the model file, which defaults to `TRUE` |
| `inits_from_nca` | `FALSE` | Replace starting values with NCA-based ones (`TRUE` or a method name) |

The other `ferx_fit()` arguments are covered where they matter: `threads`, `gradient`, `optimizer_trace` and `settings` in @sec-estimation-methods; `ignore`, `accept` and `ignore_ids` in @sec-data; `bloq_method` in @sec-censoring; `mu_referencing` and `scale_params` in @sec-variability; `sir` and `fd_hessian_step` in @sec-uncertainty; `output` and `include_data` in @sec-reproducibility.

For example, skipping the covariance step:

```{r}
#| label: no-covariance
fit_nocov <- ferx_fit(base$model, base$data, covariance = FALSE, verbose = FALSE)
fit_nocov$covariance_status
head(fit_nocov$estimates[, c("estimate", "se")], 3)
```

| Function | Arguments |
|---|---|
| `ferx_inits_from_nca()` | `model`, `data`, `method` (`"nca_sweep"`, `"nca"`, `"nca_ebe"`) |
| `ferx_check_init()` | `model`, `data`, `method` (default `"focei"`), `maxiter` (pilot iteration limit), further arguments passed to `ferx_fit()` |
| `ferx_coef()`, `ferx_se()` | `fit`, `param` (names; `NULL` for all) |
| `ferx_get_warnings()` | `fit`, `as_df` |
| `check_diagnostics()` | `fit` |

## Variants

### Starting from NCA-based values

Fit the base model three ways: from the model file's starting values, from NCA-based starting values, and from NCA-based values with `gradient = "auto"` instead of the model file's finite differences:

```{r}
#| label: tbl-nca-fits
#| tbl-cap: "The same model from different starting values."
fit_nca <- ferx_fit(base$model, base$data, inits_from_nca = TRUE, verbose = FALSE)
fit_nca_auto <- ferx_fit(base$model, base$data, inits_from_nca = TRUE, gradient = "auto",
                         verbose = FALSE)

worst <- function(f) {
  sev <- ferx_get_warnings(f, as_df = TRUE)$severity
  if ("critical" %in% sev) "critical" else if ("warning" %in% sev) "warning" else "none"
}
data.frame(
  start = c("model file", "NCA", "NCA, gradient = auto"),
  converged = c(fit$converged, fit_nca$converged, fit_nca_auto$converged),
  ofv = c(fit$ofv, fit_nca$ofv, fit_nca_auto$ofv),
  worst_warning = c(worst(fit), worst(fit_nca), worst(fit_nca_auto))
)
```

All three report `converged = TRUE` (with `gradient = "auto"`, ferx used the `r fit_nca_auto$gradient_used` gradient). The NCA start with finite differences, however, ended at an OFV about `r round(fit_nca$ofv - fit$ofv)` points higher than the other two, with a critical warning. Its warnings show why the result cannot be trusted:

```{r}
#| label: nca-warnings
nca_warnings <- ferx_get_warnings(fit_nca, as_df = TRUE)
nca_warnings[, c("severity", "category")]
# The paragraph below names this category, so its absence has to fail the render.
stopifnot("ebe_start_dependent" %in% nca_warnings$category)
```

The NCA-based absorption rate constant (`TVKA` = `r round(nca$theta[["TVKA"]], 3)`) is far from the estimate of the successful fits (`r round(ferx_coef(fit, "TVKA"), 3)`). Convergence only means the optimizer stopped. Always compare the OFV and read the warnings before accepting a fit.

`ebe_start_dependent` in that list is a second reason to distrust this one: the random effects it reports depend on where the inner loop started, so every per-subject diagnostic built on them (@sec-diagnostics) inherits that dependence. @sec-estimation-methods shows what to change.

## Pitfalls

**A fit stopped by its iteration limit.** `maxiter` limits the outer optimizer iterations. A fit that hits the limit is marked unconverged with a critical `convergence` warning. For the warfarin example with `method = "focei"`:

```{r}
#| label: maxiter-stop
warf <- ferx_example("warfarin")
fit_short <- ferx_fit(warf$model, warf$data, method = "focei",
                      settings = list(maxiter = 5), verbose = FALSE)
fit_short$converged
ferx_get_warnings(fit_short, as_df = TRUE)[, c("severity", "category", "message")]
```

**Other points**:

- `n_iterations` and the rows of the optimizer trace count objective evaluations, not outer iterations, so they can be much larger than `maxiter`.
- Reading `fit$estimates["NAME", ]` with a misspelt name silently returns `NA`. `ferx_coef()` and `ferx_se()` stop with an error instead.

### Warnings you may see here

- `convergence` (critical): the optimizer did not converge. Reject the fit, then change starting values, method or bounds and refit. See [convergence and covariance warnings](https://ferx-nlme.github.io/ferx-core/warnings.html#codes-convergence).
- `general` (warning): the fallback category for messages not assigned to a specific code. Examples in this book are notes from NCA starting values, the location of an optimizer trace file, a stopped fit, and a misspelt data-selection column (@sec-data). Read the message. See [solver, numerics and unclassified warnings](https://ferx-nlme.github.io/ferx-core/warnings.html#codes-solver).

## Summary

- Starting values come from `[parameters]`. `ferx_inits_from_nca()` proposes data-based ones, and `ferx_check_init()` runs a pilot.
- `ferx_fit(model, data)` fits. `print()` and `summary()` show the result, and `fit$estimates`, `ferx_coef()` and `ferx_se()` give the numbers.
- Judge a fit by convergence, OFV **and** warnings. `ferx_get_warnings(as_df = TRUE)` makes them scriptable.

Next: @sec-estimation-methods chooses and tunes the estimation method.

::: {.callout-tip}
## Reference
- R help: `?ferx_fit`, `?summary.ferx_fit`, `?ferx_inits_from_nca`, `?ferx_check_init`, `?ferx_coef`, `?ferx_get_warnings`, `?check_diagnostics`
- ferx-core: [fit options](https://ferx-nlme.github.io/ferx-core/model-file/fit-options.html), [warnings](https://ferx-nlme.github.io/ferx-core/warnings.html)
:::
