---
title: "Simulation & VPC"
---

## Overview {#sec-simulation}

`ferx_simulate()` generates replicated datasets from the fitted model.
The primary use case is the **Visual Predictive Check (VPC)**: overlay observed data on simulation-based prediction intervals to assess model adequacy visually.

```{r}
#| label: setup
#| message: false
library(ferx)
library(ggplot2)
library(dplyr)

ex  <- ferx_example("warfarin")
obs <- read.csv(ex$data)
```

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "foce", covariance = TRUE,
                verbose = FALSE)
```

## Simulating from the fitted model

The typical VPC workflow passes `fit` to `ferx_simulate()` so that simulation uses the fitted theta, omega, and sigma rather than the model file's initial values:

```{r}
#| label: simulate
#| cache: true
sim <- ferx_simulate(
  model  = ex$model,
  data   = ex$data,
  n_sim  = 500,
  seed   = 42,
  fit    = fit
)
head(sim)
```

`n_sim = 500` replicates is sufficient for a standard VPC.
Set `seed` for reproducibility.

## What the simulation returns

| Column | Meaning |
|---|---|
| `SIM` | Replicate index (1 to `n_sim`) |
| `ID` | Subject ID (same population structure as observed data) |
| `TIME` | Time (matched to observation schedule in `data`) |
| `IPRED` | Individual prediction for this replicate (no residual error) |
| `DV_SIM` | Simulated observation (`IPRED` + residual noise) |

To propagate parameter uncertainty across draws, use
`ferx_simulate_with_uncertainty()`, which adds a `DRAW` column indexing the
population-parameter draw — see @sec-uncertainty.

## Simulating at initial values (without a fit)

Omit `fit` to simulate at the initial estimates in the model file.
Useful for a prior predictive check:

```{r}
#| eval: false
sim_prior <- ferx_simulate(ex$model, ex$data, n_sim = 200, seed = 1)
```

## Propensity-score-matched simulation (RWD / pmVPC)

VPCs built from **real-world data (RWD)** can be biased by *treatment
adaptation*: a subject's observed dosing and sampling design is often a
response to that subject's own disposition. For example, high-clearance
patients may be given longer dosing intervals. A naïve simulation draws etas
independently of the design, so the simulated and observed populations no
longer share the design–disposition coupling, and the VPC misleads.

The `match = TRUE` argument to `ferx_simulate()` addresses this. For each
replicate, the drawn etas are reassigned to subjects by **optimal Mahalanobis
matching** (under the model omega) against the subjects' fitted (posthoc) etas.
Each subject's observed dosing/sampling design is thereby paired with a similar
drawn eta, restoring the design–disposition coupling and producing a
propensity-matched VPC (pmVPC). This requires `data` to be real observed data —
every subject must have observations so its posthoc eta can be computed. When a
`fit` is supplied, the posthoc etas are computed at the fitted parameters;
otherwise at the model file's initial values.

```{r}
#| label: simulate-matched
#| eval: false
sim_pm <- ferx_simulate(
  model = ex$model,
  data  = ex$data,
  n_sim = 500,
  seed  = 42,
  fit   = fit,
  match = TRUE
)
```

The returned columns are identical to the unmatched call (`SIM`, `ID`, `TIME`,
`IPRED`, `DV_SIM`); feed `sim_pm` into the same VPC construction below.

## Building a VPC

### Using the `vpc` package

```{r}
#| label: vpc-pkg
#| eval: false
library(vpc)

vpc(
  obs  = obs |> mutate(DV = as.numeric(DV)),
  sim  = sim,
  obs_cols = list(),
  sim_cols = list(dv = "DV_SIM")
)
```

(The `vpc` package is an optional dependency; install with
`install.packages("vpc")`.)

## Population predictions

`ferx_predict()` computes population predictions (ETA = 0) without residual noise.
Useful for overlaying the typical profile on observed data:

```{r}
#| label: predict
#| cache: true
preds <- ferx_predict(ex$model, ex$data, fit = fit)
head(preds)
```

`ferx_predict()` returns a data frame with `ID`, `TIME`, and `PRED` (population
prediction at ETA = 0). For uncertainty-propagated predictions, see
`ferx_simulate_with_uncertainty()` in @sec-uncertainty.

## `[simulation]` block

For CLI-driven simulation without an external dataset, the `.ferx` model file
can embed a trial design via the `[simulation]` block. This is a DSL feature
rather than an R workflow — see the [Model file reference](https://ferx-nlme.github.io/ferx-core/model-file/index.html)
in the ferx-core book for syntax and usage.
