---
title: "Parameter Transforms"
---

## Overview {#sec-parameter-transforms}

How an ETA is incorporated into the individual parameter determines the **distribution** of that parameter across subjects.
ferx supports three common transforms: log-normal, additive, and logit-normal.

```{r}
#| label: setup
#| message: false
library(ferx)
```

## Log-normal (default)

```
[individual_parameters]
  CL = TVCL * exp(ETA_CL)
```

The individual clearance `CL_i = TVCL × exp(η_i)` where `η_i ~ N(0, ω²)`.

This parameterisation:

- Keeps `CL_i > 0` by construction
- `ω² ≈ (CV%)²` for small-to-moderate variability (the log-normal approximation)
- The geometric mean of CL across subjects equals `TVCL`
- The arithmetic CV% is approximately `√ω × 100`

**When to use**: clearance, volume, rate constants — parameters that are always positive and whose variability is naturally multiplicative.

### CV% formula

For a log-normal ETA with omega variance $\omega^2$:

$$\text{CV\%} \approx \sqrt{\omega^2} \times 100$$

This is the _approximate_ formula.
The exact formula is:

$$\text{CV\%} = \sqrt{e^{\omega^2} - 1} \times 100$$

For $\omega^2 < 0.25$ the two agree to within 1–2 percentage points.

```{r}
#| label: cv-formula
omega_sq <- c(0.04, 0.09, 0.16, 0.25, 0.50)
data.frame(
  omega_sq = omega_sq,
  cv_approx = sqrt(omega_sq) * 100,
  cv_exact  = sqrt(exp(omega_sq) - 1) * 100
)
#>   omega_sq cv_approx cv_exact
#>       0.04      20.0     20.2
#>       0.09      30.0     30.5
#>       0.16      40.0     41.5
#>       0.25      50.0     52.8
#>       0.50      70.7     79.7
```

## Additive ETA

```
[individual_parameters]
  TLAG = TVTLAG + ETA_TLAG
```

The individual lag time `TLAG_i = TVTLAG + η_i` where `η_i ~ N(0, ω²)`.
The omega variance is now in the same units as the parameter (e.g., hours²).

**When to use**:

- Parameters that can be zero or negative by physiology (some effect parameters)
- Lag times constrained near zero — the additive form avoids inflation of CV% at small values
- Log-scale parameters (take `log(TVCL)` as the typical value)

Example from the bundled `warfarin_additive_eta` model:

```{r}
#| label: additive-model
ex <- ferx_example("warfarin_additive_eta")
ferx_model_show(ex$model)
```

```
[individual_parameters]
  CL   = TVCL  * exp(ETA_CL)
  V    = TVV   * exp(ETA_V)
  KA   = TVKA
  TLAG = TVTLAG + ETA_TLAG      # additive — variance in h²
```

```{r}
#| label: fit-additive
#| cache: true
fit_add <- ferx_fit(ex$model, ex$data, method = "focei", covariance = FALSE)
summary(fit_add)
```

## Logit-normal for bounded parameters {#sec-logit-normal}

For parameters bounded to `(0, 1)` — bioavailability, fraction absorbed — use the logit transformation:

```
[individual_parameters]
  F = inv_logit(logit(THETA_F) + ETA_F)
```

`THETA_F` is specified on the `(0, 1)` scale in `[parameters]` with bounds `(0.001, 0.999)`.
ferx maps it to the logit scale internally so the optimiser works on an unbounded parameter space.

The individual bioavailability is:

$$F_i = \text{logit}^{-1}(\text{logit}(\theta_F) + \eta_i) = \frac{e^{\text{logit}(\theta_F) + \eta_i}}{1 + e^{\text{logit}(\theta_F) + \eta_i}}$$

**CV% is not meaningful** for logit-normal parameters because the distribution is asymmetric.
Report the omega variance on the logit scale and the 5th–95th percentile range:

```{r}
#| label: logit-range
theta_f <- 0.80   # typical bioavailability
omega_f <- 0.10   # omega variance on logit scale

logit    <- function(p) log(p / (1 - p))
inv_logit <- function(x) 1 / (1 + exp(-x))

logit_typical <- logit(theta_f)
lo <- inv_logit(logit_typical - 1.96 * sqrt(omega_f))
hi <- inv_logit(logit_typical + 1.96 * sqrt(omega_f))
cat(sprintf("F: typical = %.2f, 95%% range = [%.2f, %.2f]\n", theta_f, lo, hi))
#> F: typical = 0.80, 95% range = [0.64, 0.90]
```

Example using the bundled `warfarin_logit_f` model:

```{r}
#| label: fit-logit
#| cache: true
ex_lf <- ferx_example("warfarin_logit_f")
fit_lf <- ferx_fit(ex_lf$model, ex_lf$data, method = "focei", covariance = FALSE)
summary(fit_lf)
```

## Choosing the right transform

| Situation | Parameterisation |
|---|---|
| Always positive, multiplicative variability | `TVCL * exp(ETA_CL)` |
| Near-zero or could go negative | `TVTLAG + ETA_TLAG` |
| Bounded `(0, 1)` | `inv_logit(logit(THETA_F) + ETA_F)` |
| Log-scale typical value | `exp(log_TVCL + ETA_CL)` (equivalent to log-normal) |

## Omega display in `summary()`

The `summary()` output shows omega on the variance scale.
To convert manually:

```{r}
#| label: omega-to-cv
fit <- ferx_fit(ferx_example("warfarin")$model,
                ferx_example("warfarin")$data)
omega_diag <- diag(fit$omega)
cv_approx  <- sqrt(omega_diag) * 100
names(cv_approx) <- names(omega_diag)
round(cv_approx, 1)
#>  ETA_CL  ETA_V ETA_KA
#>    20.6   12.7   57.2
```
