---
title: "ODE Models"
---

## Overview {#sec-ode-models}

When no analytical solution exists — saturable elimination, receptor occupancy, PK/PD links — use the `[odes]` section.
ferx integrates the ODE system numerically using an adaptive Runge-Kutta-45 solver (Dormand-Prince) for each subject and each EBE evaluation.

## The Michaelis-Menten example

The `mm_oral` bundled example is a one-compartment oral model with Michaelis-Menten (saturable) elimination.
It cannot be solved analytically.

```{r}
#| label: setup
#| message: false
library(ferx)
library(ggplot2)

ex <- ferx_example("mm_oral")
ferx_model_show(ex$model)
```

```
# 1-compartment oral PK with Michaelis-Menten elimination

[parameters]
  theta TVVMAX(4.0, 0.1, 50.0)
  theta TVKM(6.0, 0.1, 100.0)
  theta TVV(12.0, 1.0, 200.0)
  theta TVKA(1.5, 0.05, 20.0)

  omega ETA_VMAX ~ 0.15
  omega ETA_V    ~ 0.10

  sigma PROP_ERR ~ 0.02

[individual_parameters]
  VMAX = TVVMAX * exp(ETA_VMAX)
  KM   = TVKM
  V    = TVV * exp(ETA_V)
  KA   = TVKA

[structural_model]
  ode(obs_cmt=central, states=[depot, central])

[odes]
  d/dt(depot)   = -KA * depot
  d/dt(central) = KA * depot / V - VMAX * central / (KM + central)

[error_model]
  DV ~ proportional(PROP_ERR)
```

### Key DSL syntax for ODEs

**`[structural_model]`** declares the ODE system:

```
ode(obs_cmt=central, states=[depot, central])
```

- `obs_cmt` — the compartment name whose amount is divided by `V` to give the `IPRED` concentration
- `states` — list of all compartment names in the order used in dose records (`CMT` column)

**`[odes]`** contains one `d/dt(name) = ...` line per state:

```
d/dt(central) = KA * depot / V - VMAX * central / (KM + central)
```

Note that `central` here is the **amount** (not concentration).
Concentrations are computed as `amount / V` and should appear in the ODE where needed.

## Fitting an ODE model

```{r}
#| label: fit-ode
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
summary(fit)
```

## Dosing into compartments

The `CMT` column in the dataset routes doses to compartments.
With `states=[depot, central]`:

- `CMT = 1` → dose into `depot`
- `CMT = 2` → dose into `central`

Infusions use `RATE > 0` in the dose row; the solver handles the rate as a continuous input over the infusion duration.

## Analytical vs ODE: when to choose each

| Model feature | Analytical | ODE |
|---|---|---|
| Linear compartments, first-order absorption | Preferred | Possible |
| Saturable (MM) elimination | Not possible | Required |
| Effect compartment / PD link | Not possible | Required |
| Two-cpt with zero-order absorption | Not directly | Possible |
| Transit compartments | Not directly | Possible |

Use analytical solutions when available — they are exact, faster, and have no step-size error.
Reserve ODEs for nonlinear or structurally complex models.

## Performance

Each function evaluation requires integrating the ODE for every subject.
A few tips:

1. Use `method = c("gn", "focei")` as a chain — Gauss-Newton is fast for a warm start, then FOCEI provides the rigorous final estimate
2. Pass `threads = parallel::detectCores(logical = FALSE)` to parallelise subject evaluations
3. Cache the fit with `cache: true` in the Quarto chunk so the book does not re-run it every render

## Tolerance and step size

The Dormand-Prince solver uses adaptive step size with default tolerances:

- Relative tolerance: `1e-6`
- Absolute tolerance: `1e-8`

These are set in the backend and cannot currently be overridden from R.
For stiff systems (e.g. fast-equilibrating effect compartments), the solver will automatically take smaller steps, which increases runtime.
