---
title: "ODE Models"
---

## Overview {#sec-ode-models}

When no analytical solution exists — saturable elimination, receptor occupancy, PK/PD links — use the `[odes]` section.
ferx integrates the ODE system numerically using an adaptive Runge-Kutta-45 solver (Dormand-Prince) for each subject and each EBE evaluation.

## The Michaelis-Menten example

The `mm_oral` bundled example is a one-compartment oral model with Michaelis-Menten (saturable) elimination.
It cannot be solved analytically.

```{r}
#| label: setup
#| message: false
library(ferx)
library(ggplot2)

ex <- ferx_example("mm_oral")
ferx_model_show(ex$model)
```

```
# 1-compartment oral PK with Michaelis-Menten elimination

[parameters]
  theta TVVMAX(4.0, 0.1, 50.0)
  theta TVKM(6.0, 0.1, 100.0)
  theta TVV(12.0, 1.0, 200.0)
  theta TVKA(1.5, 0.05, 20.0)

  omega ETA_VMAX ~ 0.15
  omega ETA_V    ~ 0.10

  sigma PROP_ERR ~ 0.02

[individual_parameters]
  VMAX = TVVMAX * exp(ETA_VMAX)
  KM   = TVKM
  V    = TVV * exp(ETA_V)
  KA   = TVKA

[structural_model]
  ode(obs_cmt=central, states=[depot, central])

[odes]
  d/dt(depot)   = -KA * depot
  d/dt(central) = KA * depot / V - VMAX * central / (KM + central)

[error_model]
  DV ~ proportional(PROP_ERR)
```

### Key DSL syntax for ODEs

**`[structural_model]`** declares the ODE system:

```
ode(obs_cmt=central, states=[depot, central])
```

- `obs_cmt` — the compartment name whose amount is divided by `V` to give the `IPRED` concentration
- `states` — list of all compartment names in the order used in dose records (`CMT` column)

**`[odes]`** contains one `d/dt(name) = ...` line per state:

```
d/dt(central) = KA * depot / V - VMAX * central / (KM + central)
```

Note that `central` here is the **amount** (not concentration).
Concentrations are computed as `amount / V` and should appear in the ODE where needed.

## Fitting an ODE model

```{r}
#| label: fit-ode
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
summary(fit)
```

## Dosing into compartments

The `CMT` column in the dataset routes doses to compartments.
With `states=[depot, central]`:

- `CMT = 1` → dose into `depot`
- `CMT = 2` → dose into `central`

Infusions use `RATE > 0` in the dose row; the solver handles the rate as a continuous input over the infusion duration.

## PK/PD effect compartment {#sec-effect-cmt}

Plasma concentration and pharmacodynamic response often do not peak at the same time.
An **effect compartment** (or _link model_) captures this equilibration lag: drug transfers from the central compartment to the effect site with a first-order rate constant **ke0**, and the response is driven by the effect-compartment concentration Ce rather than the plasma concentration.

Because the effect compartment is assumed to be pharmacokinetically negligible (it does not measurably drain the central compartment), its ODE is:

```
d/dt(effect) = ke0 * (central / V1) - ke0 * effect
```

At equilibrium `d/dt(effect) = 0` implies `effect = central / V1 = Cp`, so Ce equals plasma concentration at steady-state — only the rate of approach differs.

## Multi-dose event handling

The dataset drives dosing via standard NONMEM-style columns.
Understanding how the solver interprets event records is essential for multi-dose studies.

**Bolus doses (`EVID = 1`, `RATE = 0`)**

The AMT is added instantaneously to the compartment specified by `CMT` at the event time.
The solver resets the state to `state[CMT] + AMT` and continues integration.

**IV infusions (`EVID = 1`, `RATE > 0`)**

A positive `RATE` column turns the dose record into a zero-order input lasting `AMT / RATE` hours.
The solver treats it as a continuous forcing function added to the ODE right-hand side for the infusion duration:

```
d/dt(central) = RATE_input(t) - (CL/V1) * central
```

`RATE_input(t)` equals `RATE` during the infusion window and 0 otherwise.
No special model syntax is needed — the event record is sufficient.

**Steady-state flag (`SS = 1`)**

Setting `SS = 1` in a dose record instructs the solver to compute the steady-state initial conditions analytically (for linear models) or iteratively (for ODE models) before advancing time.
This avoids simulating hundreds of wash-in doses explicitly and is especially useful for rich Phase I designs with a loading dose followed by QD maintenance.

**`CMT` column and `states` order**

The integer in the `CMT` column maps to the `states=[...]` list by position (1-indexed):

| CMT | Compartment |
|-----|-------------|
| 1   | `central`   |
| 2   | `peripheral`|
| 3   | `effect`    |

Observation records (`EVID = 0`) use `CMT` to indicate which compartment the measured DV belongs to, which must match `obs_cmt`.

## Common ODE pitfalls

| Pitfall | What goes wrong | Fix |
|---------|-----------------|-----|
| Amounts vs concentrations in ODEs | Writing `d/dt(central) = -CL * central` treats `central` as a concentration, not an amount — CL is then dimensionless | Use `d/dt(central) = -(CL/V) * central`; divide by V wherever concentration is needed |
| `obs_cmt` not in `states` | Parser error at validation time | Add the compartment name to the `states=[...]` list |
| Stiff systems | Solver takes thousands of tiny steps, fit becomes very slow or fails to converge | Check for rate-constant mismatches (e.g. ke0 >> CL/V); consider re-parameterising on log scale or tightening bounds |
| Fast absorption + slow elimination | The stiff ratio between KA and CL/V causes the solver to over-refine early time points | Either use an analytical solution for the PK part, or supply tighter initial bounds on KA |
| Off-by-one in CMT | Dose enters the wrong compartment | Print `states` order explicitly and cross-check with your dataset's `CMT` values before fitting |

A quick stiffness check: if the ratio of the largest to smallest eigenvalue of the Jacobian exceeds ~1000, the system is stiff.
In pharmacokinetics this often surfaces when ke0 (effect-site equilibration) is much faster than the slowest elimination rate constant.

## Analytical vs ODE: when to choose each

| Model feature | Analytical | ODE |
|---|---|---|
| Linear compartments, first-order absorption | Preferred | Possible |
| Saturable (MM) elimination | Not possible | Required |
| Effect compartment / PD link | Not possible | Required |
| Two-cpt with zero-order absorption | Not directly | Possible |
| Transit compartments | Not directly | Possible |

Use analytical solutions when available — they are exact, faster, and have no step-size error.
Reserve ODEs for nonlinear or structurally complex models.

## Performance

Each function evaluation requires integrating the ODE for every subject.
A few tips:

1. Use `method = c("gn", "focei")` as a chain — Gauss-Newton is fast for a warm start, then FOCEI provides the rigorous final estimate
2. Pass `threads = parallel::detectCores(logical = FALSE)` to parallelise subject evaluations
3. Cache the fit with `cache: true` in the Quarto chunk so the book does not re-run it every render

## Tolerance and step size

The Dormand-Prince solver uses adaptive step size with default tolerances:

- Relative tolerance: `1e-6`
- Absolute tolerance: `1e-8`

These are set in the backend and cannot currently be overridden from R.
For stiff systems (e.g. fast-equilibrating effect compartments), the solver will automatically take smaller steps, which increases runtime.
