---
title: "Uncertainty & SIR"
---

## Overview {#sec-uncertainty}

The covariance step gives **asymptotic** standard errors — exact only for a
locally quadratic log-likelihood and infinite data. When that assumption is
shaky (small sample, parameters near a bound, highly nonlinear model), ferx
offers two non-asymptotic tools:

| Tool | What it gives you |
|---|---|
| `ferx_sir()` | Sampling-importance-resampling: a non-parametric set of joint parameter draws |
| `ferx_simulate_with_uncertainty()` | Population-parameter uncertainty propagated through `ferx_simulate()` |
| `method = c("...", "imp")` | A final importance-sampling stage that yields a marginal log-likelihood and effective sample size |

```{r}
#| label: setup
#| message: false
library(ferx)
ex  <- ferx_example("warfarin")
```

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "foce", covariance = TRUE,
                verbose = FALSE)
```

## Sampling Importance Resampling

`ferx_sir()` takes a converged fit and returns an updated `ferx_fit` object
that carries a non-parametric joint sample of the population parameters in
`$sir_resamples`. It draws candidate vectors from a multivariate proposal
centred at the point estimate (built from the asymptotic covariance), weights
each by the likelihood ratio, and resamples to give a usable empirical
distribution.

```{r}
#| label: sir
#| eval: false
sir_res <- ferx_sir(
  fit,
  sir_samples       = 1000,    # candidate draws
  sir_resamples     = 250,     # kept after resampling
  sir_seed          = 1,
  sir_keep_samples  = TRUE     # keep the raw draws for diagnostics
)
```

When the asymptotic ellipsoid and the true likelihood diverge, the resampled
distribution becomes skewed or has heavier tails than the SE-based 95% CI —
that disagreement is the signal SIR is designed to surface.

### SIR diagnostics

The resampling step itself has diagnostics: the effective sample size (ESS) of
the importance weights, and how many of the `sir_samples` were actually kept.
A low ESS means the proposal was a poor match for the true likelihood — widen
the proposal or run more samples.

For a deeper treatment of SIR practice, see
[Example: Uncertainty (SIR)](https://ferx-nlme.github.io/ferx-site/learn/uncertainty-sir.html)
on the ferx site.

## Propagating uncertainty into simulations

`ferx_simulate()` simulates with the **point estimate** held fixed. For dose-
response predictions, label predictions, or "what does my model say at a new
weight band" questions, the right object to propagate is the *joint*
uncertainty of the population parameters — not just the asymptotic SE on each
theta one at a time.

`ferx_simulate_with_uncertainty()` does exactly that:

```{r}
#| label: sim-uncertainty
#| eval: false
sim_u <- ferx_simulate_with_uncertainty(
  model               = ex$model,
  data                = ex$data,
  fit                 = fit,
  n_uncertainty_draws = 100,         # draws of (theta, omega, sigma)
  n_sim_per_draw      = 1,           # ETA/EPS replicates per draw
  method              = "asymptotic",  # or "sir"
  seed                = 1
)
```

The returned data frame carries the same columns as `ferx_simulate()` plus a
`DRAW` index that identifies which population-parameter draw produced each
row. Summarise across `DRAW` to get prediction intervals that include
estimation uncertainty:

```{r}
#| eval: false
library(dplyr)

sim_u |>
  group_by(ID, TIME) |>
  summarise(lo  = quantile(DV_SIM, 0.05),
            mid = median(DV_SIM),
            hi  = quantile(DV_SIM, 0.95),
            .groups = "drop")
```

### Choosing `method`

| `method` | Where the parameter draws come from |
|---|---|
| `"asymptotic"` (default) | Multivariate normal centred at the point estimate, covariance from the Hessian. Requires the covariance step to have succeeded. |
| `"sir"` | Resamples from a SIR run. Honest about non-Gaussian likelihood shape but requires `sir = TRUE` (or a separate `ferx_sir()` call) so the resamples exist. |

Use `"asymptotic"` for routine reporting once the fit is clearly identified.
Use `"sir"` when SIR has already flagged that the asymptotic covariance is a
poor approximation.

## The `imp` final stage

For a converged FOCE/FOCEI fit, the asymptotic OFV is an approximation of
$-2\log L$ — adequate for likelihood-ratio tests on nested models, but biased
in absolute terms. Appending `"imp"` to the method chain runs an importance
sampling pass at the converged estimate and reports the marginal
log-likelihood directly:

```{r}
#| eval: false
fit_imp <- ferx_fit(
  ex$model, ex$data,
  method = c("focei", "imp"),
  settings = list(
    is_samples     = 1000,    # importance-sampling draws per subject
    is_proposal_df = 5        # heavy-tailed proposal (t with df)
  )
)
```

The IMP stage adds the marginal log-likelihood and an effective-sample-size
diagnostic to the fit object (`fit_imp$importance_sampling`). Low ESS means
the FOCE-conditional proposal does not cover the integrand well — try
more samples, fewer degrees of freedom (heavier tails), or check whether the
Laplace approximation is breaking down (highly nonlinear models, sparse data).

## When to reach for which tool

| Situation | Reach for |
|---|---|
| Reporting parameter SE/CI on a well-identified model | `covariance = TRUE` (default) |
| Boundary estimates, weak identifiability, small N | `ferx_sir()` |
| Forward predictions with full uncertainty (dosing, exposure-response) | `ferx_simulate_with_uncertainty()` |
| Honest marginal log-likelihood for model selection | `method = c("focei", "imp")` |
