---
title: "Uncertainty & SIR"
---

## Overview {#sec-uncertainty}

The covariance step gives **asymptotic** standard errors — exact only for a
locally quadratic log-likelihood and infinite data. When that assumption is
shaky (small sample, parameters near a bound, highly nonlinear model), ferx
offers two non-asymptotic tools:

| Tool | What it gives you |
|---|---|
| `ferx_sir()` | Sampling-importance-resampling: a non-parametric set of joint parameter draws |
| `ferx_simulate_with_uncertainty()` | Population-parameter uncertainty propagated through `ferx_simulate()` |
| `method = c("...", "imp")` | A final importance-sampling stage that yields a marginal log-likelihood and effective sample size |

```{r}
#| label: setup
#| message: false
library(ferx)
ex  <- ferx_example("warfarin")
```

```{r}
#| label: fit
#| cache: true
fit <- ferx_fit(ex$model, ex$data, method = "foce", covariance = TRUE,
                verbose = FALSE)
```

## Strengthening the asymptotic covariance step

Before reaching for SIR, it is usually worth trying to fix or strengthen the
asymptotic covariance step itself. ferx exposes three controls for this, and
they are the first line of defense when `fit$covariance_status` is not
`"computed"`:

| Control | Where | What it does |
|---|---|---|
| `fd_hessian_step` | dedicated `ferx_fit()` argument (default `0.01`) | Relative step for the finite-difference Hessian. Increase (e.g. `0.1`) when the fit warns about an ill-conditioned Hessian; decrease (e.g. `1e-3`) on smooth surfaces where FD noise dominates. |
| `covariance_method` | `settings = list(...)` key | Covariance estimator: `"r"` (inverse Hessian, default), `"s"` (score cross-product), or `"rsr"` (Huber-White sandwich SEs, robust to model misspecification). `"s"` and `"rsr"` are rejected under non-interaction FOCE. |
| `covariance_fallback` | `settings = list(...)` key | `"none"` (default) or `"sir"`. With `"sir"`, a non-positive-definite finite-difference Hessian triggers an automatic SIR run instead of a failed covariance step, and `fit$covariance_status` becomes `"sir_fallback"`. |

```{r}
#| label: cov-robust
#| eval: false
fit_cov <- ferx_fit(
  ex$model, ex$data,
  method          = "focei",
  covariance      = TRUE,
  fd_hessian_step = 0.1,                 # rescue an ill-conditioned FD Hessian
  settings = list(
    covariance_method   = "rsr",         # Huber-White sandwich SEs
    covariance_fallback = "sir"          # auto-SIR if the Hessian is non-PD
  )
)
```

If the covariance step still cannot be trusted after these adjustments — or if
`covariance_status` reports `"sir_fallback"` — move on to an explicit SIR run
below for a fully non-asymptotic picture.

## Sampling Importance Resampling

`ferx_sir()` takes a converged fit and returns an updated `ferx_fit` object
that carries a non-parametric joint sample of the population parameters in
`$sir_resamples`. It draws candidate vectors from a multivariate proposal
centred at the point estimate (built from the asymptotic covariance), weights
each by the likelihood ratio, and resamples to give a usable empirical
distribution.

```{r}
#| label: sir
#| eval: false
sir_res <- ferx_sir(
  fit,
  sir_samples       = 1000,    # candidate draws
  sir_resamples     = 250,     # kept after resampling
  sir_seed          = 1,
  sir_keep_samples  = TRUE     # keep the raw draws for diagnostics
)
```

When the asymptotic ellipsoid and the true likelihood diverge, the resampled
distribution becomes skewed or has heavier tails than the SE-based 95% CI —
that disagreement is the signal SIR is designed to surface.

### Models with fixed parameters

When the model contains `FIX`-ed parameters, `ferx_sir()` excludes them from
the proposal distribution and holds them constant across all resamples.
Only freely estimated parameters are varied. The resampled distribution and
all downstream uncertainty intervals reflect only the free parameters.

### SIR diagnostics

The resampling step itself has diagnostics: the effective sample size (ESS) of
the importance weights, and how many of the `sir_samples` were actually kept.
A low ESS means the proposal was a poor match for the true likelihood — widen
the proposal or run more samples.

For a deeper treatment of SIR practice, see
[Example: Uncertainty (SIR)](https://ferx-nlme.github.io/learn/uncertainty-sir.html)
on the ferx site.

## Propagating uncertainty into simulations

`ferx_simulate()` simulates with the **point estimate** held fixed. For dose-
response predictions, label predictions, or "what does my model say at a new
weight band" questions, the right object to propagate is the *joint*
uncertainty of the population parameters — not just the asymptotic SE on each
theta one at a time.

`ferx_simulate_with_uncertainty()` does exactly that:

```{r}
#| label: sim-uncertainty
#| eval: false
sim_u <- ferx_simulate_with_uncertainty(
  model               = ex$model,
  data                = ex$data,
  fit                 = fit,
  n_uncertainty_draws = 100,         # draws of (theta, omega, sigma)
  n_sim_per_draw      = 1,           # ETA/EPS replicates per draw
  method              = "asymptotic",  # or "sir"
  seed                = 1
)
```

The returned data frame carries the same columns as `ferx_simulate()` plus a
`DRAW` index that identifies which population-parameter draw produced each
row. Summarise across `DRAW` to get prediction intervals that include
estimation uncertainty:

```{r}
#| eval: false
library(dplyr)

sim_u |>
  group_by(ID, TIME) |>
  summarise(lo  = quantile(DV_SIM, 0.05),
            mid = median(DV_SIM),
            hi  = quantile(DV_SIM, 0.95),
            .groups = "drop")
```

### Choosing `method`

| `method` | Where the parameter draws come from |
|---|---|
| `"asymptotic"` (default) | Multivariate normal centred at the point estimate, covariance from the Hessian. Requires the covariance step to have succeeded. |
| `"sir"` | Resamples from a SIR run. Honest about non-Gaussian likelihood shape but requires `sir = TRUE` (or a separate `ferx_sir()` call) so the resamples exist. |

Use `"asymptotic"` for routine reporting once the fit is clearly identified.
Use `"sir"` when SIR has already flagged that the asymptotic covariance is a
poor approximation.

## The `imp` final stage

For a converged FOCE/FOCEI fit, the asymptotic OFV is an approximation of
$-2\log L$ — adequate for likelihood-ratio tests on nested models, but biased
in absolute terms. Appending `"imp"` to the method chain runs an importance
sampling pass at the converged estimate and reports the marginal
log-likelihood directly:

```{r}
#| eval: false
fit_imp <- ferx_fit(
  ex$model, ex$data,
  method = c("focei", "imp"),
  settings = list(
    is_samples     = 1000,    # importance-sampling draws per subject
    is_proposal_df = 5        # heavy-tailed proposal (t with df)
  )
)
```

The IMP stage adds the marginal log-likelihood and an effective-sample-size
diagnostic to the fit object (`fit_imp$importance_sampling`). Low ESS means
the FOCE-conditional proposal does not cover the integrand well — try
more samples, fewer degrees of freedom (heavier tails), or check whether the
Laplace approximation is breaking down (highly nonlinear models, sparse data).

`"imp"` can also run **standalone** — `method = "imp"` with no preceding
estimator. In that case it does not optimise anything; it evaluates the
importance-sampling marginal log-likelihood at the model's *initial*
parameters and reports the same `$importance_sampling` diagnostics. This is
useful for scoring a fixed parameter set (e.g. values carried over from a
reference fit) without re-estimating:

```{r}
#| eval: false
fit_imp0 <- ferx_fit(
  ex$model, ex$data,
  method   = "imp",          # standalone: score the initial parameters
  settings = list(is_samples = 1000, is_proposal_df = 5)
)
```

## When to reach for which tool

| Situation | Reach for |
|---|---|
| Reporting parameter SE/CI on a well-identified model | `covariance = TRUE` (default) |
| Boundary estimates, weak identifiability, small N | `ferx_sir()` |
| Forward predictions with full uncertainty (dosing, exposure-response) | `ferx_simulate_with_uncertainty()` |
| Honest marginal log-likelihood for model selection | `method = c("focei", "imp")` |
| Marginal log-likelihood at a fixed/initial parameter set (no re-fit) | `method = "imp"` (standalone) |
| Asymptotic covariance failed or is ill-conditioned | `fd_hessian_step`, `covariance_method`, `covariance_fallback` (before SIR) |
