---
title: "BLOQ & M3 Method"
---

## Overview {#sec-bloq-m3}

Concentrations below the lower limit of quantification (LLOQ) are a routine challenge in PK/PD analysis.
The simplest approach — discarding or replacing BLOQ values — introduces bias.
Beal's M3 method handles BLOQ observations correctly by integrating the likelihood over the censored region.

## The M3 likelihood

For an uncensored observation:

$$\ell_i = \log f(y_i \mid \theta, \eta_i)$$

For a censored observation with LLOQ $c$:

$$\ell_i = \log P(Y_i < c \mid \theta, \eta_i) = \log \Phi\!\left(\frac{c - \hat{y}_i}{\sigma_i}\right)$$

where $\Phi$ is the standard normal CDF and $\hat{y}_i$ is the individual prediction at the LLOQ time.

The M3 method treats each BLOQ observation as contributing its probability mass below the LLOQ rather than a point observation.
This eliminates the downward bias in IPRED introduced by dropping or imputing BLOQ values.

## Dataset requirements

Add a `CENS` column to your dataset:

| EVID | DV | CENS | Interpretation |
|---|---|---|---|
| 0 | observed value | 0 | Normal observation |
| 0 | LLOQ value | 1 | Censored — DV carries the LLOQ |
| 1 | AMT | 0 | Dose record |

The `DV` column for censored rows should contain the LLOQ value (e.g., `2.0 mg/L`), not `NA` or `0`.

## Model specification

Enable M3 in `[fit_options]`:

```
[fit_options]
  method       = focei
  covariance   = true
  bloq_method  = m3
```

Or override at run time:

```{r}
#| eval: false
fit_m3 <- ferx_fit(ex$model, ex$data, bloq_method = "m3")
```

Setting `bloq_method = "m3"` at the call level overrides whatever is in the model file.
Use `bloq_method = "drop"` to explicitly drop BLOQ rows (not recommended in general).

## Fitting with M3

```{r}
#| label: setup
#| message: false
library(ferx)
library(ggplot2)
library(dplyr)

ex     <- ferx_example("warfarin_bloq")
obs    <- read.csv(ex$data)
ferx_model_show(ex$model)
```

```{r}
#| label: fit-m3
#| cache: true
fit_m3 <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
summary(fit_m3)
```

`summary(fit_m3)` includes the model structure, covariance condition number,
and any warnings — use `ferx_warnings(fit_m3)` to inspect them grouped by severity.

## Comparing M3 vs naive BLOQ removal

To quantify the bias from discarding BLOQ values, fit the same dataset both ways:

```{r}
#| label: fit-drop
#| cache: true
fit_drop <- ferx_fit(ex$model, ex$data, method = "focei",
                     bloq_method = "drop", covariance = FALSE)
```

```{r}
#| label: compare
cat(sprintf("M3   OFV: %.1f\n", fit_m3$ofv))
cat(sprintf("Drop OFV: %.1f\n", fit_drop$ofv))

# Compare TVCL estimates
cat(sprintf("TVCL M3:   %.3f\n", fit_m3$theta["TVCL"]))
cat(sprintf("TVCL drop: %.3f\n", fit_drop$theta["TVCL"]))
```

Typical finding: dropping BLOQ underestimates CL because the model predicts lower concentrations than truly exist at BLOQ times.

## Diagnostics with censored data

Residuals (CWRES, IWRES) are only defined for uncensored observations.
Filter before plotting:

```{r}
#| label: gof-m3
sdtab <- fit_m3$sdtab |> filter(CENS == 0 | is.na(CENS))

ggplot(sdtab, aes(x = PRED, y = CWRES)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0,      colour = "#1F6B7A", linewidth = 1) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "#C8893E") +
  labs(x = "PRED", y = "CWRES", title = "CWRES (uncensored obs only)") +
  theme_minimal()
```

## VPC with censored data

For VPCs with BLOQ data, censor the simulated data at the same LLOQ:

```{r}
#| label: vpc-bloq
#| cache: true
lloq <- 2.0   # mg/L

sim <- ferx_simulate(ex$model, ex$data, n_sim = 500, seed = 42, fit = fit_m3)

# Apply same censoring to simulations
sim <- sim |> mutate(DV_SIM_CENS = ifelse(DV_SIM < lloq, lloq, DV_SIM),
                     CENS_SIM    = ifelse(DV_SIM < lloq, 1L, 0L))

# Fraction BLOQ by time bin
pct_bloq_obs <- obs |>
  filter(EVID == 0) |>
  group_by(TIME) |>
  summarise(pct = mean(CENS == 1) * 100, .groups = "drop")

pct_bloq_sim <- sim |>
  group_by(SIM, TIME) |>
  summarise(pct = mean(CENS_SIM == 1) * 100, .groups = "drop") |>
  group_by(TIME) |>
  summarise(lo = quantile(pct, 0.05), mid = median(pct),
            hi = quantile(pct, 0.95), .groups = "drop")

ggplot() +
  geom_ribbon(data = pct_bloq_sim, aes(x = TIME, ymin = lo, ymax = hi),
              fill = "#C9582A", alpha = 0.25) +
  geom_line(data = pct_bloq_sim, aes(x = TIME, y = mid),
            colour = "#C9582A", linewidth = 1) +
  geom_point(data = pct_bloq_obs, aes(x = TIME, y = pct)) +
  labs(x = "Time (h)", y = "% BLOQ",
       title = "% BLOQ over time: observed vs simulated") +
  theme_minimal()
```
