# Model development and selection {#sec-model-selection}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

```{r}
#| label: stage-banner
#| echo: false
#| output: asis
book_stage_banner("EVALUATE")
```

## Where you are

The base model is fitted (@sec-first-fit) and evaluated (@sec-diagnostics, @sec-vpc). The diagnostics hinted at covariate relationships (`ETA_CL` correlated with `CRCL` and `WT`) and flagged autocorrelated residuals. Model development means fitting alternatives and choosing between them. You can do that by hand: write a candidate, fit it and compare. Or let ferx's search tools generate, fit, gate and rank candidates from a search space. This chapter shows both.

## The data

```{r}
#| label: examples
base <- ferx_example("two_cpt_oral_base")
cov <- ferx_example("two_cpt_oral_cov")
```

## Comparing fitted models

`two_cpt_oral_cov` is a prespecified covariate model for the same data. Body weight scales `CL` and `V1` through a shared exponent `THETA_WT`, and creatinine clearance scales `CL`:

```{r}
#| label: cov-model
invisible(ferx_model_get_section(cov$model, "individual_parameters"))
```

```{r}
#| label: fits
fit_base <- ferx_fit(base$model, base$data, verbose = FALSE)
fit_cov <- ferx_fit(cov$model, cov$data, verbose = FALSE)
```

### Likelihood ratio test and information criteria

Both are FOCEI fits to the same data, and the base model is nested in the covariate model (fix both exponents at 0). The difference in OFV is therefore a likelihood ratio statistic, with degrees of freedom equal to the number of added parameters:

```{r}
#| label: lrt
d_ofv <- fit_base$ofv - fit_cov$ofv
df <- fit_cov$n_parameters - fit_base$n_parameters
c(d_ofv = d_ofv, df = df, p_value = pchisq(d_ofv, df, lower.tail = FALSE))
```

For non-nested models, use information criteria. `fit$aic` and `fit$bic` are the classical AIC and BIC. `ferx_bic()` computes four BIC variants that differ in how parameters are penalised:

| `type` | Penalty added to the OFV |
|---|---|
| `"mixed"` (default) | random-effect parameters × log(subjects) + fixed-effect parameters × log(observations) |
| `"fixed"` | all parameters × log(observations); equals `fit$bic` |
| `"iiv"` | variance parameters × log(subjects), for comparing variability structures |
| `"random"` | all parameters × log(subjects) |

```{r}
#| label: tbl-criteria
#| tbl-cap: "Base and covariate model compared on OFV and information criteria (lower is better)."
criteria <- function(f) {
  c(ofv = f$ofv, parameters = f$n_parameters, aic = f$aic, bic = f$bic,
    bic_mixed = ferx_bic(f, "mixed"), bic_iiv = ferx_bic(f, type = "iiv"),
    bic_random = ferx_bic(f, "random"))
}
round(rbind(base = criteria(fit_base), covariate = criteria(fit_cov)), 2)
```

Here the covariate model is better on every criterion.

### Is the fit eligible at all?

A converged fit can still be unfit for comparison: it may never have left its initial estimates, have a parameter at a bound, or have an ill-conditioned covariance matrix. `check_strictness()` applies such gates and returns the reasons for failure:

```{r}
#| label: strictness
check_strictness(fit_cov, require_covariance = TRUE)
```

The fit from NCA-based starting values in @sec-first-fit converged to a much worse OFV. The gates reject it:

```{r}
#| label: strictness-fail
fit_nca <- ferx_fit(base$model, base$data, inits_from_nca = TRUE, verbose = FALSE)
check_strictness(fit_nca)
```

The gates and their defaults: `require_converged = TRUE`, `require_covariance = FALSE`, `max_condition_number = 1000`, `max_correlation = 0.95`, `reject_on_boundary = TRUE` and `reject_init_stall = TRUE`. Set a threshold to `NULL`, or a flag to `FALSE`, to disable that gate. `reject_init_stall` assumes a cold start from the model file's initial values, so turn it off when candidates start from a previous model's estimates.

## Search configurations

::: {.callout-warning}
## Maturity: alpha
The ferx-core pages for the search tools carry the maturity label *alpha*. See [feature maturity](https://ferx-nlme.github.io/ferx-core/maturity.html).
:::

The search tools read a **`.ferxsearch`** configuration file (TOML) that names the base model and data and states the search. Four bundled examples carry one as `$search`:

| Example | Search file for |
|---|---|
| `two_cpt_oral_base` | covariate search (`[covsearch]`), plus an `[allometry]` section (@sec-covariates) |
| `two_cpt_oral_cov` | a configuration with a BIC ranking, used here to show loading |
| `warfarin` | structural search (`[modelsearch]`) |
| `one_cpt_transit` | residual error search (`[ruvsearch]`) |

`ferx_search_config()` loads a file with the engine's own strict loader, so an invalid space or a misspelt section fails before anything is fitted:

```{r}
#| label: config
cfg <- ferx_search_config(base$search)
cfg
```

The file's sections, as shown in the printout: `base` and `data` (paths relative to the file); `[space]` with the search space as `mfl`; `[rank]` with the ranking criterion; `[strictness]` with the gates from above; `[run]` with `threads`, `retries` and `resume`; and a section per tool. The object returned holds the same information as a list:

```{r}
#| label: config-rank
cfg_cov <- ferx_search_config(ferx_example("two_cpt_oral_cov")$search)
cfg_cov$rank$type
cfg$tools
```

### The search space

The space is written in the Model Feature Language (MFL). `COVARIATE?(@IIV, @CONTINUOUS, [pow, lin])` means: every parameter with a random effect, crossed with every continuous covariate, as a power or linear effect, each optional (`?`). `ferx_search_space()` parses MFL text and, given a model and data, expands the symbols into the concrete candidates:

```{r}
#| label: space
ferx_search_space("COVARIATE?(@IIV, @CONTINUOUS, [pow, lin])")
space <- ferx_search_space(cfg$mfl, model = base$model, data = base$data)
nrow(attr(space, "covariate_effects"))
head(attr(space, "covariate_effects"))
```

`ferx_search_coverage()` reports, without fitting, whether the engine can build candidates for each feature of a space:

```{r}
#| label: coverage
ferx_search_coverage(space)
ferx_search_coverage("ABSORPTION(SEQ-ZO-FO)")
```

Sequential zero- then first-order absorption is not generated by the search. @sec-absorption shows the bundled model for it, which you can use as a base model instead.

## Covariate search

`ferx_covsearch()` runs a stepwise covariate search. In each forward step it fits every remaining candidate effect and adds the one with the largest significant OFV drop at `p_forward`. The backward phase then removes effects whose removal is not significant at the stricter `p_backward`. The bundled configuration uses `p_forward = 0.01` and `p_backward = 0.001`:

```{r}
#| label: covsearch
cs_dir <- book_tempdir("covsearch")
cs <- ferx_covsearch(config = base$search, directory = cs_dir, progress = FALSE)
cs
```

The forward phase added `CL ~ CRCL` (linear) at p = `r signif(cs$steps$p_value[cs$steps$selected & cs$steps$phase == "forward"][1], 2)`, below `p_forward`. That p-value is above `p_backward`, so the backward phase removed the effect again. The search returns the base model. Candidates excluded by the strictness gate are kept in the step table with their reason:

```{r}
#| label: covsearch-excluded
cs$steps[!cs$steps$passed, c("step", "parameter", "covariate", "form", "failures")]
```

`print()` shows the selected step of each phase. `summary()` on the result adds every candidate of every step; the same table is `cs$steps`. The run directory holds the per-step journals, `steps.csv` and `final.ferx`. `ferx_search_results()` reads the step table back from disk:

```{r}
#| label: covsearch-results
list.files(cs_dir)
steps <- ferx_search_results(cs_dir, type = "steps")
attr(steps, "tool")
```

The search can also be stated inline. Here it is restricted to `CL`, with a less strict backward level:

```{r}
#| label: covsearch-inline
cs_cl <- ferx_covsearch(model = base$model, data = base$data,
                        search_space = "COVARIATE?(CL, [WT, CRCL], [pow, lin])",
                        p_forward = 0.01, p_backward = 0.01,
                        directory = book_tempdir("covsearch-cl"), progress = FALSE)
cs_cl$included
c(base_ofv = cs_cl$base_ofv, final_ofv = cs_cl$final_ofv)
```

The selected covariate model depends on the search space and the significance levels. A search result is evidence, not a decision. The rest of *The analysis workflow* continues with the prespecified `two_cpt_oral_cov` model, which the comparison above supports.

## Structural search

`ferx_modelsearch()` searches structural features: absorption, elimination, peripheral and transit compartments, and lag time. It ranks the candidates on a criterion, the mixed BIC by default. The warfarin configuration searches a peripheral compartment and a lag time:

```{r}
#| label: modelsearch
ms_dir <- book_tempdir("modelsearch")
warf <- ferx_example("warfarin")
ms <- ferx_modelsearch(config = warf$search, directory = ms_dir, progress = FALSE)
ms
```

The base structure ranks best on the mixed BIC. The lag-time model has a lower OFV, but not by enough to pay for its extra parameters. Every candidate's model text is kept, so the runner-up can be inspected or refitted:

```{r}
#| label: modelsearch-runner-up
runner_up <- file.path(ms_dir, "runner_up.ferx")
writeLines(ms$model_text[["run1"]], runner_up)
invisible(ferx_model_get_section(runner_up, "structural_model"))
```

With `resume = TRUE` and the same `directory`, a search reuses the fits it already journalled:

```{r}
#| label: modelsearch-resume
ms_again <- ferx_modelsearch(config = warf$search, directory = ms_dir, resume = TRUE, progress = FALSE)
ms_again$models[, c("id", "structure", "reused")]
nrow(ferx_search_results(ms_dir, type = "models"))
```

## Residual error search

`ferx_ruvsearch()` tests four families of residual error extensions against the current error model, using the likelihood ratio test at `p_value`:

- `IIV_on_RUV`: a random effect on the residual error magnitude. It needs an estimation method with interaction.
- `power`: a proportional error raised to an estimated power.
- `combined`: an additive component added to the proportional one.
- `time_varying`: a residual error magnitude that differs below a time-after-dose cutoff, set at quantiles given by `groups`.

Up to `max_iter` accepted features are stacked. The bundled configuration searches the analytic transit model `one_cpt_transit`:

```{r}
#| label: ruvsearch
transit <- ferx_example("one_cpt_transit")
rs <- ferx_ruvsearch(config = transit$search, directory = book_tempdir("ruvsearch"), progress = FALSE)
rs
rs$final_features
```

The search accepted a time-varying residual error in the first iteration, and nothing further in the second. The base model of this workflow, which carried a residual autocorrelation warning in @sec-diagnostics, can be searched the same way with the inline arguments (`model`, `data`, `p_value`, `max_iter`, …).

## Variability search

`ferx_iivsearch()` decides the random-effect structure: which parameters carry an eta, and which etas share a covariance block. The space is written with `IIV` and `COVARIANCE` statements, and `?` marks what the search may decide. The bundled `warfarin_block_omega` configuration makes every PK parameter's eta exploratory and offers a block over whichever survive:

```{r}
#| label: iivsearch
bo <- ferx_example("warfarin_block_omega")
iiv <- ferx_iivsearch(config = bo$search, directory = book_tempdir("iivsearch"), progress = FALSE)
iiv
```

The printed output has three steps. The eta-count stage removes etas from the input structure -- `top_down_exhaustive` tries every subset, not one deletion at a time, so the candidates include structures with two or three etas gone. The block stage then offers covariance blocks over the etas that survived. The third step, `compare_to_input`, puts the winner of the block stage back against the input structure, and it is there that the selection is made. Candidates are ranked on the BIC(iiv), which is what `rank = "bic"` means for this tool.

Read the printed output carefully. The input model is a candidate in its own right: the step tables rank it, and in the eta-count stage it ranks first. The strictness gate is separate from that ranking. It rejected `r sum(!iiv$models$passed)` of the fitted models, and one of them reached a better criterion than the winner did:

```{r}
#| label: iivsearch-gate
iiv$models[, c("id", "structure", "criterion", "rank", "passed")]
```

`run8` scores `r round(min(iiv$models$criterion[!iiv$models$passed]), 2)` against the selected `r round(iiv$final_criterion, 2)`, and is out because it did not pass. The input carries no rank in this combined table because it is the parent each stage starts from. A search result is a ranking of what passed, not proof that the winner is the best model fitted.

```{r}
#| label: iivsearch-final
iiv$final_structure
iiv$final_etas
```

With `algorithm = "skip"` the eta-count stage is left out and only the block stage runs, which is the cheap way to ask one question: should `CL` and `V` share a covariance block? Skipping the eta-count stage means naming the block stage explicitly -- asking for neither is an error rather than a search with nothing left to do. The statements name parameters, not etas:

```{r}
#| label: iivsearch-blocks
iiv_block <- ferx_iivsearch(model = bo$model, data = bo$data,
                            search_space = "COVARIANCE?(IIV, [CL, V])",
                            algorithm = "skip",
                            correlation_algorithm = "top_down_exhaustive",
                            block_retries = 1,
                            directory = book_tempdir("iivsearch-block"), progress = FALSE)
c(input = iiv_block$input_structure, final = iiv_block$final_structure)
iiv_block$models[, c("id", "structure", "criterion", "rank", "passed")]
```

Two structures are compared here: the input's `block_omega (ETA_CL, ETA_V)` and the diagonal alternative. The input is kept -- but read why. The diagonal candidate reaches the better criterion of the two, and it is out because it fails the strictness gate. The `failures` column gives the reason:

```{r}
#| label: iivsearch-blocks-reason
with(iiv_block$models, failures[!is.na(failures)])
```

No free parameter moved more than 1% of its initial value, so `reject_init_stall` refused the fit. Whether that is the right call here depends on where the candidate started, which the printed output does not say: a candidate handed its parent's optimum can sit still because it is already converged, which is why `ferx_amd()` turns this gate off inside its steps (below). Read a gated-out candidate as unjudged rather than as rejected on merit, and refit it yourself if it matters. A selection is only as meaningful as the gate that produced it.

## Inter-occasion variability search

`ferx_iovsearch()` decides which parameters carry an occasion random effect. It reads the occasion column from the base model's `iov_column`, so a model without one has nothing to search. The bundled `warfarin_iov` configuration offers a kappa on every parameter that has an eta:

```{r}
#| label: iovsearch
iov_ex <- ferx_example("warfarin_iov")
iov <- ferx_iovsearch(config = iov_ex$search, directory = book_tempdir("iovsearch"), progress = FALSE)
iov
```

`distribution = "same-as-iiv"`, the default, declares each added kappa the way its eta is declared. The first step added kappas on `KA` and `V`; the second then found that dropping `ETA_V` improved the BIC(random) further: an IOV search can change the IIV structure, because the two are fitted together. `CL` already carries a kappa in the base model and is left as it is.

```{r}
#| label: iovsearch-final
iov$final_kappas
iov$final_structure
```

## The whole pipeline at once

`ferx_amd()` runs the tools above in order as one automatic model-development pipeline: structural, IIV, residual, IOV, allometry, covariates. One search space describes the whole pipeline, and the engine hands each step only the statements its tool can read. A step the space says nothing about is skipped, and says so.

`ferx_amd_plan()` shows that plan without fitting anything, which is the cheap way to check a configuration before spending the run:

```{r}
#| label: amd-plan
amd_ex <- ferx_example("warfarin_amd")
plan <- ferx_amd_plan(config = amd_ex$search)
plan
```

`r sum(!is.na(plan$skipped))` of the `r nrow(plan)` steps are already ruled out by the model and the space, and the `skipped` column says why: the model declares no `iov_column`, and the space carries no `ALLOMETRY` or `COVARIATE` statement because `warfarin.csv` has no covariate column. Running it fits the rest:

```{r}
#| label: amd
amd <- ferx_amd(config = amd_ex$search, directory = book_tempdir("amd"), progress = FALSE)
amd
```

The pipeline improved nothing: `warfarin_amd.ferx` is already the model these three steps would choose, so each step re-selected its own input and the final OFV equals the starting one. That is a useful result to see printed rather than hidden -- the step table carries each step's criterion before and after, its candidate count, the strictness verdict and its wall clock, so a step that "won" by having its siblings rejected looks different from one that found something. `summary()` adds every candidate of every step, and the same two tables are written to the run directory as `steps.csv` and `candidates.csv`.

The pipeline can also be stated inline, with a shorter order and steps left out. `strategy = "SIR"` runs structural, IIV and residual only; `skip` drops a step from whichever order was chosen; `retries_on` says when the extra perturbed-restart pass runs:

```{r}
#| label: amd-inline
amd_sir <- ferx_amd(model = amd_ex$model, data = amd_ex$data,
                    search_space = "PERIPHERALS(0..1); IIV?(@PK, exp)",
                    strategy = "SIR", skip = "residual", retries_on = "final",
                    retries = 0, directory = book_tempdir("amd-sir"), progress = FALSE)
amd_sir$steps[, c("step", "status", "criterion", "candidates", "seconds")]
```

## Tools not available from R

ferx-core also has a global search. It runs from the ferx command line and has no R function in the pinned ferx-r build. See the ferx-core [tools](https://ferx-nlme.github.io/ferx-core/tools/index.html) pages.

## Options that matter

| Function | Arguments |
|---|---|
| `ferx_bic()` | `fit`, `type` (`"mixed"`, `"fixed"`, `"iiv"`, `"random"`) |
| `check_strictness()` | `fit`, `require_converged`, `require_covariance`, `max_condition_number`, `max_correlation`, `reject_on_boundary`, `reject_init_stall` |
| `ferx_search_config()` | `path` |
| `ferx_search_space()` | `mfl` (text, config or space), `model`, `data` |
| `ferx_search_coverage()` | `space` (space, config or MFL text) |
| `ferx_search_results()` | `directory`, `partial` (`NULL` prefers the complete table), `type` (`"candidates"`, `"models"`, `"steps"`) |

The search functions share the entry arguments `model`, `data` and `config`; the run arguments `threads`, `retries`, `directory`, `resume` and `progress`; and, wherever candidates are ranked rather than tested, `search_space`, `rank` and `cutoff`. A search is stated either by `config` or by the inline arguments, not both:

| Function | Search arguments |
|---|---|
| `ferx_covsearch()` | `algorithm` (`"scm-forward-then-backward"` or `"scm-forward"`), `p_forward`, `p_backward`, `max_steps`, `adaptive_scope_reduction` |
| `ferx_modelsearch()` | `algorithm` (`"reduced_stepwise"`, `"exhaustive_stepwise"`, `"exhaustive"`), `iiv_strategy` (`"absorption_delay"`, `"add_diagonal"`, `"no_add"`) |
| `ferx_ruvsearch()` | `groups`, `p_value`, `skip` (families not to test), `max_iter` (1 to 3), `cwres_prescreen` (screen candidates on the parent's CWRES before refitting) |
| `ferx_iivsearch()` | `algorithm` (`"top_down_exhaustive"`, `"bottom_up_stepwise"`, `"simultaneous_stepwise"`, `"skip"`), `correlation_algorithm` (the block stage: `"top_down_exhaustive"` or `"skip"`), `as_fullblock` (a bottom-up candidate blocks every eta it carries), `block_retries` (extra starts per eta in the largest block) |
| `ferx_iovsearch()` | `column` (the occasion column; defaults to the model's `iov_column`, and disagreeing with it is an error), `distribution` (`"same-as-iiv"`, `"disjoint"`, `"joint"`, `"explicit"`), `groups` (the blocks `"explicit"` names), `block_retries` |
| `ferx_amd()`, `ferx_amd_plan()` | `strategy` (`"default"`, `"reevaluation"`, `"SIR"`, `"SRI"`, `"RSI"`), `skip` (steps to leave out), `retries_on` (`"all_final"`, `"final"`, `"skip"`) |

Only a `.ferxsearch` file can carry a per-tool section (`[modelsearch]`, `[iivsearch]`, `[covsearch]`, …); the inline form runs each step at its own defaults.

`print()` and `summary()` exist for the results of all of them: `ferx_covsearch`, `ferx_modelsearch`, `ferx_ruvsearch`, `ferx_iivsearch`, `ferx_iovsearch` and `ferx_amd` objects. `ferx_search_config` and `ferx_search_space` objects have `print()` methods.

## Pitfalls

- **Keep search directories.** The book writes them to a temporary directory. In a project, write them next to the analysis: they hold every candidate model, the tables and `final.ferx`, and they make `resume` possible.
- **`ferx_search_results()` needs the right `type`.** The covariate and residual error searches write `steps.csv`, the structural search `models.csv`. Neither wrote a `candidates.csv` here, so the default `type = "candidates"` finds no table for these runs.
- **Compare like with like.** OFV differences and likelihood ratio tests need the same data, the same estimation method and nested models.
- **Thresholds change the answer.** Record the space, the significance levels and the criterion together with the selected model. The `.ferxsearch` file is that record.

## Summary

- Compare fitted models by likelihood ratio test (nested) or with `fit$aic`, `fit$bic` and `ferx_bic()`. Gate candidates with `check_strictness()`.
- Describe searches in `.ferxsearch` files; check them with `ferx_search_config()`, `ferx_search_space()` and `ferx_search_coverage()`.
- Run covariate, structural and residual error searches with `ferx_covsearch()`, `ferx_modelsearch()` and `ferx_ruvsearch()`, and read their tables with `ferx_search_results()`.
- Search the random-effect structure with `ferx_iivsearch()` and the occasion structure with `ferx_iovsearch()`; both can change the IIV structure, so read what they selected rather than only the criterion.
- Run the whole order as one pipeline with `ferx_amd()`, after checking it with `ferx_amd_plan()`. A step the space says nothing about is skipped, and the step table says why.

Next: @sec-uncertainty quantifies the uncertainty of the selected covariate model.

::: {.callout-tip}
## Reference
- R help: `?ferx_bic`, `?check_strictness`, `?ferx_search_config`, `?ferx_search_space`, `?ferx_search_coverage`, `?ferx_search_results`, `?ferx_covsearch`, `?ferx_modelsearch`, `?ferx_ruvsearch`, `?ferx_iivsearch`, `?ferx_iovsearch`, `?ferx_amd`, `?ferx_amd_plan`
- ferx-core: [search configuration](https://ferx-nlme.github.io/ferx-core/tools/search.html), [covsearch](https://ferx-nlme.github.io/ferx-core/tools/covsearch.html), [modelsearch](https://ferx-nlme.github.io/ferx-core/tools/modelsearch.html), [ruvsearch](https://ferx-nlme.github.io/ferx-core/tools/ruvsearch.html), [iivsearch](https://ferx-nlme.github.io/ferx-core/tools/iivsearch.html), [iovsearch](https://ferx-nlme.github.io/ferx-core/tools/iovsearch.html), [amd](https://ferx-nlme.github.io/ferx-core/tools/amd.html), [tools](https://ferx-nlme.github.io/ferx-core/tools/index.html)
:::
