---
title: "Inter-Occasion Variability"
---

## Overview {#sec-iov}

**Inter-occasion variability (IOV)** captures variability in PK parameters that changes between dosing occasions within the same subject — for example, day 1 vs day 7 clearance after a multi-dose study.

In the ferx DSL, IOV is expressed with the `kappa` keyword.
Kappas are analogous to ETAs, but they are re-drawn for each occasion rather than being fixed per subject.

The individual parameter on occasion $k$ for subject $i$:

$$CL_{ik} = \text{TVCL} \times \exp(\eta_i^{CL} + \kappa_{ik}^{CL})$$

where $\eta_i^{CL} \sim N(0, \omega_{CL}^2)$ is the between-subject component (constant across occasions) and $\kappa_{ik}^{CL} \sim N(0, \omega_{IOV}^2)$ is the occasion-specific component.

## Dataset requirements

Add an `OCC` column (integer occasion index per row).
Convention: `OCC = 1` for the first occasion, `OCC = 2` for the second, etc.

```
ID  TIME  DV   EVID  AMT  OCC
 1   0.0  NA     1    70    1
 1   1.0  5.2    0     0    1
 ...
 1  24.0  NA     1    70    2     <- second dose, new occasion
 1  25.0  4.8    0     0    2
```

## The `kappa` keyword (diagonal IOV)

Declare IOV parameters with `kappa` in `[parameters]`:

```
[parameters]
  theta TVCL(0.2, 0.001, 10.0)
  theta TVV(10.0, 0.1, 500.0)
  theta TVKA(1.5, 0.01, 50.0)

  omega ETA_CL ~ 0.09   # between-subject variance (constant per subject)
  omega ETA_V  ~ 0.04
  omega ETA_KA ~ 0.30

  kappa KAPPA_CL ~ 0.01  # IOV variance (re-drawn each occasion)

  sigma PROP_ERR ~ 0.02
```

Reference `KAPPA_CL` in `[individual_parameters]` just like an ETA:

```
[individual_parameters]
  CL = TVCL * exp(ETA_CL + KAPPA_CL)
  V  = TVV  * exp(ETA_V)
  KA = TVKA * exp(ETA_KA)
```

Enable the IOV column in `[fit_options]`:

```
[fit_options]
  method     = foce
  iov_column = OCC
  covariance = false
```

## Fitting an IOV model

```{r}
#| label: setup
#| message: false
library(ferx)
```

The IOV example models live in the `ferx-core` repository.
Here we write the model inline using a temporary file:

```{r}
#| label: iov-model-write
iov_model <- tempfile(fileext = ".ferx")
writeLines(c(
  "# Warfarin IOV model",
  "[parameters]",
  "  theta TVCL(0.2, 0.001, 10.0)",
  "  theta TVV(10.0, 0.1, 500.0)",
  "  theta TVKA(1.5, 0.01, 50.0)",
  "  omega ETA_CL ~ 0.09",
  "  omega ETA_V  ~ 0.04",
  "  omega ETA_KA ~ 0.30",
  "  kappa KAPPA_CL ~ 0.01",
  "  sigma PROP_ERR ~ 0.02",
  "[individual_parameters]",
  "  CL = TVCL * exp(ETA_CL + KAPPA_CL)",
  "  V  = TVV  * exp(ETA_V)",
  "  KA = TVKA * exp(ETA_KA)",
  "[structural_model]",
  "  pk one_cpt_oral(cl=CL, v=V, ka=KA)",
  "[error_model]",
  "  DV ~ proportional(PROP_ERR)",
  "[fit_options]",
  "  method     = foce",
  "  iov_column = OCC",
  "  covariance = false"
), iov_model)
```

Fit using a dataset that has an `OCC` column:

```{r}
#| eval: false
fit_iov <- ferx_fit(iov_model, "warfarin_multi_dose.csv")
summary(fit_iov)
```

The summary output includes an **OMEGA_IOV** block:

```
--- OMEGA_IOV Estimates (Inter-Occasion Variability) ---
  KAPPA_CL = 0.012400  (CV% = 11.1)  SE = 0.004000  Shrinkage = N/A
```

## Correlated IOV: `block_kappa`

When IOV on multiple parameters is correlated (e.g. CL and V co-vary across occasions), use `block_kappa`:

```
[parameters]
  ...
  block_kappa (KAPPA_CL, KAPPA_V) = [0.02, 0.005, 0.01]
```

The three values are the lower-triangle of the 2×2 covariance matrix (column-major order):

| Position | Value | Meaning |
|---|---|---|
| `[1]` | 0.02 | Variance of KAPPA_CL |
| `[2]` | 0.005 | Covariance of KAPPA_CL and KAPPA_V |
| `[3]` | 0.01 | Variance of KAPPA_V |

Both kappas are used in `[individual_parameters]` the same way:

```
[individual_parameters]
  CL = TVCL * exp(ETA_CL + KAPPA_CL)
  V  = TVV  * exp(ETA_V  + KAPPA_V)
```

## Interpreting IOV results

### Partitioning variability

The total clearance variance (on the log scale) is:

$$\text{Var}(\log CL_i) = \omega_{CL}^2 + \omega_{IOV}^2$$

The fraction of total variability attributable to IOV:

```{r}
#| label: iov-fraction
#| eval: false
omega_bsv <- fit_iov$omega["ETA_CL", "ETA_CL"]
omega_iov <- fit_iov$omega_iov["KAPPA_CL", "KAPPA_CL"]
cat(sprintf("BSV: %.1f%%   IOV: %.1f%%   of total log-scale variance\n",
            100 * omega_bsv / (omega_bsv + omega_iov),
            100 * omega_iov / (omega_bsv + omega_iov)))
```

### IOV shrinkage

The IOV kappa shrinkage is reported in `fit$shrinkage_kappa`.
High shrinkage (> 30%) means the data cannot reliably estimate per-occasion kappas — consider removing IOV from that parameter.

## When to add IOV

Add IOV when:

- The study has multiple distinct dosing occasions with washout between them
- You see residual correlation in ETA-vs-time plots that suggests within-subject variation across occasions
- A model comparison (OFV difference) shows significant improvement with IOV included

Do not add IOV to studies with sparse sampling where the data cannot support additional random effects.
