# Simulating scenarios {#sec-simulation}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

```{r}
#| label: stage-banner
#| echo: false
#| output: asis
book_stage_banner("SIMULATE")
```

## Where you are

The covariate model is selected (@sec-model-selection) and its uncertainty quantified (@sec-uncertainty). The purpose of a model is usually to answer questions the study did not test directly. What is the typical profile of a light or a heavy patient? What exposure does another dose or dosing interval give, and how variable is it? How much of that spread comes from uncertainty in the estimates themselves?

In ferx you answer these by simulating from a **design dataset**: a CSV in the usual data format that describes the virtual subjects, their doses and the times you want predictions at, with the `DV` column left empty.

## The data

```{r}
#| label: fit
cov <- ferx_example("two_cpt_oral_cov")
fit <- ferx_fit(cov$model, cov$data, verbose = FALSE)
```

A small helper builds a design dataset from a table of subjects, one dose record each, with optional `II` and `ADDL` columns for repeated dosing:

```{r}
#| label: design-helper
design_dir <- book_tempdir("simulation")

write_design <- function(subjects, times, file) {
  doses <- subjects |>
    transmute(ID, TIME = 0, DV = NA, EVID = 1, AMT = dose, CMT = 1, RATE = 0, MDV = 1,
              II = interval, ADDL = additional, WT, CRCL)
  samples <- subjects |>
    select(ID, WT, CRCL) |>
    tidyr::crossing(TIME = times) |>
    mutate(DV = NA, EVID = 0, AMT = NA, CMT = 1, RATE = 0, MDV = 0, II = NA, ADDL = NA)
  design <- bind_rows(doses, samples) |>
    arrange(ID, TIME, desc(EVID)) |>
    select(ID, TIME, DV, EVID, AMT, CMT, RATE, MDV, II, ADDL, WT, CRCL)
  write.csv(design, file, row.names = FALSE, na = ".")
  file
}
```

The first scenario is a single dose of 250 or 500 mg for subjects of 50, 70 and 90 kg with a creatinine clearance of 100:

```{r}
#| label: single-dose-design
single <- tidyr::expand_grid(dose = c(250, 500), WT = c(50, 70, 90)) |>
  mutate(ID = row_number(), CRCL = 100, interval = 0, additional = 0)
single_file <- write_design(single, times = c(0.25, 0.5, 1, 2, 3, 4, 6, 8, 12, 16, 24, 36, 48),
                            file = file.path(design_dir, "single_dose.csv"))
head(read.csv(single_file), 4)
```

## Minimal runnable call: typical profiles

`ferx_predict()` returns the population prediction `PRED` (all random effects zero) for every observation row of the dataset. With `fit` it uses the estimated thetas; without it, the model file's initial values:

```{r}
#| label: predict
typical <- ferx_predict(cov$model, single_file, fit = fit)
head(typical)
```

```{r}
#| label: fig-typical
#| fig-cap: "Typical concentration profiles after a single oral dose, by dose and body weight."
typical |>
  mutate(ID = as.integer(ID)) |>
  left_join(single, by = "ID") |>
  ggplot(aes(TIME, PRED, colour = factor(WT))) +
  geom_line() +
  facet_wrap(~ paste(dose, "mg")) +
  labs(x = "Time (h)", y = "Typical concentration", colour = "Weight (kg)")
```

## Reading the result: variability around a regimen

`ferx_simulate()` adds between-subject variability and residual error to the design: each design subject is simulated `n_sim` times with new random effects (@sec-vpc). The scenario below compares two regimens with the same daily dose for a 70 kg subject: 250 mg once daily and 125 mg twice daily, each for 7 days (`II`, `ADDL`):

```{r}
#| label: regimen-design
regimens <- tibble::tibble(ID = 1:2, regimen = c("250 mg q24h", "125 mg q12h"),
                           dose = c(250, 125), interval = c(24, 12), additional = c(6, 13),
                           WT = 70, CRCL = 100)
regimen_file <- write_design(regimens, times = seq(0.5, 168, by = 0.5),
                             file = file.path(design_dir, "regimens.csv"))
sim <- ferx_simulate(cov$model, regimen_file, n_sim = 500, seed = 1, fit = fit)
```

The warning is expected: the design rows have an empty `DV`, so ferx simulates them as design points. It also records this in `attr(sim, "simulation_warnings")`.

```{r}
#| label: fig-regimens
#| fig-cap: "Median and 90% prediction interval of simulated concentrations for two regimens with the same daily dose."
sim_summary <- sim |>
  mutate(ID = as.integer(ID)) |>
  left_join(select(regimens, ID, regimen), by = "ID") |>
  group_by(regimen, TIME) |>
  summarise(p05 = quantile(DV_SIM, 0.05), p50 = median(DV_SIM), p95 = quantile(DV_SIM, 0.95),
            .groups = "drop")

ggplot(sim_summary, aes(TIME, p50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration", colour = NULL, fill = NULL)
```

Summaries come straight from the simulated rows. For example, the trough at 168 h, which is one dosing interval after the last dose of both regimens (144 h and 156 h):

```{r}
#| label: troughs
sim |>
  mutate(ID = as.integer(ID)) |>
  left_join(select(regimens, ID, regimen), by = "ID") |>
  filter(TIME == 168) |>
  group_by(regimen) |>
  summarise(median_trough = median(DV_SIM),
            p05 = quantile(DV_SIM, 0.05), p95 = quantile(DV_SIM, 0.95))
```

## Options that matter

| Function | Argument | Meaning |
|---|---|---|
| `ferx_predict()` | `model` | Model file |
| | `data` | Dataset or design; `NULL` uses the model's `[data]` block |
| | `fit` | Use the fitted thetas; `NULL` uses the model file's initial values |
| `ferx_simulate()` | `match` | Propensity-score matching of simulated random effects to the observed design (below) |
| `ferx_simulate_with_uncertainty()` | `model`, `data`, `fit` | Model, design and a fit carrying a covariance matrix (or SIR resamples) |
| | `n_uncertainty_draws` | Parameter sets drawn from the uncertainty distribution (default 100) |
| | `n_sim_per_draw` | Random-effect replicates per parameter set (default 1) |
| | `method` | `"asymptotic"` (multivariate normal from the covariance matrix) or `"sir"` (the fit's kept SIR resamples) |
| | `seed` | Random seed |

`ferx_simulate()`'s other arguments are described in @sec-vpc.

## Variants

### Adding parameter uncertainty

`ferx_simulate()` treats the estimates as exact. `ferx_simulate_with_uncertainty()` first draws parameter sets from their uncertainty distribution, then simulates each set with new random effects and residual errors. The `DRAW` column identifies the parameter set:

```{r}
#| label: with-uncertainty
sim_unc <- ferx_simulate_with_uncertainty(cov$model, single_file, fit,
                                          n_uncertainty_draws = 200, n_sim_per_draw = 5,
                                          method = "asymptotic", seed = 1)
nrow(sim_unc) == 200 * 5 * nrow(typical)
length(unique(sim_unc$DRAW))
```

The spread of the *median* profile across parameter draws shows how uncertain the typical prediction is. Here it is for the 70 kg subject on 500 mg:

```{r}
#| label: fig-uncertainty
#| fig-cap: "Uncertainty in the median concentration (90% interval of the per-draw median) for a 70 kg subject after 500 mg, from 200 parameter draws."
id_500_70 <- single$ID[single$dose == 500 & single$WT == 70]
sim_unc |>
  filter(as.integer(ID) == id_500_70) |>
  group_by(DRAW, TIME) |>
  summarise(median_conc = median(DV_SIM), .groups = "drop") |>
  group_by(TIME) |>
  summarise(lo = quantile(median_conc, 0.05), mid = median(median_conc),
            hi = quantile(median_conc, 0.95)) |>
  ggplot(aes(TIME, mid)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.3) +
  geom_line() +
  labs(x = "Time (h)", y = "Median concentration")
```

With `method = "sir"`, the parameter sets are sampled from SIR resamples kept on the fit (@sec-uncertainty). The fit must be made with `sir = TRUE` and `sir_keep_samples = TRUE`:

```{r}
#| label: with-uncertainty-sir
fit_sir <- ferx_fit(cov$model, cov$data, sir = TRUE, settings = list(sir_keep_samples = TRUE),
                    verbose = FALSE)
sim_sir <- ferx_simulate_with_uncertainty(cov$model, single_file, fit_sir,
                                          n_uncertainty_draws = 100, n_sim_per_draw = 5,
                                          method = "sir", seed = 1)
length(unique(sim_sir$DRAW))
```

### Matched simulation of observed data

In real-world data, dosing often depends on the patient: higher clearance, more frequent dosing. A plain simulation assigns random effects to subjects at random, which breaks that link. It can then make a VPC look wrong even when the model is right. `match` reassigns each replicate's simulated random effects to the observed subjects by propensity-score matching against their estimated random effects. The options are `TRUE` or `"optimal"` (recommended), `"nearest"` and `"rank"`:

```{r}
#| label: match
sim_matched <- ferx_simulate(cov$model, cov$data, n_sim = 100, seed = 1, fit = fit, match = "optimal")
nrow(sim_matched)
```

Matching needs every subject to have observations, so it works on the observed dataset but not on a design with empty `DV`:

```{r}
#| label: match-design
try(ferx_simulate(cov$model, single_file, n_sim = 2, seed = 1, fit = fit, match = TRUE))
```

### The `[simulation]` block

ferx-core model files can carry a `[simulation]` block that describes a virtual trial (number of subjects, dose, sampling times). It is used by the ferx command line for simulation-estimation studies. The ferx-r simulation functions do not read it: they always need a dataset.

```{r}
#| label: simulation-block
sim_model <- file.path(design_dir, "with_simulation_block.ferx")
writeLines(c(readLines(cov$model), "", "[simulation]", "  n_subjects = 5", "  dose_amt = 250",
             "  dose_cmt = 1", "  seed = 1", "  times = [1, 2, 4, 8]",
             "  covariate WT = 70", "  covariate CRCL = 100"), sim_model)
try(ferx_simulate(sim_model, n_sim = 1))
```

Build the equivalent design dataset in R instead, as above. See the ferx-core [simulation](https://ferx-nlme.github.io/ferx-core/model-file/simulation.html) page for the block itself.

## Pitfalls

- **Pass `fit`.** Without it, predictions and simulations use the model file's initial values (@sec-vpc).
- **Empty `DV` is simulated.** In a design this is what you want. In an observed dataset, such rows are simulated although a fit skips them, so they have no observed counterpart. Set `MDV = 1` to exclude rows from both.
- **Covariates must be in the design.** Every covariate the model uses needs a column in the design dataset.
- **Parameter uncertainty is a separate step.** Intervals from `ferx_simulate()` reflect variability only. Use `ferx_simulate_with_uncertainty()` when the uncertainty of the estimates matters for the question.

## Summary

- Build design datasets in R with doses (`AMT`, `II`, `ADDL`), sampling times, covariates and an empty `DV`.
- `ferx_predict()` gives typical profiles, and `ferx_simulate()` adds between-subject variability and residual error.
- `ferx_simulate_with_uncertainty()` adds parameter uncertainty (asymptotic or SIR).
- `match` keeps the link between dosing design and random effects when simulating observed data.

Next: @sec-tables-figures turns the results into report tables and figures.

::: {.callout-tip}
## Reference
- R help: `?ferx_predict`, `?ferx_simulate`, `?ferx_simulate_with_uncertainty`
- ferx-core: [data format](https://ferx-nlme.github.io/ferx-core/data-format.html), [simulation](https://ferx-nlme.github.io/ferx-core/model-file/simulation.html)
:::
