---
title: "Multi-endpoint PK/PD"
---

## Overview {#sec-multi-endpoint}

A model with **more than one type of observation** — for example, plasma
concentration **and** a pharmacodynamic readout — must apply a different
residual error model to each endpoint. The historical workaround is the
*sequential* approach (fit PK, freeze it, then fit PD), which is convenient
but biased: it ignores the cross-information the joint likelihood carries.

ferx solves this directly with **per-CMT dispatch in `[error_model]`**: each
observation row's `CMT` column selects which residual error model contributes
its likelihood. PK and PD are fit jointly, in one FOCEI likelihood, with their
own sigmas.

```{r}
#| label: setup
#| message: false
library(ferx)
```

## The bundled example: oral PK + Emax PD

The `emax_pkpd` example is an oral one-compartment PK linked to an effect
compartment with a Hill-Emax readout. Plasma concentration (CMT = 2) has
proportional error; the PD effect (CMT = 3) has additive error.

```{r}
#| label: show-model
ex <- ferx_example("emax_pkpd")
ferx_model_show(ex$model)
```

Three things make this work:

1. **Multi-state ODE.** `[structural_model]` declares `ode(states = [...])`
   *without* a single `obs_cmt`, because there is no single observation
   compartment.
2. **Per-CMT readout in `[scaling]`.** `y[CMT=2] = central / V` and
   `y[CMT=3] = E0 + ...` define exactly what is predicted for each
   compartment's observations.
3. **Per-CMT residual in `[error_model]`.** Each `CMT=N: DV ~ ...` line is a
   separate error model with its own sigma.

```
[scaling]
  y[CMT=2] = central / V
  y[CMT=3] = E0 + EMAX * effect^GAMMA / (EC50^GAMMA + effect^GAMMA)

[error_model]
  CMT=2: DV ~ proportional(PROP_ERR_PK)
  CMT=3: DV ~ additive(ADD_ERR_PD)
```

## Data layout

For a joint dataset, each subject contributes both PK and PD observation
rows, distinguished by the `CMT` column:

```
ID  TIME  DV   EVID  AMT  CMT   ...
 1   0.0   .     1    50    1    <- dose into depot
 1   0.5  5.4    0     .    2    <- PK observation (plasma)
 1   1.0  7.1    0     .    2
 1   1.0  4.8    0     .    3    <- PD observation (effect)
 ...
```

ferx routes each row to the appropriate readout and error model purely from
the `CMT` value — no separate datasets or merging required.

## Fitting

```{r}
#| label: fit
#| cache: true
#| eval: false
fit_pkpd <- ferx_fit(ex$model, ex$data, method = "focei", verbose = FALSE)
print(fit_pkpd)
```

The printed fit shows both sigmas on separate lines (one labelled
`[proportional]`, one `[additive]`), and the SIGMA block lists them in the
order they appear in `[parameters]`.

::: callout-note
**FD gradients.** Per-CMT error models live in the finite-difference
likelihood path. The model file therefore sets `gradient = fd` — automatic
differentiation is not available for the per-CMT dispatch in the current
release. Fits remain numerically reliable, just somewhat slower per
iteration than the AD-supported analytical PK shortcuts.
:::

## Diagnostics with multiple endpoints

Goodness-of-fit plots should be **separated by `CMT`** so each endpoint can be
inspected on its own scale:

```{r}
#| label: gof
#| eval: false
library(ggplot2)
library(dplyr)

sdtab <- fit_pkpd$sdtab |>
  mutate(endpoint = factor(CMT, levels = c(2, 3),
                           labels = c("Plasma (PK)", "Effect (PD)")))

ggplot(sdtab, aes(x = IPRED, y = DV)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, colour = "#1F6B7A") +
  facet_wrap(~ endpoint, scales = "free") +
  labs(x = "Individual prediction", y = "Observed") +
  theme_minimal()
```

Shrinkage is reported per ETA (not per endpoint); high shrinkage on a PD-only
ETA usually means the PD design is too sparse to estimate that effect
individually.

## When sequential is still appropriate

Joint estimation is the gold standard, but a sequential PK → PD workflow is
defensible when:

- The PK is already a published, well-validated model and you do not want PD
  data to perturb the PK estimates.
- PD endpoints are far slower than PK (very different time scales) and the
  joint Hessian becomes ill-conditioned.
- You are exploring a new PD model and want to iterate quickly without
  re-fitting PK each time.

For routine modeling, prefer joint estimation — the residuals on each
endpoint are then conditional on the *correct* PK uncertainty, not a fixed
point estimate.
