---
title: "Special features"
---

## Overview {#sec-special-features}

This chapter is a tour of optional `.ferx` blocks and `ferx_fit()` modes that
each cover a narrow but important case: unit conversion (`[scaling]`),
post-fit derived columns (`[derived]` / `[output]`), data filtering
(`[data_selection]`), steady-state and multiple additional doses (`SS` / `II` /
`ADDL` columns), absorption lag time, persisting fits to disk for later
analysis, log-transform-both-sides residual error, and stochastic differential
equations (`[diffusion]` / EKF process noise).

```{r}
#| label: setup
#| message: false
library(ferx)
```

## `[scaling]` — unit conversion and custom readouts

The `[scaling]` block performs a post-prediction transform before the
likelihood is computed. It has three forms.

### Form A — scalar `obs_scale`

The simplest case: divide every prediction by a constant, e.g. when `AMT` is
recorded in micrograms but `DV` in mg/L. The bundled `warfarin_scaled` example
uses this:

```
[scaling]
  obs_scale = 1000
```

```{r}
#| label: scaled-show
ferx_model_show(ferx_example("warfarin_scaled")$model)
```

The scalar form is gradient-safe: it does not interfere with the AD path, so
the model can keep `gradient = ad`.

### Form B — per-CMT `obs_scale[CMT=N]`

Different scale per observation compartment:

```
[scaling]
  obs_scale[CMT=2] = 1       # plasma — already in obs units
  obs_scale[CMT=3] = 0.001   # effect — convert ng to ug
```

### Form C — explicit `y[CMT=N] = <expression>`

A full custom readout. Used in `emax_pkpd` (see @sec-multi-endpoint) to drive
a Hill-Emax response off the effect-compartment amount:

```
[scaling]
  y[CMT=2] = central / V
  y[CMT=3] = E0 + EMAX * effect^GAMMA / (EC50^GAMMA + effect^GAMMA)
```

Form C and the per-CMT variants of Form B move the model onto the
finite-difference gradient path (`gradient = fd`).

## `[derived]` and `[output]` — post-fit sdtab columns {#sec-derived}

After a fit, `fit$sdtab` contains prediction-level columns (IPRED, PRED, CWRES,
…). The `[derived]` and `[output]` blocks add arbitrary extra columns to that
table — computed from the fitted individual parameters, predicted concentrations,
and time variables — without rerunning the likelihood.

### Three computation kinds

| Syntax | Kind | Evaluated |
|---|---|---|
| `COL = <expr>` | PerRow | At every observation row |
| `COL = max(IPRED, TIME < 24)` | Aggregate | One value per subject (max/min/tmax with optional filter) |
| `COL = integral(IPRED, from=0, to=72, step=0.2)` | Integral | Numerical trapezoid over a time window per subject |

All three share the same expression namespace: thetas, individual parameters
(`CL`, `V`, …), covariates (`AGE`, `WT`, …), `TIME` / `TAFD` / `TAD`, and
`IPRED`.

### Per-row: KE and half-life

```
[derived]
  KE     = CL / V
  T_HALF = 0.6931472 / KE
```

Columns defined earlier in the block can be reused in later expressions — `KE`
is in scope when `T_HALF` is evaluated.

### Aggregate: Cmax and Tmax

```
[derived]
  CMAX = max(IPRED)
  TMAX = tmax(IPRED)
  CMAX_48 = max(IPRED, TIME < 48)
```

`max`, `min`, and `tmax` aggregate over all observation rows for the subject.
An optional filter expression (second argument) restricts which rows are
included.

### Integral: AUC over a defined window

```
[derived]
  AUC_72 = integral(IPRED, from=0, to=72, step=0.2)
```

The integrand is re-evaluated on a fine time grid (`step` in the same units as
TIME) rather than just at the observation rows. Any per-row expression is valid
as the integrand.

### Time-above-threshold (TAT)

The constant integrand `1.0` with a filter measures duration above a threshold:

```
[derived]
  TAT_0_5 = integral(1.0, IPRED > 0.5, from=0, to=72, step=0.1)
```

This gives hours above 0.5 mg/L within the window. For per-interval TAT (using
TAD) the anchor and window form should be used — see the
[Derived columns reference](https://ferx-nlme.github.io/model-dsl/derived.html)
for the full syntax.

### Trough detection with `TAD < 1e-10`

Because `TAD` (time after dose) resets at each dose event, the smallest
positive `TAD` values are at dose times — but floating-point arithmetic can
make them `1e-13` rather than `0`. Use a tolerance:

```
[derived]
  CMIN_TAU = min(IPRED, TAD < 1e-10)
```

Do **not** use `TAD == 0`.

### `[output]` — individual parameters and covariates in sdtab

The `[output]` block echoes individual parameters and covariates into sdtab.
Without it, `fit$sdtab` contains only `IPRED`, `PRED`, residuals, and any
`[derived]` columns — not `CL`, `V`, `AGE`, etc.

```
[output]
  CL V KA AGE WT
```

Only items that are **individual parameters** (defined in `[individual_parameters]`)
or **covariates** (columns in the data) can appear in `[output]`. Thetas cannot.

### Runnable example

```{r}
#| label: derived-show
ex_d <- ferx_example("warfarin_derived")
ferx_model_show(ex_d$model)
```

```{r}
#| label: derived-fit
#| cache: true
#| eval: false
fit_d <- ferx_fit(ex_d$model, ex_d$data, verbose = FALSE)
names(fit_d$sdtab)
head(unique(fit_d$sdtab[, c("ID", "KE", "T_HALF", "CMAX", "AUC_0_72")]))
```

::: {.callout-tip}
For the full reference and more examples see
[Model DSL: Derived columns](https://ferx-nlme.github.io/model-dsl/derived.html)
on the ferx site.
:::

---

## `[data_selection]` — excluding records without touching the CSV {#sec-data-selection}

The `[data_selection]` block filters out records at read time. It is the ferx
equivalent of NONMEM's `$DATA IGNORE=`. The CSV is never modified; the filter
is part of the model specification.

### Syntax

```
[data_selection]
  ignore = DV < 1.0
  ignore = EVID == 0 && TIME > 120   # multiple conditions: use && within a line
  accept = BW >= 30                  # keep only rows where BW >= 30
  ignore_subjects = [7, 8]           # exclude subjects 7 and 8 entirely
```

Multiple `ignore` lines are OR-combined (any fired line excludes the record).
Use `&&` to combine conditions within a single `ignore` or `accept` line; `||`
within a single line is **not** supported — use separate lines instead.

`ignore_subjects` takes a list of subject IDs (integers matching the `ID`
column). A single ID can be written without brackets: `ignore_subjects = 7`.
It excludes all records for those subjects regardless of any other condition.

An `accept` condition keeps only rows where the expression is true.

### R-side preview with `ferx_selection()`

Before fitting, preview which records would be dropped:

```{r}
#| label: sel-preview
ex_sel <- ferx_example("warfarin_data_selection")
sel <- ferx_selection(ex_sel$data, ignore = "DV < 1.0")
cat("Excluded obs:", sel$exclusions$n_obs_excluded, "\n")
ferx_selection_excluded(sel)[, c("ID", "TIME", "DV")]
```

`ferx_selection()` returns the filtered data frame (plus an `$exclusions` list);
`ferx_selection_excluded()` returns only the dropped rows.

### What `fit$exclusions` reports

After fitting, `fit$exclusions` carries:

```{r}
#| eval: false
fit_sel <- ferx_fit(ex_sel$model, ex_sel$data, verbose = FALSE)
fit_sel$exclusions$n_records_total   # total rows read
fit_sel$exclusions$n_obs_excluded    # excluded observations
fit_sel$exclusions$fired_ignore      # which conditions fired
```

The `print(fit)` header also shows a `DATA SELECTION` line when any condition
fired.

### NONMEM equivalents

| NONMEM | ferx |
|---|---|
| `$DATA IGNORE=@` (comment rows) | handled automatically |
| `$DATA IGNORE=(column.EQ.value)` | `ignore = column == value` |
| `$DATA ACCEPT=(column.GE.value)` | `accept = column >= value` |

::: {.callout-tip}
For the full reference see
[Model DSL: Data selection](https://ferx-nlme.github.io/model-dsl/data-selection.html)
and the
[worked example](https://ferx-nlme.github.io/examples/warfarin-data-selection.html).
:::

---

## Steady-state dosing (`SS`, `II` columns)

For studies recorded after the pre-study dosing has reached steady state, add
two columns to the dose record:

- `SS = 1` on the dose row: ferx computes analytical steady-state initial
  conditions from the dose, interval, and the current individual parameters.
- `II` (inter-dose interval) in the same time units as `TIME`.

The bundled `warfarin_ss` example is one-compartment oral at QD steady state:

```{r}
#| label: ss-data
ex_ss <- ferx_example("warfarin_ss")
head(read.csv(ex_ss$data))
```

```{r}
#| label: ss-fit
#| cache: true
#| eval: false
fit_ss <- ferx_fit(ex_ss$model, ex_ss$data, verbose = FALSE)
```

The structural model is unchanged from the single-dose case — only the data
columns and the dose row's `SS`/`II` values differ. ferx resolves the
steady-state initial condition from the dose, the interval, and the current
individual parameters; the supported PK shortcuts use an analytical
closed-form.

## ADDL — multiple additional doses {#sec-addl}

Recording repeated identical doses one row at a time inflates the dataset.
The `ADDL` column (NONMEM-compatible) collapses a train of doses into a single
row:

```
ID,TIME,DV,EVID,AMT,CMT,RATE,MDV,II,ADDL
1,0,.,1,70,1,0,1,24,6     # day 1 dose + 6 more = 7 total, q24h
1,24,3.8,0,.,1,0,0,.,.
1,168,1.4,0,.,1,0,0,.,.
```

`ADDL=6` with `II=24` means: repeat the same dose 6 more times (so 7 doses
total), each 24 h apart. `TAD` resets to zero at each expanded dose event.

**No special DSL syntax is required.** The structural model is identical to a
single-dose model — multi-dose bookkeeping lives entirely in the data columns.

### Combining ADDL with `[derived]`

Because `TAD` resets correctly after each expanded dose, trough detection and
within-interval AUC work as expected:

```
[derived]
  CMIN_TAU = min(IPRED, TAD < 1e-10)   # trough
  AUC_TAU  = integral(IPRED, window=24, anchor=0, step=0.1)
```

::: {.callout-note}
The `warfarin_addl` model file is bundled and demonstrates this design.
However, the matching dataset is not yet bundled — supply your own ADDL-format
CSV that matches the column schema above.
:::

---

## Absorption lag time {#sec-lagtime}

Lag time can be expressed in two equivalent ways depending on which structural
model you use.

### As an argument to an analytical PK shortcut

```
[structural_model]
  pk one_cpt_oral(cl=CL, v=V, ka=KA, tlag=TLAG)
```

The named keyword `tlag` is accepted by the oral analytical solvers.

### As an `LAGTIME` keyword in `[individual_parameters]` (ODE models)

For ODE models the convention is to assign a special-named individual
parameter, which the solver consumes when processing dose events. The
`transit_2cpt` example uses this form:

```
[individual_parameters]
  ...
  LAGTIME = TVLAG
```

```{r}
#| label: lag-show
ex_lag <- ferx_example("transit_2cpt")
ferx_model_show(ex_lag$model)
```

In both forms the individual lag value flows through `[individual_parameters]`
like any other PK parameter — meaning it can be ETA-driven and recovered
per subject from `fit$individual_estimates`.

## Saving and reloading fits

Long-running fits should be persisted so you do not pay the estimation cost
twice. `ferx_save_fit()` writes a `.fitrx` file with the full fit object plus
SHA-256 hashes of the model file and dataset; `ferx_load_fit()` restores it
and warns if either has changed.

```{r}
#| label: save-load
#| eval: false
ferx_save_fit(fit, "warfarin.fitrx")                     # data not embedded
ferx_save_fit(fit, "warfarin.fitrx", include_data = TRUE)

fit_reloaded <- ferx_load_fit("warfarin.fitrx")
```

`include_data = TRUE` embeds the dataset as a base64 blob in the `.fitrx`
file. The file is then self-contained — useful for sharing across machines,
attaching to a regulatory submission, or moving a fit between branches of an
analysis without the dataset's path drifting.

## Bioavailability {#sec-bioavailability}

Estimating bioavailability `F` as a population parameter benefits from a
logit transform so it stays in `(0, 1)`. The bundled `bioavailability`
example uses `inv_logit(logit(THETA_F) + ETA_F)`; see @sec-logit-normal for
the parameterisation and reporting conventions.

## Log-transform-both-sides (LTBS) residual error {#sec-ltbs}

The familiar NONMEM idiom `Y = LOG(F) + EPS(1)` — fit log-transformed
observations against log-predictions with additive error on the log scale
— is written in `.ferx` as one of two equivalent forms:

```
[error_model]
  log(DV) ~ additive(SIGMA_LOG)   # DV on the natural scale; engine logs it

[error_model]
  DV ~ log_additive(SIGMA_LOG)    # DV already log-transformed in the data
```

The bundled `warfarin_ltbs` example uses the first form:

```{r}
#| label: ltbs-show
ferx_model_show(ferx_example("warfarin_ltbs")$model)
```

Under LTBS the diagnostic outputs (`IPRED`, `PRED`, `CWRES`, `IWRES` in
`sdtab`, and `DV_SIM` from `ferx_simulate()`) are on the log scale.
`ferx_model_inspect()` reports the residual type as
`additive (log-transformed)`.

## Stochastic differential equations (SDE / `[diffusion]`) {#sec-sde}

When IWRES shows autocorrelation that cannot be absorbed by the residual error
alone — residual structural misspecification, unmodelled absorption variability,
within-subject biological noise — the standard ODE treats the system as
deterministic and forces all unexplained variation into ETA and EPS.
The `[diffusion]` block turns the ODE into a stochastic differential equation
(SDE): the structural dynamics are themselves subject to continuous Gaussian
process noise.

ferx uses the **Extended Kalman Filter (EKF)** to propagate both the state and
its covariance forward in time. At each observation the integrated covariance
inflates the residual variance:

$$V_\text{total}(t) = P_\text{central}(t) + \sigma^2(\hat{y})$$

The diffusion parameter is fitted **on the variance scale** (not SD) alongside
the structural thetas.

### DSL syntax

Add a `[diffusion]` block after `[odes]`:

```
[diffusion]
  central ~ 0.01
```

The number is the initial estimate for `DIFF_CENTRAL`, the process-noise
variance on the `central` compartment state.

The bundled `warfarin_sde` example is a one-compartment oral ODE with SDE
process noise:

```{r}
#| label: sde-show
ferx_model_show(ferx_example("warfarin_sde")$model)
```

### Fitting an SDE model

```{r}
#| label: sde-fit
#| eval: false
ex_sde <- ferx_example("warfarin_sde")
fit_sde <- ferx_fit(ex_sde$model, ex_sde$data)
```

Key outputs that differ from a standard ODE fit:

```{r}
#| eval: false
fit_sde$uses_sde               # TRUE
fit_sde$theta["DIFF_CENTRAL"]  # fitted diffusion variance
ferx_estimates(fit_sde)        # DIFF_CENTRAL appears in the theta block
```

### SDE-specific notes

::: {.callout-important}
**Gradient method**: SDE models require `gradient_method = fd`; Enzyme autodiff
is not compatible with the EKF covariance propagation step.
`gradient_method = fd` is set automatically in the model file — no manual
override is needed. Finite-difference gradients make each function evaluation
significantly slower than an equivalent analytical PK model; this slowdown is
inherent to SDE fitting and cannot be mitigated by the Enzyme autodiff build.
:::

| Property | SDE model | Standard ODE |
|---|---|---|
| `fit$uses_sde` | `TRUE` | `FALSE` |
| Gradient method | `fd` (required) | `ad` (preferred) |
| Additional theta | `DIFF_CENTRAL` (variance) | — |
| CWRES / IWRES | Available | Available |
| EKF covariance propagation | Yes | No |

Standard diagnostics (`sdtab`, CWRES, IWRES) remain available; they reflect
the EKF-inflated total variance.

::: {.callout-tip}
For a full worked example see [Example: SDE](https://ferx-nlme.github.io/examples/sde.html)
on the ferx website.
:::

## Beyond this chapter

- **`[covariate_nn NAME]`** — Deep Compartment Models, a neural-network map
  from covariates to PK parameters. Feature-flagged (`cargo --features nn`);
  see the ferx-site
  [DCM example](https://ferx-nlme.github.io/examples/dcm.html).
- **`[derived]` / `[output]` full reference** —
  [Model DSL: Derived columns](https://ferx-nlme.github.io/model-dsl/derived.html)
  covers all aggregate forms, anchor/window integrals, naming rules, and sequential
  scoping rules.
- **`[data_selection]` full reference** —
  [Model DSL: Data selection](https://ferx-nlme.github.io/model-dsl/data-selection.html)
  documents evaluation order, `ignore` vs `accept` vs `ignore_subjects`, and
  supported column types.
- **Worked examples** — additional examples on the ferx site:
  [Derived columns](https://ferx-nlme.github.io/examples/warfarin-derived.html),
  [Derived PK/PD](https://ferx-nlme.github.io/examples/warfarin-derived-pkpd.html),
  [Data selection](https://ferx-nlme.github.io/examples/warfarin-data-selection.html),
  [ODE with TIME/TAFD/TAD](https://ferx-nlme.github.io/examples/warfarin-ode-time.html).
