---
title: "Special features"
---

## Overview {#sec-special-features}

This chapter is a tour of optional `.ferx` blocks and `ferx_fit()` modes that
each cover a narrow but important case: unit conversion (`[scaling]`),
steady-state dosing (`SS` / `II` columns), absorption lag time, persisting
fits to disk for later analysis, and log-transform-both-sides residual
error.

```{r}
#| label: setup
#| message: false
library(ferx)
```

## `[scaling]` — unit conversion and custom readouts

The `[scaling]` block performs a post-prediction transform before the
likelihood is computed. It has three forms.

### Form A — scalar `obs_scale`

The simplest case: divide every prediction by a constant, e.g. when `AMT` is
recorded in micrograms but `DV` in mg/L. The bundled `warfarin_scaled` example
uses this:

```
[scaling]
  obs_scale = 1000
```

```{r}
#| label: scaled-show
ferx_model_show(ferx_example("warfarin_scaled")$model)
```

The scalar form is gradient-safe: it does not interfere with the AD path, so
the model can keep `gradient = ad`.

### Form B — per-CMT `obs_scale[CMT=N]`

Different scale per observation compartment:

```
[scaling]
  obs_scale[CMT=2] = 1       # plasma — already in obs units
  obs_scale[CMT=3] = 0.001   # effect — convert ng to ug
```

### Form C — explicit `y[CMT=N] = <expression>`

A full custom readout. Used in `emax_pkpd` (see @sec-multi-endpoint) to drive
a Hill-Emax response off the effect-compartment amount:

```
[scaling]
  y[CMT=2] = central / V
  y[CMT=3] = E0 + EMAX * effect^GAMMA / (EC50^GAMMA + effect^GAMMA)
```

Form C and the per-CMT variants of Form B move the model onto the
finite-difference gradient path (`gradient = fd`).

## Steady-state dosing (`SS`, `II` columns)

For studies recorded after the pre-study dosing has reached steady state, add
two columns to the dose record:

- `SS = 1` on the dose row: ferx computes analytical steady-state initial
  conditions from the dose, interval, and the current individual parameters.
- `II` (inter-dose interval) in the same time units as `TIME`.

The bundled `warfarin_ss` example is one-compartment oral at QD steady state:

```{r}
#| label: ss-data
ex_ss <- ferx_example("warfarin_ss")
head(read.csv(ex_ss$data))
```

```{r}
#| label: ss-fit
#| cache: true
#| eval: false
fit_ss <- ferx_fit(ex_ss$model, ex_ss$data, verbose = FALSE)
```

The structural model is unchanged from the single-dose case — only the data
columns and the dose row's `SS`/`II` values differ. ferx resolves the
steady-state initial condition from the dose, the interval, and the current
individual parameters; the supported PK shortcuts use an analytical
closed-form.

## Absorption lag time

Lag time can be expressed in two equivalent ways depending on which structural
model you use.

### As an argument to an analytical PK shortcut

```
[structural_model]
  pk one_cpt_oral(cl=CL, v=V, ka=KA, tlag=TLAG)
```

The named keyword `tlag` is accepted by the oral analytical solvers.

### As an `LAGTIME` keyword in `[individual_parameters]` (ODE models)

For ODE models the convention is to assign a special-named individual
parameter, which the solver consumes when processing dose events. The
`transit_2cpt` example uses this form:

```
[individual_parameters]
  ...
  LAGTIME = TVLAG
```

```{r}
#| label: lag-show
ex_lag <- ferx_example("transit_2cpt")
ferx_model_show(ex_lag$model)
```

In both forms the individual lag value flows through `[individual_parameters]`
like any other PK parameter — meaning it can be ETA-driven and recovered
per subject from `fit$individual_estimates`.

## Saving and reloading fits

Long-running fits should be persisted so you do not pay the estimation cost
twice. `ferx_save_fit()` writes a `.fitrx` file with the full fit object plus
SHA-256 hashes of the model file and dataset; `ferx_load_fit()` restores it
and warns if either has changed.

```{r}
#| label: save-load
#| eval: false
ferx_save_fit(fit, "warfarin.fitrx")                     # data not embedded
ferx_save_fit(fit, "warfarin.fitrx", include_data = TRUE)

fit_reloaded <- ferx_load_fit("warfarin.fitrx")
```

`include_data = TRUE` embeds the dataset as a base64 blob in the `.fitrx`
file. The file is then self-contained — useful for sharing across machines,
attaching to a regulatory submission, or moving a fit between branches of an
analysis without the dataset's path drifting.

## Bioavailability

Estimating bioavailability `F` as a population parameter benefits from a
logit transform so it stays in `(0, 1)`. The bundled `bioavailability`
example uses `inv_logit(logit(THETA_F) + ETA_F)`; see @sec-logit-normal for
the parameterisation and reporting conventions.

## Log-transform-both-sides (LTBS) residual error {#sec-ltbs}

The familiar NONMEM idiom `Y = LOG(F) + EPS(1)` — fit log-transformed
observations against log-predictions with additive error on the log scale
— is written in `.ferx` as one of two equivalent forms:

```
[error_model]
  log(DV) ~ additive(SIGMA_LOG)   # DV on the natural scale; engine logs it

[error_model]
  DV ~ log_additive(SIGMA_LOG)    # DV already log-transformed in the data
```

The bundled `warfarin_ltbs` example uses the first form:

```{r}
#| label: ltbs-show
ferx_model_show(ferx_example("warfarin_ltbs")$model)
```

Under LTBS the diagnostic outputs (`IPRED`, `PRED`, `CWRES`, `IWRES` in
`sdtab`, and `DV_SIM` from `ferx_simulate()`) are on the log scale.
`ferx_model_inspect()` reports the residual type as
`additive (log-transformed)`.

## Beyond this chapter

The ferx engine has two further advanced blocks that this book does not yet
cover in depth:

- **`[diffusion]`** — stochastic differential equations (process noise) for
  ODE models where the structural dynamics themselves are uncertain. See
  [`warfarin_sde`](https://github.com/FeRx-NLME/ferx-r/tree/main/inst/examples)
  for a working example and the ferx-site
  [SDE example page](https://ferx-nlme.github.io/ferx-site/examples/sde.html).
- **`[covariate_nn NAME]`** — Deep Compartment Models, a neural-network map
  from covariates to PK parameters. Feature-flagged (`cargo --features nn`);
  see the ferx-site
  [DCM example](https://ferx-nlme.github.io/ferx-site/examples/dcm.html).
