---
title: "Model Search and Automatic Model Development"
---

## Overview {#sec-model-search}

Model development is a sequence of decisions: one compartment or two, an
absorption lag or not, which parameters carry between-subject variability,
whether the residual error needs a time-varying term, which covariates earn
their place.

Each of those decisions is a small, mechanical search over a space of
candidate models, ranked by a criterion and filtered by whether the fit can
be trusted at all. ferx can run them for you — the same tool chain as
Pharmpy's `amd`, driven from R:

| Function | Decides | Ranks on |
|---|---|---|
| `ferx_modelsearch()` | structural model: absorption, peripherals, transits, lag time | BIC (mixed) |
| `ferx_iivsearch()` | which parameters carry an eta, and which etas are correlated | BIC (IIV) |
| `ferx_iovsearch()` | which parameters carry inter-occasion variability | BIC (random) |
| `ferx_ruvsearch()` | the residual-error model | likelihood-ratio test |
| `ferx_covsearch()` | which covariate effects to include (SCM) | likelihood-ratio test |
| `ferx_allometry()` | allometric scaling on body size — a transform, not a search | — |
| `ferx_amd()` | all of the above in order, each step starting from what the last selected | per step |

```{r}
#| label: setup
#| message: false
library(ferx)
run_dir <- file.path(tempdir(), "search")
```

::: {.callout-note}
A search is only as good as the fits underneath it. Read
@sec-fitting first: everything here assumes a base model that fits, and
strictness settings you have thought about.
:::

## The search space is MFL

A space is written in Pharmpy's **Model Feature Language** and quoted
verbatim. ferx never translates R arguments into MFL — the space you write is
the space the engine reads, which is also what makes it portable to and from
Pharmpy.

```
ABSORPTION(FO); PERIPHERALS(0..1); LAGTIME([OFF,ON])
IIV?(@PK, exp); COVARIANCE?(IIV, *)
```

- A statement **without** `?` is structural: every candidate has it.
- A statement **with** `?` is exploratory: the search decides.
- `@PK`, `@IIV`, `@CONTINUOUS` and `@CATEGORICAL` resolve against your model
  and your dataset, so one space travels across models.

`ferx_search_space()` expands those symbols so you can see what a one-line
space actually means on your model, before committing to a run:

```{r}
#| label: space
ex_cov <- ferx_example("two_cpt_oral_cov")
ferx_search_space("COVARIATE?(@IIV, @CONTINUOUS, [pow, lin])",
                  model = ex_cov$model,
                  data  = ex_cov$data)
```

Not every feature the language can express is one the engine can build.
`ferx_search_coverage()` reports a gap as a row, with the engine's reason:

```{r}
#| label: coverage
ferx_search_coverage("ABSORPTION([FO, ZO, SEQ-ZO-FO]); ELIMINATION(FO)")
```

The tools themselves *refuse* an uncovered feature rather than quietly
narrowing the space — a search that silently drops what you asked for answers
a different question than the one you posed.

## The configuration file

A search can be driven two ways: a `.ferxsearch` file, or inline arguments.
The file is the reproducible artifact — it carries the space, the criterion,
the strictness gate, the run settings and the per-tool sections in one place,
and its paths are relative to itself, so it travels with the model.

```{r}
#| label: config
ex  <- ferx_example("warfarin_amd")
cfg <- ferx_search_config(ex$search)
cfg
```

The loader is strict, and it is strict *here* rather than three fits into a
run: an unknown section, an unparseable space, an empty space, an
unrecognised `[rank] type`, an invalid `[rank.penalties]` charge and a
coverage gap are all errors at load, each naming the offender.

::: {.callout-tip}
## Both entry forms are the same search

Every tool accepts **either** `config =` **or** inline arguments, never a
mix. The inline form builds the same configuration in the engine and goes
through the same loader, so a one-off run and a filed one cannot drift apart.
:::

## One decision at a time

### Structural model

```{r}
#| label: modelsearch
#| cache: true
ex_w <- ferx_example("warfarin")

ms <- ferx_modelsearch(
  model        = ex_w$model,
  data         = ex_w$data,
  search_space = "ABSORPTION(FO); PERIPHERALS(0..1); LAGTIME([OFF,ON])",
  retries      = 0,
  directory    = file.path(run_dir, "modelsearch"),
  progress     = FALSE
)
ms
```

Two things in that output matter more than the winner.

**The base model is allowed to win.** A search that cannot answer "none of
the above" is a search that always finds something.

**The excluded models are still in the table.** A candidate that did not
converge, pinned a parameter to a bound or came back ill-conditioned is not a
result worth ranking — but hiding it turns a rejected fit into an invisible
one:

```{r}
#| label: modelsearch-excluded
ms$models[!ms$models$passed, c("id", "structure", "failures")]
```

On this space those are the *lowest-OFV* models in the run. A winner-only
view would have handed you one of them.

### Variability structure

`ferx_iivsearch()` works in two stages: which parameters keep an eta, then
which of the survivors are correlated.

```{r}
#| label: iivsearch
#| cache: true
iv <- ferx_iivsearch(
  model        = ex_w$model,
  data         = ex_w$data,
  search_space = "IIV?(@PK, exp); COVARIANCE?(IIV, *)",
  retries      = 0,
  directory    = file.path(run_dir, "iivsearch"),
  progress     = FALSE
)
iv
```

The etas are named, not numbered. A block prints as `ETA_V ~ ETA_CL` and an
IOV term as `KAPPA_CL`, with `OMEGA(i,j)` appearing only where the model gave
no name — see @sec-iov for where those names come from.

`ferx_iovsearch()` is the same idea over occasions, and it reads its
occasions from the base model's `iov_column`: a model that declares none has
nothing to search, and the tool says so rather than searching an empty space.

### Residual error

```{r}
#| label: ruvsearch
#| cache: true
rv <- ferx_ruvsearch(
  model     = ex_w$model,
  data      = ex_w$data,
  retries   = 0,
  directory = file.path(run_dir, "ruvsearch"),
  progress  = FALSE
)
rv
```

Each candidate is a residual-error feature — a power term, a combined error,
a time-varying error — tested against the parent by likelihood-ratio test at
`p_value`. The per-iteration table carries the p-value it decided on, so
"which residual form won, and at what p" is a table rather than prose.

### Covariates and allometry

`ferx_covsearch()` is stepwise covariate modelling: `scm-forward` or
`scm-forward-then-backward`, with `p_forward` and `p_backward` thresholds and
adaptive scope reduction.

```{r}
#| label: covsearch
#| eval: false
ferx_covsearch(
  model        = ex_cov$model,
  data         = ex_cov$data,
  search_space = "COVARIATE?(@IIV, @CONTINUOUS, [pow, lin])",
  p_forward    = 0.01,
  p_backward   = 0.001,
  directory    = file.path(run_dir, "covsearch")
)
```

`ferx_allometry()` is not a search at all — it is the convention (`fix =
0.75` on clearances, `fix = 1` on volumes, centred on a reference weight)
applied as a model transform. With `fit = FALSE` it returns the edited model
without fitting anything, so allometry can be a step in a hand-built
workflow:

```{r}
#| label: allometry
al <- ferx_allometry(ex_cov$model, ex_cov$data,
                     covariate = "WT", reference = 70, fit = FALSE)
al
```

## The whole pipeline: `ferx_amd()`

`ferx_amd()` runs the tools in order, each step starting from the model the
previous step selected. One space describes the whole pipeline: the engine
partitions it by statement kind and hands each step only the statements its
own tool can read.

### Plan before you pay

`ferx_amd_plan()` answers "what would run, and what would not" without
fitting anything:

```{r}
#| label: amd-plan
ferx_amd_plan(config = ex$search)
```

A step that will not run says why: no `iov_column` in the model, no
`ALLOMETRY` statement in the space, no `COVARIATE` statement in the space.
That is the cheap way to find out your space says nothing about the step you
cared about.

### Run it

```{r}
#| label: amd-run
#| cache: true
res <- ferx_amd(config = ex$search,
                directory = file.path(run_dir, "amd"),
                progress  = FALSE)
res
```

The strategy sets the order. `default` is
structural → IIV → residual → IOV → allometry → covariates;
`reevaluation` appends a second IIV and residual pass once the model has its
covariates; `SIR`, `SRI` and `RSI` run three steps in the order their
initials spell.

### What comes back

```{r}
#| label: amd-steps
res$steps[, c("index", "step", "status", "criterion", "d_value",
              "converged", "passed", "selected")]
```

```{r}
#| label: amd-candidates
head(res$candidates[, c("step", "tool", "id", "description", "criterion",
                        "value", "rank", "passed", "selected")], 8)
```

Each step that ran has an entry under `$tools`, named by its position and
pipeline step (`01-structural`, `02-iivsearch`, `03-residual`) — not by its
directory, which is the tool's own name and is on the entry as `$directory`
(`01-modelsearch`, `02-iivsearch`, `03-ruvsearch`):

```{r}
#| label: amd-final
names(res$tools)
vapply(res$tools, function(t) t$directory, character(1))
```

An entry is a plain list, not a `ferx_modelsearch()` / `ferx_iivsearch()`
object: what the step decided (`status`, `reason`, `criterion`, `selected`,
`seconds`), its own slice of `$candidates`, and — when the run kept a
`directory` — the engine's own table read back from disk (`models` or
`steps`, whichever that tool writes) plus `input_model_path`,
`final_model_path` and `model_paths`. Run a tool on its own when you want its
object, with the fit and the options only a direct call returns.

```{r}
#| label: amd-tool-entry
names(res$tools[["01-structural"]])
```

The two "final" fields are different things: `res$final_model` is the model
*text* the pipeline ended on (written to `res$final_model_path`), while the
fitted `ferx_fit` is `res$fit`.

```{r}
#| label: amd-final-fit
res$final_ofv
res$d_ofv          # final - start; negative is an improvement
class(res$final_model)
class(res$fit)
```

::: {.callout-warning}
## `criterion` says what `value` is on

It is not always the step's own scale. With `cwres_prescreen = TRUE`,
ruvsearch screens its candidates against the parent's CWRES and refits only
the winner to the data: those rows are marked `screened` and carry
`criterion = "cwres_ofv"`, and their numbers must not be compared with a data
OFV. The engine default is `cwres_prescreen = FALSE` — every candidate is
fitted to the data, and the rows read `criterion = "ofv"` like any other
step, which is what the runs in this chapter do. The column is there so you
do not have to remember which of the two you are reading.
:::

## Reading a run back

Every tool writes its run to `directory`, and the tables there are the ones
the R object carries — no recomputation in R, so the two cannot disagree. An
interrupted run, or one driven from the `ferx` command-line tool, reads back
the same way:

```{r}
#| label: results
ferx_search_results(file.path(run_dir, "modelsearch"), type = "models")[,
  c("id", "parent", "layer", "criterion", "rank", "converged", "passed")]
```

- `type = "candidates"` (the default) reads `candidates.csv`
- `type = "models"` reads a structural or variability search's `models.csv`
- `type = "steps"` reads a stepwise search's `steps.csv`

Columns come from the engine's own column list and are typed on the way in:
logicals are logical, an empty cell is `NA` rather than `NaN`. A partial
table, written when a run is cancelled, reads back too.

`resume = TRUE` picks a run up where it stopped using the journal in
`directory`; `threads` and `retries` mirror `ferx_bootstrap()`, which is the
closest existing tool.

## Notes and limitations

- **Some `.ferxsearch` sections have no R tool.** The engine accepts a
  section for every tool *it* knows, which is more than this package binds —
  `[globalsearch]` and `[structsearch]` today. Such a file loads, and ferx
  warns that no R tool consumes the section, because the stepwise tools would
  otherwise run a stepwise search without saying so. The warning names where
  the section can be run instead when there is somewhere: `[globalsearch]` is
  the engine's own tool, behind `ferx globalsearch`. `[structsearch]` is
  accepted vocabulary with no engine module behind it yet, so it is reported
  without a remediation rather than with one that names a command that does
  not exist.
- **The strictness gate is shared.** `[strictness]` applies to every tool,
  and `converged` / `passed` are columns of every table that reports on
  fitted candidates — the candidate tables, the structural and variability
  `models.csv`, the covsearch and ruvsearch `steps.csv`, and AMD's own step
  table. They are not on the per-stage *ranking* summaries an R object
  carries alongside those (`ferx_iivsearch()$steps` is `step`, `kind`,
  `parent`, `id`, `criterion`, `d_criterion`, `rank`, `best`), which rank
  models the candidate rows already reported on. Inside an AMD step the init-stall gate is turned off
  by design: each step starts from a
  model already at its own optimum, where "did not move from the initial
  estimates" is the expected outcome rather than a failed fit.
- **Search is not a substitute for judgement.** These tools decide within the
  space you give them, on the criterion you choose, among the fits the gate
  lets through. All three are yours.
