# PK/PD and multiple endpoints {#sec-pkpd}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

## Where you are

So far every model had one endpoint, a drug concentration. Pharmacodynamic (PD) models describe a response. They can be linked to the pharmacokinetics and fitted together with the concentrations, used to compute a predicted effect from a PK fit, or stand on their own as an equation of time or dose. This chapter covers all three with the bundled examples.

::: {.callout-warning}
## Maturity: stable, beta and experimental parts
The error models are **stable** in ferx-core. ODE models, `[scaling]` and `[derived]` are **beta**. Compartment-free models are documented in the ferx-core [structural model](https://ferx-nlme.org/ferx-core/model-file/structural-model.html#compartment-free-models-the-pred-equivalent) page. See [feature maturity](https://ferx-nlme.org/ferx-core/maturity.html).
:::

## The data

```{r}
#| label: packages
#| message: false
library(ferx)
library(dplyr)
library(ggplot2)
```

`emax_pkpd` has 3 subjects with an oral dose, plasma concentrations and effect measurements in one dataset. The `CMT` column tells the endpoints apart: 2 for the concentration, 3 for the effect:

```{r}
#| label: data
pkpd <- ferx_example("emax_pkpd")
pkpd_dir <- file.path(tempdir(), "pkpd")
dir.create(pkpd_dir, showWarnings = FALSE)
pkpd_data <- read.csv(pkpd$data, na.strings = ".")
filter(pkpd_data, ID == 1) |> arrange(CMT, TIME)
count(filter(pkpd_data, EVID == 0), CMT)
```

## Minimal runnable call: a joint PK/PD fit

The model has three ODE states: the depot, the central compartment and an effect compartment that follows the plasma concentration with rate `KE0`. `[scaling]` gives one readout per observed `CMT` (@sec-scaling), and `[error_model]` one residual error model per `CMT`:

```{r}
#| label: pkpd-model
invisible(ferx_model_get_section(pkpd$model, "odes"))
invisible(ferx_model_get_section(pkpd$model, "scaling"))
invisible(ferx_model_get_section(pkpd$model, "error_model"))
invisible(ferx_model_get_section(pkpd$model, "fit_options"))
fit_pkpd <- ferx_fit(pkpd$model, pkpd$data, gradient = "auto", verbose = FALSE)
```

Both endpoints enter one likelihood, so the PK and PD parameters are estimated together. The model file asks for `gradient = fd`, and its comment calls finite differences required for this readout. At this build they are not, and the comment is out of date ([ferx-core #1625](https://github.com/FeRx-NLME/ferx-core/issues/1625)): ferx computes the analytic gradient for this model (`fit_pkpd$gradient_used` is "`r fit_pkpd$gradient_used`"), and the R warning above reports the override. The fits in this chapter pass `gradient = "auto"`, because the finite-difference fit stops short of the optimum, as the simulation–estimation run below shows.

## Reading the result

`sdtab` keeps the `CMT` column, so each endpoint can be plotted separately:

```{r}
#| label: fig-pkpd-fit
#| fig-cap: "Joint PK/PD fit: observations (points) and individual predictions (lines) by endpoint."
fit_pkpd$sdtab |>
  mutate(endpoint = ifelse(CMT == 2, "Concentration (CMT 2)", "Effect (CMT 3)")) |>
  ggplot(aes(TIME, colour = factor(ID))) +
  geom_point(aes(y = DV)) +
  geom_line(aes(y = IPRED)) +
  facet_wrap(~ endpoint, scales = "free_y") +
  labs(x = "Time (h)", y = NULL, colour = "ID")
```

The predictions follow the data, but three subjects do not determine the PD parameters:

```{r}
#| label: pkpd-estimates
fit_pkpd$estimates[, c("estimate", "rse_pct")]
pd <- c("TVE0", "TVEMAX", "TVEC50", "TVGAMMA")
round(fit_pkpd$cor_matrix[pd, pd], 2)
pkpd_warnings <- ferx_get_warnings(fit_pkpd, as_df = TRUE)
pkpd_warnings[, c("severity", "category")]
```

```{r}
#| label: check-pkpd-estimates
#| include: false
# The paragraph below reads these numbers and warnings, so a different picture has to fail the render.
pk <- c("TVCL", "TVV", "TVKA", "TVKE0")
pd_cor <- fit_pkpd$cor_matrix[pd, pd]
diag(pd_cor) <- 0
stopifnot(fit_pkpd$gradient_used == "analytic",
          all(c("high_correlation", "condition_number") %in% pkpd_warnings$category),
          max(fit_pkpd$estimates[pk, "rse_pct"]) < 10, fit_pkpd$estimates["TVE0", "rse_pct"] > 100,
          max(abs(pd_cor)) > 0.95)
```

The PK parameters are well determined, with relative standard errors of at most `r ceiling(max(fit_pkpd$estimates[pk, "rse_pct"]))`%. The PD parameters are not: the baseline `TVE0` has a relative standard error of `r round(fit_pkpd$estimates["TVE0", "rse_pct"])`%, and the PD parameters are almost perfectly correlated with each other (|r| up to `r round(max(abs(pd_cor)), 2)`). The model structure is fine; the question is what data would determine the PD parameters, which the simulation–estimation runs below answer.

## Options that matter

| Where | Syntax | Meaning |
|---|---|---|
| Data | `CMT` on observation rows | Which endpoint the row is |
| `[structural_model]` | `ode(states=[depot, central, effect])` | No single `obs_cmt`; `[scaling]` defines each readout |
| `[scaling]` | `y[CMT=2] = central / V`, `y[CMT=3] = <expr>` | One readout per observed `CMT` |
| `[error_model]` | `CMT=2: DV ~ proportional(...)`, `CMT=3: DV ~ additive(...)` | One residual error model per `CMT` (ODE models only) |
| `[structural_model]` | `y = <expr>` with no `pk` or `ode` line | A compartment-free model |
| `[derived]` | `EFF = EMAX * IPRED / (EC50 + IPRED)` | An effect computed from a PK fit |

Every observed `CMT` needs a readout and an error model. The two forms cannot be mixed in one block ([error model](https://ferx-nlme.org/ferx-core/model-file/error-model.html#multiple-endpoints-per-cmt-error-models)).

## Variants

### Simulation and estimation: can a design support the model?

To see whether the joint model can be estimated with a realistic design, simulate a study from the model's initial values and fit it again. `simulate_study()` below writes a design, simulates it once with `ferx_simulate()` (without `fit`, so with the model file's parameters, @sec-simulation) and returns the simulated study as a data file. The first design is the bundled one, scaled up: 30 subjects on 100, with six concentration and six effect samples each. `recovery()` puts each estimate beside the value it was simulated with:

```{r}
#| label: pkpd-sim-design
simulate_study <- function(doses, file, seed = 42) {
  ids <- seq_along(doses)
  design <- bind_rows(
    data.frame(ID = ids, TIME = 0, DV = NA, EVID = 1, AMT = doses, CMT = 1, RATE = 0, MDV = 1),
    expand.grid(ID = ids, TIME = c(0.5, 1, 2, 4, 8, 12)) |>
      mutate(DV = NA, EVID = 0, AMT = NA, CMT = 2, RATE = 0, MDV = 0),
    expand.grid(ID = ids, TIME = c(1, 2, 4, 8, 12, 24)) |>
      mutate(DV = NA, EVID = 0, AMT = NA, CMT = 3, RATE = 0, MDV = 0)
  ) |> arrange(ID, TIME, desc(EVID), CMT)
  design_file <- sub("\\.csv$", "_design.csv", file)
  write.csv(design, design_file, row.names = FALSE, na = ".")
  # the empty DV cells are the points to simulate, which ferx notes as W_DESIGN_DV
  simulated <- suppressWarnings(ferx_simulate(pkpd$model, design_file, n_sim = 1, seed = seed))
  rows <- design$EVID == 0
  design$DV[rows] <- simulated$DV_SIM[match(paste(design$ID, design$TIME, design$CMT)[rows],
                                            paste(simulated$ID, simulated$TIME, simulated$CMT))]
  write.csv(design, file, row.names = FALSE, na = ".")
  file
}
recovery <- function(f) {
  data.frame(parameter = c(names(f$theta), f$eta_names, f$sigma_names),
             simulated_with = c(f$theta_init, f$omega_init[c(1, 4)], f$sigma_init),
             estimate = c(f$theta, diag(f$omega), f$sigma),
             rse_pct = f$estimates$rse_pct)
}
one_dose <- simulate_study(rep(100, 30), file.path(pkpd_dir, "pkpd_100.csv"))
# suppress the R warning that reports overriding the model file's gradient
fit_one_dose <- suppressWarnings(ferx_fit(pkpd$model, one_dose, gradient = "auto", verbose = FALSE))
recovery(fit_one_dose)
```

The PK parameters are recovered and the PD parameters are not: `TVEMAX` ends at `r signif(fit_one_dose$theta[["TVEMAX"]], 3)` and `TVEC50` at `r signif(fit_one_dose$theta[["TVEC50"]], 2)`, where the simulation used 40 and 5, and their relative standard errors of `r round(fit_one_dose$estimates["TVEMAX", "rse_pct"])`% and `r round(fit_one_dose$estimates["TVEC50", "rse_pct"])`% do not flag it. This is where the data put the optimum, not a fit that went wrong. `maxiter = 0` evaluates the objective at the starting values without moving, which here are the simulation values themselves:

```{r}
#| label: pkpd-sim-truth
# suppress the R warnings that report overriding the model file's gradient and covariance setting
at_simulation_values <- suppressWarnings(ferx_fit(pkpd$model, one_dose, gradient = "auto", covariance = FALSE,
                                                  settings = list(maxiter = 0), verbose = FALSE))
data.frame(fit = fit_one_dose$ofv, simulation_values = at_simulation_values$ofv)
one_dose_data <- read.csv(one_dose, na.strings = ".")
max(one_dose_data$DV[one_dose_data$CMT == 2])
```

The true parameters score `r round(at_simulation_values$ofv - fit_one_dose$ofv, 1)` OFV points worse than the fit. The cause is the design: with 100 for every subject, the concentration never rises above `r round(max(one_dose_data$DV[one_dose_data$CMT == 2]), 1)`, below the `EC50` of 5, so the data see only the lower part of the Emax curve, and a smaller `EMAX` with a smaller `EC50` traces nearly the same part. More subjects on the same dose do not help. Doses that reach past `EC50` do -- the same number of subjects, ten each on 100, 300 and 1000:

```{r}
#| label: pkpd-sim-doses
dose_range <- simulate_study(rep(c(100, 300, 1000), each = 10), file.path(pkpd_dir, "pkpd_dose_range.csv"))
fit_dose_range <- suppressWarnings(ferx_fit(pkpd$model, dose_range, gradient = "auto", verbose = FALSE))
recovery(fit_dose_range)
```

```{r}
#| label: check-pkpd-sim
#| include: false
# The paragraphs around these chunks say the 100 design misses EMAX and EC50 at an optimum the
# simulation values cannot match, because its concentrations stay below EC50, and that the
# dose range recovers every typical value with moderate standard errors.
dose_range_data <- read.csv(dose_range, na.strings = ".")
pk <- c("TVCL", "TVV", "TVKA", "TVKE0")
emax_ec50 <- c("TVEMAX", "TVEC50")
miss <- function(f, p) abs(f$theta[p] / f$theta_init[match(p, names(f$theta))] - 1)
stopifnot(max(miss(fit_one_dose, pk)) < 0.1,
          fit_one_dose$theta[["TVEMAX"]] < 20, fit_one_dose$theta[["TVEC50"]] < 2.5,
          max(fit_one_dose$estimates[emax_ec50, "rse_pct"]) < 50,
          at_simulation_values$ofv - fit_one_dose$ofv > 5,
          max(one_dose_data$DV[one_dose_data$CMT == 2]) < 5,
          max(dose_range_data$DV[dose_range_data$CMT == 2]) > 10,
          max(miss(fit_dose_range, emax_ec50)) < 0.25,
          max(fit_dose_range$estimates$rse_pct) < 50)
```

With concentrations up to `r round(max(dose_range_data$DV[dose_range_data$CMT == 2]))`, `TVEMAX` and `TVEC50` come out at `r signif(fit_dose_range$theta[["TVEMAX"]], 3)` and `r signif(fit_dose_range$theta[["TVEC50"]], 2)`, within `r ceiling(100 * max(abs(fit_dose_range$theta[c("TVEMAX", "TVEC50")] / c(40, 5) - 1)))`% of 40 and 5 instead of several-fold off, and the largest relative standard error of any parameter is `r round(max(fit_dose_range$estimates$rse_pct))`%. One simulated study is one draw: repeat it with other seeds before settling a design. A simulation–estimation run answers the design question before the study is run.

::: {.callout-warning}
## The model file's finite-difference gradient stops short at this build

The fits above pass `gradient = "auto"`. With the model file's own `gradient = fd`, the fits of both designs converge well above their optimum, and refitting with a gradient-based optimizer, as @sec-estimation-methods suggests when `bobyqa` stops on a slope, does not rescue them:

```{r}
#| label: pkpd-sim-fd
gradient_runs <- function(data, fit_auto) {
  list(`gradient = fd (model file)` = ferx_fit(pkpd$model, data, verbose = FALSE),
       # suppress the R warning that reports overriding the model file's optimizer
       `gradient = fd, optimizer = slsqp` = suppressWarnings(
         ferx_fit(pkpd$model, data, settings = list(optimizer = "slsqp"), verbose = FALSE)),
       `gradient = "auto"` = fit_auto)
}
fd_runs <- list(`100 for all` = gradient_runs(one_dose, fit_one_dose),
                `100, 300, 1000` = gradient_runs(dose_range, fit_dose_range))
do.call(rbind, lapply(names(fd_runs), function(design) {
  do.call(rbind, lapply(names(fd_runs[[design]]), function(run) {
    f <- fd_runs[[design]][[run]]
    data.frame(design = design, fit = run, ofv = f$ofv, TVEMAX = f$theta[["TVEMAX"]],
               TVEC50 = f$theta[["TVEC50"]], max_rse_pct = max(f$estimates$rse_pct))
  }))
}))
```

```{r}
#| label: check-pkpd-sim-fd
#| include: false
# The callout says every fd fit stops well above the analytic optimum, that the objective is the
# same on both paths (the analytic estimates, written into the model file, score the same OFV on
# the fd path), and that on the 100 design the fd fits stay close to the simulation values.
fd_gap <- unlist(lapply(fd_runs, function(runs) sapply(runs[1:2], function(f) f$ofv - runs[[3]]$ofv)))
at_optimum <- readLines(pkpd$model)
est <- fit_one_dose$estimates
for (nm in names(fit_one_dose$theta)) {
  at_optimum <- sub(sprintf("theta %s\\(([^,]+),", nm), sprintf("theta %s(%.15g,", nm, est[nm, "estimate"]), at_optimum)
}
for (nm in fit_one_dose$eta_names) {
  at_optimum <- sub(sprintf("omega %s\\s*~.*$", nm), sprintf("omega %s ~ %.15g", nm, est[nm, "estimate"]), at_optimum)
}
for (nm in fit_one_dose$sigma_names) {
  at_optimum <- sub(sprintf("sigma %s\\s*~.*$", nm), sprintf("sigma %s ~ %.15g (sd)", nm, est[nm, "estimate"]), at_optimum)
}
at_optimum_file <- file.path(pkpd_dir, "emax_pkpd_at_optimum.ferx")
writeLines(at_optimum, at_optimum_file)
fd_at_optimum <- suppressWarnings(ferx_fit(at_optimum_file, one_dose, covariance = FALSE,
                                           settings = list(maxiter = 0), verbose = FALSE))
fd_emax <- sapply(fd_runs[[1]][1:2], function(f) f$theta[["TVEMAX"]])
stopifnot(all(fd_gap > 2), abs(fd_at_optimum$ofv - fit_one_dose$ofv) < 0.01,
          all(abs(fd_emax / 40 - 1) < 0.25),
          "covariance_regularized" %in% ferx_get_warnings(fd_runs[[1]][[1]], as_df = TRUE)$category)
```

Every finite-difference fit stops between `r round(min(fd_gap), 1)` and `r round(max(fd_gap), 1)` OFV points above the optimum that the analytic gradient finds. The objective is the same on both paths (the analytic optimum scores the same OFV on the finite-difference path), so it is the gradient that loses the optimum. On the 100 design this is the trap of a simulation–estimation run: the fit starts at the simulation values and stops near them, so `TVEMAX` comes out close to 40 and the design looks able to support the model, which it is not. Fit a simulated study with the most reliable settings you have, and compare its OFV with the objective at the simulation values. For this model, pass `gradient = "auto"` until the model file drops its `gradient = fd` ([ferx-core #1625](https://github.com/FeRx-NLME/ferx-core/issues/1625)).
:::

### Effects computed from a PK fit

When the PD parameters are known, an effect can be computed from the PK predictions without PD data. `warfarin_derived_pkpd` fixes `EMAX`, `EC50` and a minimum effective concentration `MEC` as thetas and uses them in `[derived]`:

```{r}
#| label: derived-pkpd
derived_pkpd <- ferx_example("warfarin_derived_pkpd")
invisible(ferx_model_get_section(derived_pkpd$model, "derived"))
fit_derived <- ferx_fit(derived_pkpd$model, derived_pkpd$data, verbose = FALSE)
filter(fit_derived$sdtab, ID == 1) |> select(TIME, IPRED, EFF, CMAX_INTERVAL, TAM_D1)
```

The fixed thetas are used only in `[derived]`, so the fit warns that they do not affect the predictions, and R warns about the correlation matrix because a fixed parameter has no standard error. Both are intended here. With `MEC = 0.5` every concentration in these data is above the MEC, so the time above it is the whole day. A copy with `MEC = 8` shows a threshold within the concentration range, per 24-hour window:

```{r}
#| label: time-above-mec
mec8_model <- file.path(pkpd_dir, "warfarin_derived_pkpd_mec8.ferx")
writeLines(sub("theta MEC(0.5)", "theta MEC(8.0)", readLines(derived_pkpd$model), fixed = TRUE), mec8_model)
fit_mec8 <- ferx_fit(mec8_model, derived_pkpd$data, verbose = FALSE)
filter(fit_mec8$sdtab, ID == 1) |> select(TIME, IPRED, TAM_TAU, TAM_D1)
```

Subject 1 is above 8 for most of the first day and about half of the second. From day 3 on its concentration stays below 8, and the time above the MEC in those windows is reported as `NaN`, not 0. Treat `NaN` as "never above" when you summarise it.

### Models without compartments

A response that is a plain equation of time, dose or other data needs no compartments. `emax_timecourse` writes the prediction directly in `[structural_model]`: `EFF` is an intermediate and `y` is the prediction. The dataset has no doses:

```{r}
#| label: timecourse
timecourse <- ferx_example("emax_timecourse")
invisible(ferx_model_get_section(timecourse$model, "structural_model"))
head(read.csv(timecourse$data, na.strings = "."), 3)
fit_timecourse <- ferx_fit(timecourse$model, timecourse$data, verbose = FALSE)
fit_timecourse$estimates[, c("estimate", "rse_pct")]
```

Such a model works with prediction, simulation and the diagnostics like any other. `TAD` and `TAFD` are `NaN` in `sdtab`, since there are no doses. For the same reason `ferx_predict()` passes on the data reader's `W_NO_DOSES` warning, and `W_CMT_DEFAULTED` because the dataset has no `CMT` column. For a compartment-free model with one endpoint, both state the design rather than a problem:

```{r}
#| label: fig-timecourse
#| fig-cap: "Compartment-free Emax time-course: observations (points) and the typical prediction (line)."
head(fit_timecourse$sdtab, 3)
ggplot(fit_timecourse$sdtab, aes(TIME)) +
  geom_line(aes(y = DV, group = ID), alpha = 0.2) +
  geom_line(data = ferx_predict(timecourse$model, timecourse$data, fit = fit_timecourse) |> distinct(TIME, PRED),
            aes(y = PRED), linewidth = 1) +
  labs(x = "Time", y = "Response")
```

A compartment-free model cannot have `[odes]`, `[scaling]` or references to `central` and `depot`. `y[CMT=N]` gives it several endpoints ([structural model](https://ferx-nlme.org/ferx-core/model-file/structural-model.html#compartment-free-models-the-pred-equivalent)).

## Pitfalls

- **Joint PK/PD models need data for every parameter.** Check relative standard errors and correlations before interpreting PD parameters; a simulation–estimation run tells whether a design can support the model.
- **An Emax design has to reach the `EC50`.** With concentrations that stay below it, `EMAX` and `EC50` cannot be told apart, whatever the number of subjects, and their standard errors need not say so. Compare the fit's OFV with the objective at the simulation values: a fit that stops near its starting values looks like a recovery.
- **Per-CMT error models need an ODE model.** An analytical `pk` model with `CMT=N:` lines is rejected. Use `ode_template` to write the PK part as ODEs.
- **Every observed `CMT` needs a readout and an error model.** A missing entry stops the fit and names the compartment.
- **Thresholds that are never crossed give `NaN`** in `[derived]` filters and conditional integrals.

### Warnings you may see here

- `general` (warning): parameters declared but not used by the predictions, such as the fixed PD thetas of `warfarin_derived_pkpd` that only `[derived]` reads.

Other warnings above (`inflated_rse`, `high_correlation`, `condition_number`) are covered in @sec-uncertainty.

## Summary

- Put all endpoints in one dataset and tell them apart with `CMT`.
- Write the PK and PD states as ODEs, one `[scaling]` readout and one error model per `CMT`, and fit them jointly.
- Check with simulation and estimation whether a design can support the PD parameters.
- Compute effects from a PK fit with `[derived]` when the PD parameters are known.
- Write responses that are plain equations as compartment-free models.

Next: @sec-binary covers binary endpoints.

::: {.callout-tip}
## Reference
- R help: `?ferx_fit`, `?ferx_simulate`, `?ferx_predict`
- ferx-core: [error model (multiple endpoints)](https://ferx-nlme.org/ferx-core/model-file/error-model.html#multiple-endpoints-per-cmt-error-models), [scaling](https://ferx-nlme.org/ferx-core/model-file/scaling.html), [structural model](https://ferx-nlme.org/ferx-core/model-file/structural-model.html), [`[derived]`](https://ferx-nlme.org/ferx-core/model-file/derived.html)
:::
