# Experimental features {#sec-experimental}

```{r}
#| label: setup
#| include: false
#| cache: false
source(if (file.exists("_common.R")) "_common.R" else "chapters/_common.R")
```

## Where you are

Two model extensions in ferx are explicitly experimental: stochastic differential equations, which add random noise to the model states between observations, and neural networks for the covariate model, known as deep compartment models. Both fit from R with the bundled examples. This chapter runs them, compares them with their conventional counterparts, and shows what the fits report. It closes with parameter priors, which carry the same label without being a model extension.

::: {.callout-warning}
## Maturity: experimental
ferx-core labels `[diffusion]` (SDE) and `[covariate_nn]` (neural networks) **experimental**: validated on a small set of examples, with syntax that may still change. Every fit that uses them carries an `experimental` warning. The [feature maturity](https://ferx-nlme.github.io/ferx-core/maturity.html) table's third experimental entry is the Vern7 ODE stepper (@sec-structural-models). Parameter priors, at the end of this chapter, are labelled experimental on their [own page](https://ferx-nlme.github.io/ferx-core/estimation/priors.html) but are absent from that table, and a priored fit raises no `experimental` warning.
:::

## The data

`warfarin_sde` uses the warfarin concentrations (10 subjects). `warfarin_dcm` uses the 30-subject dataset of the covariate thread, with weight and creatinine clearance:

```{r}
#| label: data
sde <- ferx_example("warfarin_sde")
dcm <- ferx_example("warfarin_dcm")
experimental_dir <- book_tempdir("experimental")
c(sde_subjects = n_distinct(read.csv(sde$data)$ID), dcm_subjects = n_distinct(read.csv(dcm$data)$ID))
```

## Minimal runnable call: an SDE model

An SDE model is an ODE model with a `[diffusion]` block. Each line gives a state a diffusion variance, estimated as the parameter `DIFF_<STATE>`. ferx evaluates the likelihood with an extended Kalman filter, which adds the propagated state uncertainty to the residual variance at each observation:

```{r}
#| label: sde-model
invisible(ferx_model_get_section(sde$model, "odes"))
invisible(ferx_model_get_section(sde$model, "diffusion"))
fit_sde <- ferx_fit(sde$model, sde$data, verbose = FALSE)
fit_sde$uses_sde
fit_sde$estimates[, c("estimate", "rse_pct")]
```

## Reading the result

The fit carries the experimental notice with its other warnings:

```{r}
#| label: sde-warnings
ferx_get_warnings(fit_sde, as_df = TRUE)[, c("severity", "category")]
substr(ferx_get_warnings(fit_sde, as_df = TRUE)$message[1], 1, 150)
```

The diffusion variance is small, and the covariance step does not support it: the relative standard error of every parameter, not only of `DIFF_CENTRAL`, runs to thousands of percent (the smallest is `r format(round(min(fit_sde$estimates$rse_pct)), big.mark = ",")`%), and the fit carries `covariance_regularized`, `inflated_rse`, `high_correlation` and a critical `condition_number` warning. The standard errors of this fit are not usable. Removing the `[diffusion]` block gives the ordinary ODE model, which reaches a similar OFV:

```{r}
#| label: sde-vs-ode
sde_lines <- readLines(sde$model)
start <- grep("^\\[diffusion\\]", sde_lines)
ode_model <- file.path(experimental_dir, "warfarin_ode_without_diffusion.ferx")
writeLines(sde_lines[-(start:(start + 2))], ode_model)
fit_ode <- ferx_fit(ode_model, sde$data, verbose = FALSE)
data.frame(model = c("with [diffusion]", "without"), ofv = c(fit_sde$ofv, fit_ode$ofv),
           durbin_watson = c(fit_sde$dw_statistic, fit_ode$dw_statistic))
```

These data show no system noise. A `[diffusion]` term is not a remedy for residuals that are correlated in time within a subject (@sec-diagnostics). The filter propagates the state *variance* and adds it to the residual variance; the predicted state itself stays the ODE solution and is never pulled towards a subject's observations, so the term cannot follow a drift. Here the Durbin-Watson statistic barely moves, `r round(fit_sde$dw_statistic, 2)` with the block and `r round(fit_ode$dw_statistic, 2)` without. What `[diffusion]` does estimate is how much unexplained variance builds up between observations.

## Options that matter

| Block or setting | Meaning |
|---|---|
| `[diffusion]` `STATE ~ variance` | Diffusion variance for an ODE state, estimated as `DIFF_STATE` (`FIX` to hold it) |
| `[covariate_nn NAME]` | A neural network with `inputs` (covariates), `outputs` (parameter names), hidden `layers`, `activation` and `output` activation |
| `NAME.OUTPUT` in `[individual_parameters]` | A network output, used like any other quantity |

```{r}
#| label: tbl-settings-nn
book_settings_table(c("nn_l2", "nn_smooth"))
```

Constraints from the ferx-core [diffusion](https://ferx-nlme.github.io/ferx-core/model-file/diffusion.html) page: `[diffusion]` needs an ODE model, cannot be combined with SAEM or `[scaling]` (@sec-scaling), and uses finite-difference gradients.

## Variants

### Estimation method for the SDE model

The bundled model uses FOCE. With FOCEI the diffusion variance is even smaller. The OFVs of the two methods are different objectives:

```{r}
#| label: sde-focei
# suppress the R warning that reports overriding the model file's method
fit_sde_focei <- suppressWarnings(ferx_fit(sde$model, sde$data, method = "focei", verbose = FALSE))
rbind(foce = c(ofv = fit_sde$ofv, fit_sde$theta), focei = c(ofv = fit_sde_focei$ofv, fit_sde_focei$theta))
```

Simulation from an SDE fit works like any other simulation (@sec-simulation):

```{r}
#| label: sde-simulate
head(ferx_simulate(sde$model, sde$data, n_sim = 1, seed = 1, fit = fit_sde), 3)
```

### A neural-network covariate model

`warfarin_dcm` replaces the covariate expressions of `two_cpt_oral_cov` with a network. The network reads `WT` and `CRCL` and outputs a multiplier for each PK parameter. The multipliers start at 1, so the thetas carry the typical values:

```{r}
#| label: dcm-model
# print the two blocks without their comment lines
section_lines <- function(block) {
  capture.output(lines <- ferx_model_get_section(dcm$model, block, strip = TRUE))
  trimws(lines[nzchar(trimws(lines)) & !grepl("^\\s*#", lines)])
}
section_lines("covariate_nn TYPICAL_PK")
section_lines("individual_parameters")
```

The network has 2 inputs, two hidden layers of 8 and 5 outputs: 141 weights, estimated as thetas. The covariance step is skipped here:

```{r}
#| label: dcm-fit
# suppress the R warning that reports overriding the model file's covariance setting
fit_dcm <- suppressWarnings(ferx_fit(dcm$model, dcm$data, covariance = FALSE, verbose = FALSE))
network <- fit_dcm$neural_networks$TYPICAL_PK
network[c("shape", "hidden_activation", "output_activation", "n_weights", "input_names", "output_names")]
ferx_get_warnings(fit_dcm, as_df = TRUE)[, c("severity", "category")]
```

The same data with the base model and with the explicit covariate model:

```{r}
#| label: tbl-dcm-comparison
#| tbl-cap: "Neural-network covariate model compared with the base and the explicit covariate model on the same data (covariance step skipped)."
base <- ferx_example("two_cpt_oral_base")
cov <- ferx_example("two_cpt_oral_cov")
fit_base <- suppressWarnings(ferx_fit(base$model, base$data, covariance = FALSE, verbose = FALSE))
fit_cov <- suppressWarnings(ferx_fit(cov$model, cov$data, covariance = FALSE, verbose = FALSE))
data.frame(model = c("two_cpt_oral_base (no covariates)", "two_cpt_oral_cov (WT, CRCL)", "warfarin_dcm (network)"),
           parameters = c(fit_base$n_parameters, fit_cov$n_parameters, fit_dcm$n_parameters),
           ofv = c(fit_base$ofv, fit_cov$ofv, fit_dcm$ofv),
           aic = c(fit_base$aic, fit_cov$aic, fit_dcm$aic))
```

With 30 subjects the network does not improve on the model without covariates, and its 141 weights make the AIC far worse than either alternative. The explicit covariate model captures the covariate effects with two parameters.

How much the network departs from multipliers of 1 can be read from the individual parameters. For clearance, `CL / (TVCL * exp(ETA_CL))` is the network's multiplier for each subject:

```{r}
#| label: dcm-multiplier
subjects <- read.csv(dcm$data) |> distinct(ID, WT, CRCL)
multipliers <- mutate(fit_dcm$individual_estimates, ID = as.integer(ID)) |>
  select(ID, CL) |>
  left_join(mutate(fit_dcm$ebe_etas, ID = as.integer(ID)) |> select(ID, ETA_CL), by = "ID") |>
  left_join(subjects, by = "ID") |>
  mutate(multiplier = CL / (fit_dcm$theta[["TVCL"]] * exp(ETA_CL)))
summary(multipliers$multiplier)
range(subjects$WT)
range(subjects$CRCL)
```

The multiplier is the same for every subject to seven significant digits, across the whole range of weight and creatinine clearance: the fitted network learned no covariate dependence at all, and only rescaled the typical clearance.

### Regularizing the network

`nn_l2` (weight decay) and `nn_smooth` (a penalty on curvature) keep a network from fitting spurious covariate effects. They are set in `[fit_options]` or with `settings`, apply to the FOCE-family methods, and the reported OFV stays the unpenalized one, so regularized and unregularized fits remain comparable ([regularization](https://ferx-nlme.github.io/ferx-core/model-file/covariate-nn.html#regularization-nn_l2-nn_smooth)). On this example a fit with `settings = list(nn_l2 = 0.01)` did not finish within 10 minutes, with either the model's `lbfgs` or the `bobyqa` optimizer, against about half a minute without the penalty. It is therefore not run in this book.

### Parameter priors

A prior pins a parameter near a value you already believe and lets the data move it only as far as the data can pay for -- penalized maximum likelihood, the mechanism behind updating a published model on local data. It is written inline on the `[parameters]` row it belongs to, as a central value and a relative standard error, both on the scale the parameter was declared on:

```{r}
#| label: prior-inline
warf <- ferx_example("warfarin")
priored <- file.path(experimental_dir, "warfarin_prior.ferx")
writeLines(sub("theta TVCL(0.134, 0.001, 10.0)",
               "theta TVCL(0.134, 0.001, 10.0) prior(0.25, rse = 10%)",
               readLines(warf$model), fixed = TRUE), priored)
fit_plain <- ferx_fit(warf$model, warf$data, verbose = FALSE)
fit_prior <- ferx_fit(priored, warf$data, verbose = FALSE)
c(no_prior = fit_plain$theta[["TVCL"]], with_prior = fit_prior$theta[["TVCL"]])
```

These data want a `TVCL` near `r signif(fit_plain$theta[["TVCL"]], 4)`; a prior centred at 0.25 with a 10% relative standard error pulls the estimate to `r signif(fit_prior$theta[["TVCL"]], 4)`. The penalty is added to the objective, so `fit$ofv` is the penalized total and does not compare with the unpenalized fit. The fit reports the two halves separately: `fit$ofv_data` is the data's share, `fit$ofv_prior` the penalty, and they sum to `fit$ofv`. The information criteria are built from the data half alone, so `fit$aic` is `fit$ofv_data + 2 * fit$n_parameters` and not `fit$ofv + 2 * fit$n_parameters` once a prior is in the model:

```{r}
#| label: prior-objective
data.frame(fit = c("no prior", "prior on TVCL"),
           ofv = c(fit_plain$ofv, fit_prior$ofv),
           ofv_data = c(fit_plain$ofv_data, fit_prior$ofv_data),
           ofv_prior = c(fit_plain$ofv_prior, fit_prior$ofv_prior),
           aic = c(fit_plain$aic, fit_prior$aic),
           ofv_data_plus_2k = c(fit_plain$ofv_data + 2 * fit_plain$n_parameters,
                                fit_prior$ofv_data + 2 * fit_prior$n_parameters))
```

A fit without a prior has `ofv_data` equal to `ofv` and an `ofv_prior` of 0. `ofv_data` is the number to set beside the unpenalized fit: `r round(fit_prior$ofv_data, 2)` against `r round(fit_plain$ofv, 2)`, so pulling `TVCL` towards 0.25 costs the data `r round(fit_prior$ofv_data - fit_plain$ofv, 2)` points of objective, on top of a penalty of `r round(fit_prior$ofv_prior, 2)`. `print()` and `summary()` state the split under the OFV line:

```{r}
#| label: prior-print
grep("OFV|prior", capture.output(print(fit_prior)), value = TRUE)
```

`fit$prior_summary` has one row per priored parameter, and is `NULL` for a model without one:

```{r}
#| label: prior-summary
fit_prior$prior_summary
is.null(fit_plain$prior_summary)
```

`shift_in_prior_sds` is how far the estimate landed from the prior's centre, with its sign, in prior standard deviations on the scale the prior is evaluated on. For a `lognormal` family that is the log scale, so it is not `(estimate - 0.25) / 0.025`. `penalty` is its square, and that parameter's share of `ofv_prior`. `prior_lower_95` and `prior_upper_95` are the interval that `rse = 10%` implies for the `family` the engine chose (`?ferx_fit` describes every column). An estimate `r round(abs(fit_prior$prior_summary$shift_in_prior_sds), 1)` prior standard deviations from the centre, outside that interval, says the prior and these data disagree. The standard errors of a priored fit are those of the penalized objective. What the prior cost in total is `fit_prior$ofv - fit_plain$ofv`, `r round(fit_prior$ofv - fit_plain$ofv, 2)` points: the data's share plus the penalty.

A `[priors]` block takes the priors from a previous fit instead of writing them out by hand. It has exactly one key, `from_fit`, naming a saved fit. That path does not work from R at this build: the only fit file ferx-r writes is the `.fitrx` bundle of `ferx_save_fit()`, and the engine's reader rejects it.

```{r}
#| label: priors-from-fit
saved <- file.path(experimental_dir, "warfarin.fitrx")
ferx_save_fit(fit_plain, saved)
from_fit_model <- file.path(experimental_dir, "warfarin_from_fit.ferx")
# normalizePath(): `//` starts a comment in a .ferx file, and R's tempdir()
# contains one on macOS, which would truncate the path before the engine saw it.
writeLines(c(readLines(warf$model), "", "[priors]",
             paste0("  from_fit = ", normalizePath(saved))),
           from_fit_model)
try(ferx_fit(from_fit_model, warf$data, verbose = FALSE))
```

The bundle stores every one-element array as a plain value, so the first such field the reader meets -- here the method chain, which it expects as a list of methods -- ends the import. That one field is not the whole of it: a chained fit writes a correct list and then stops at the next single-element field instead ([ferx-r #379](https://github.com/FeRx-NLME/ferx-r/issues/379)). Until the writer and the format agree, write priors inline.

## Pitfalls

- **Experimental means validate yourself.** Compare an SDE or network model with its conventional counterpart, as above, before relying on it.
- **`[priors] from_fit` does not work from R at this build.** The `.fitrx` bundle `ferx_save_fit()` writes is rejected by the engine's reader ([ferx-r #379](https://github.com/FeRx-NLME/ferx-r/issues/379)). Write the prior inline on the `[parameters]` row instead, as above.
- **Extra flexibility is not extra information.** A diffusion variance or a network with many weights needs data that support it; check standard errors, the OFV change and information criteria.
- **Standard errors of network weights are not reliable,** and the covariance step with many weights is slow. Skip it with `covariance = FALSE` while exploring.
- **SDE fits are slow.** The extended Kalman filter uses finite differences; the warfarin SDE fit takes many times longer than the ODE fit.

### Warnings you may see here

- `experimental` (warning): the model uses an experimental component (`[diffusion]` or `[covariate_nn]`). The message states what has and has not been validated. It appears on every such fit.

## Summary

- `[diffusion]` adds system noise to ODE states, estimated with an extended Kalman filter; compare with the model without it.
- `[covariate_nn]` lets a neural network produce covariate multipliers; compare its OFV and AIC with an explicit covariate model.
- `nn_l2` and `nn_smooth` regularize a network.
- An inline `prior(value, rse = …)` regularizes a parameter towards a value you already believe; the reported OFV is then penalized, and the `[priors]` block's `from_fit` import is not reachable from R at this build.
- `[diffusion]` and `[covariate_nn]` fits carry an `experimental` warning.

Next: @sec-reference lists every function, option and example with the chapter that covers it.

::: {.callout-tip}
## Reference
- R help: `?ferx_fit`, `?ferx_save_fit`
- ferx-core: [SDE and diffusion](https://ferx-nlme.github.io/ferx-core/model-file/diffusion.html), [covariate neural networks](https://ferx-nlme.github.io/ferx-core/model-file/covariate-nn.html), [neural networks overview](https://ferx-nlme.github.io/ferx-core/model-file/neural-networks.html), [parameter priors](https://ferx-nlme.github.io/ferx-core/estimation/priors.html)
:::
